---
title: "B_cells_v1"
author: "Kate Petrenko"
date: "2024-06-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
library(data.table)
library(Seurat)
library(scDblFinder)
library(DoubletFinder)
library(dplyr)
library(remotes)
# remotes::install_github("carmonalab/STACAS")
library(STACAS)
library(SingleR)
# install.packages("scGate")
library(scGate)
library(ggplot2)
library(BiocParallel)
library(harmony)
library(RColorBrewer)
library(stringr)
library(cowplot)
library(scuttle)
# install.packages("devtools")
# BiocManager::install("Homo.sapiens")
# BiocManager::install("jpmam1/scalop") #forked copy of a really good guy 
# devtools::install_github("jlaffy/infercna")
library(infercna)
library(infercnv)
library(parallel)
library(future)
library(clustree)
library(dittoSeq)
# install.packages("ggh4x")
library(ggh4x)
library(EnhancedVolcano)
library(clusterProfiler)
library("org.Hs.eg.db")
library(ReactomePA)
library(msigdbr)
library(qs)
```

```{r read data, include=FALSE}
set.seed(100)
DATADIR = '~/neuroblastoma/data/'
WORKDIR = '~/neuroblastoma/'
PREPRDATADIR = '~/neuroblastoma/preprocessing_data/'

my_colors <- c("#9ACD32FF", "#4682B4FF",  "#DDA0DDFF", "#FFA07AFF", "#8FBC8BFF",
               "#40E0D0FF",  "#F0E68CFF", "#5F9EA0FF", "#D2B48CFF",  
               "#6495EDFF", "#FF69B4FF", "#BA55D3FF",  "#F08080FF", "#32CD32FF",  
               "#FFDAB9FF",  "#87CEEBFF")

b_cells <- qread(paste0(PREPRDATADIR, 'b_cells.qs'))
```

# Data overview

```{r include=FALSE}
b_cells <- b_cells %>% 
  FindNeighbors(dims = 1:20) %>%
  FindClusters(resolution = c(0.1)) %>%
  RunUMAP(reduction = "integrated.cca.bcells", dims = 1:15)

Idents(b_cells) <- 'RNA_snn_res.0.1'
```

```{r fig.height=4, fig.width=10}
plot_grid(#DimPlot(b_cells, group.by = 'celltype_humatlas_fine'),
          DimPlot(b_cells, group.by = 'Study', cols = my_colors),
          #DimPlot(b_cells, group.by = 'scGate_multi'),
          DimPlot(b_cells, group.by = 'less18M', cols = my_colors[10:12]), 
          nrow=1)
```
```{r fig.height=4, fig.width=10}
plot_grid(dittoBarPlot(b_cells, 'less18M', group.by = 'Study', color.panel = my_colors[2:3]),
          dittoBarPlot(b_cells, 'less18M', group.by = 'sex', color.panel = my_colors[2:3]),
          nrow = 1)
```

# Markers of B-cells

## B-cells

```{r fig.height=8, fig.width=10}
FeaturePlot(b_cells, c('IGHM', 'IGHD', 'CD19', 'MS4A1', 'CD79A', 'TNFRSF13C', 'CD24', 'CD52'))
```

## Plasma cells 

```{r fig.height=8, fig.width=8}
FeaturePlot(b_cells, c('MZB1', 'CD27', 'CD38', 'SDC1', 'SLAMF7'))
```

## Pro-B-cells

```{r fig.height=8, fig.width=8}
FeaturePlot(b_cells, c('DNTT', 'VPREB1', 'MME',  'SOX4', 'PTPRC', 'CD19', 'CD34'))
```

# Annotation

```{r}
new.cluster.ids <- c('0'='Mature B cells',
                     '1'='Mature B cells',
                     '2'='Mature B cells',
                     '3'='Mature B cells',
                     '4'='Pre/Pro B cells',
                     '5'='Mature B cells',
                     '6'='Plasma cells',
                     '7'='Mature B cells',
                     '8'='Mature B cells',
                     '9'='Mature B cells',
                     '10'='Mature B cells'
)

b_cells <- RenameIdents(b_cells, new.cluster.ids)

b_cells$b_cells_manual_annot <- Idents(b_cells)
DimPlot(b_cells, cols = my_colors)
```

```{r}

dittoBarPlot(b_cells, 'less18M', group.by = 'b_cells_manual_annot', color.panel = my_colors)
```

# Differential expression analysis
```{r fig.height=8, fig.width=7}
b_cells.de.markers <- read.csv(paste0(PREPRDATADIR, 'b_cells_de_markers.csv'), row.names = 1)

volcano_plot <- function(de.markers) {
  EnhancedVolcano(de.markers,
    lab = rownames(de.markers),
    x = 'avg_log2FC',
    y = 'p_val_adj',
    col = c("grey30", "#9ACD32FF", "#6495EDFF", "#FF69B4FF"),
    drawConnectors = TRUE,
    pCutoff = 0.05) +
  labs(subtitle = "< 18M vs > 18M")
}

volcano_plot(b_cells.de.markers) +
            ggtitle('B-cells')
```

Genes are **up**-regulated in patients that are younger 18 months.

```{r fig.height=13, fig.width=6}
heatmap_bcells <- dittoHeatmap(b_cells, b_cells.de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC > 1) %>% 
                            dplyr::arrange(desc(avg_log2FC)) %>% 
                            rownames(),
             order.by = 'less18M', annot.by = c('less18M', 'Study', 'sex'), silent=T,
             annot.colors = c("#87CEEBFF", "#DDA0DDFF", "#9ACD32FF", "#4682B4FF", "#FFA07AFF", "#8FBC8BFF",
                              "#BA55D3FF",  "#F08080FF", "#32CD32FF"), main = 'B cells', 
             breaks=seq(-2, 2, 0.2), heatmap.colors = colorRampPalette(c("#6495EDFF", "white", "#FF69B4FF"))(length(seq(-2, 2, 0.2)))) 

plot_grid(heatmap_bcells[[4]])
```

Genes are **down**-regulated in patients that are younger 18 months.

```{r fig.height=8, fig.width=6}
heatmap_bcells_neg <- dittoHeatmap(b_cells, b_cells.de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC < -1) %>% 
                            dplyr::arrange(desc(avg_log2FC)) %>% 
                            rownames(),
             order.by = 'less18M', annot.by = c('less18M', 'Study', 'sex'), silent=T,
             annot.colors = c("#87CEEBFF", "#DDA0DDFF", "#9ACD32FF", "#4682B4FF", "#FFA07AFF", "#8FBC8BFF",
                              "#BA55D3FF",  "#F08080FF", "#32CD32FF"), main = 'B cells', 
             breaks=seq(-2, 2, 0.2), heatmap.colors = colorRampPalette(c("#6495EDFF", "white", "#FF69B4FF"))(length(seq(-2, 2, 0.2)))) 

plot_grid(heatmap_bcells_neg[[4]])
```

## Enrichment analysis

That's an enrichment analysis of genes **up**-regulated in patients that are younger 18 months (Positive markers) and genes **down**-regulated in patients that are younger 18 months (Negative markers)

```{r fig.height=7, fig.width=20}
H_t2g <- msigdbr(species = "Homo sapiens", category = "H") %>% 
  dplyr::select(gs_name, entrez_gene)

ImmuneSigDB <- msigdbr(species = "Homo sapiens", category = "C5") %>% 
  dplyr::select(gs_name, entrez_gene)

enrichplot_pos <- function(de.markers, gene.set) {
  em <- enricher(de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC > 1) %>% 
                            rownames() %>% 
                            mapIds(org.Hs.eg.db, 
                                   keys = .,
                                   column = "ENTREZID",
                                   keytype = "SYMBOL"),
                       TERM2GENE=gene.set)
  barplot(em) 
}

enrichplot_neg <- function(de.markers, gene.set) {
  em <- enricher(de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC < -1) %>% 
                            rownames() %>% 
                            mapIds(org.Hs.eg.db, 
                                   keys = .,
                                   column = "ENTREZID",
                                   keytype = "SYMBOL"),
                       TERM2GENE=gene.set)
  barplot(em) 
}

plot_grid(enrichplot_pos(b_cells.de.markers, H_t2g) +
            ggtitle('Positive markers'),
          enrichplot_neg(b_cells.de.markers, H_t2g) +
            ggtitle('Negative markers'), 
          nrow = 1)

plot_grid(enrichplot_neg(b_cells.de.markers, ImmuneSigDB) +
            ggtitle('Negative markers') ,
          nrow = 1)
```

# Enrichment score 

We noticed that on the heatmap we have two groups of cells. We decided to calculate enrichment score for them. 

Gene signature 1 - TUBB6, HERC5, PARP9, IFI44, MX2, OAS1, SAMD9, IFI44L, IFI6, ISG15 

Gene signature 2 - LYZ, S100A8, S100A9, KANSL1-AS1, GNLY, AKT1, PHC2, MAZ, HMGB3, PHGDH, IGLL1, CENPU, KLF16, CBX4, ZNF280D, C16orf74, H2AC6, COLGALT1, KIN, SSBP3, KIAA2013, RIOK2, C3orf38, ADARB1, N4BP2L1, NTAN1, GAB1, MTHFD2L, PPP1R14B, PTP4A2, TLE5, CD81, CYB5R4, ATP8A1, TAPT1, OGFRL1, CSNK1E, CDKN2D, WASF1, CDC25B, NSMCE1, CMTM7, BCL7A, MME, SOX4, H1-10, TFDP2, TIFA, TCF3, ABHD15, APBB2, PCDH9, REXO2, CCDC112, ACSM3, DTX1, CCDC191, H1-2, LETM1, MID1IP1, ZNF880, CCDC71L, ZNF254, COG3, ELF2, ASXL1, TLE1, PCIF1, ATF7IP2, BRI3, JUND, UBALD2, MEX3C, EPC1, C16orf87, MAPRE1, MTRF1L, RASSF1, TRA2A, MYADM, YPEL5, HRK, FOXN3, IQGAP1, CYFIP2, DYRK2, UHRF2, TMEM243, SYF2, UBE2J1, CXXC5, MAP1LC3B, PPP2R2D, PGM2L1, ARL8A, MAF1, WWC3, S1PR1

```{r fig.height=4, fig.width=10}
markers1 <- b_cells.de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC > 1) %>% 
                            dplyr::arrange(desc(avg_log2FC)) %>% 
                            rownames() %>% 
  .[heatmap_bcells$tree_row$order] %>% 
  .[1:10]

markers2 <- b_cells.de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC > 1) %>% 
                            dplyr::arrange(desc(avg_log2FC)) %>% 
                            rownames() %>% 
  .[heatmap_bcells$tree_row$order] %>% 
  .[11:108]

b_cells <- AddModuleScore(b_cells,
                  features = list(markers1, markers2),
                  name="gene_signature")


FeaturePlot(b_cells, c('gene_signature1', 'gene_signature2'))
```


# DE in mature B-cells

Because gene signatures are enriched in mature B-cells I perform DE in this cell type only.

```{r fig.height=8, fig.width=7}
b_cells$annotation_age <- paste(b_cells$b_cells_manual_annot, b_cells$less18M, sep='_')
Idents(b_cells) <- "annotation_age"
b_cells.de.markers <- FindMarkers(b_cells, ident.1 = "Mature B cells_TRUE", ident.2 = "Mature B cells_FALSE") %>% 
  dplyr::filter(pct.1 > 0.05 & pct.2 > 0.05)
Idents(b_cells) <- "b_cells_manual_annot"


volcano_plot(b_cells.de.markers) +
            ggtitle('Mature B-cells')
```

Genes are **up**-regulated in patients that are younger 18 months.

```{r fig.height=13, fig.width=6}
heatmap_bcells <- dittoHeatmap(b_cells, b_cells.de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC > 1) %>% 
                            dplyr::arrange(desc(avg_log2FC)) %>% 
                            rownames(),
             order.by = 'less18M', annot.by = c('less18M', 'Study', 'sex'), silent=T,
             annot.colors = c("#87CEEBFF", "#DDA0DDFF", "#9ACD32FF", "#4682B4FF", "#FFA07AFF", "#8FBC8BFF",
                              "#BA55D3FF",  "#F08080FF", "#32CD32FF"), main = 'Mature B cells', 
             breaks=seq(-2, 2, 0.2), heatmap.colors = colorRampPalette(c("#6495EDFF", "white", "#FF69B4FF"))(length(seq(-2, 2, 0.2)))) 

plot_grid(heatmap_bcells[[4]])
```

Genes are **down**-regulated in patients that are younger 18 months.

```{r fig.height=8, fig.width=6}
heatmap_bcells_neg <- dittoHeatmap(b_cells, b_cells.de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC < -1) %>% 
                            dplyr::arrange(desc(avg_log2FC)) %>% 
                            rownames(),
             order.by = 'less18M', annot.by = c('less18M', 'Study', 'sex'), silent=T,
             annot.colors = c("#87CEEBFF", "#DDA0DDFF", "#9ACD32FF", "#4682B4FF", "#FFA07AFF", "#8FBC8BFF",
                              "#BA55D3FF",  "#F08080FF", "#32CD32FF"), main = 'B cells', 
             breaks=seq(-2, 2, 0.2), heatmap.colors = colorRampPalette(c("#6495EDFF", "white", "#FF69B4FF"))(length(seq(-2, 2, 0.2)))) 

plot_grid(heatmap_bcells_neg[[4]])
```

## Enrichment analysis

That's an enrichment analysis of genes **up**-regulated in patients that are younger 18 months (Positive markers) and genes **down**-regulated in patients that are younger 18 months (Negative markers)

```{r fig.height=7, fig.width=20}

plot_grid(enrichplot_pos(b_cells.de.markers, H_t2g) +
            ggtitle('Positive markers'),
          enrichplot_neg(b_cells.de.markers, H_t2g) +
            ggtitle('Negative markers'), 
          nrow = 1)

plot_grid(enrichplot_neg(b_cells.de.markers, ImmuneSigDB) +
            ggtitle('Negative markers'),
          nrow = 1)
```

# Enrichment score 

We noticed that on the heatmap we have two groups of cells. We decided to calculate enrichment score for them. 

Gene signature 1 - TUBB6, PARP9, PLSCR1, MX2, SAMD9L, IFI44L, IFI6, ISG15, IRF7, OAS1, EPSTI1

Gene signature 2 - CCDC71L, N4BP2L1, PCIF1, DYRK2, PPP1R14B, ATF7IP2, BRI3, JUND, UBALD2, MEX3C, EPC1, WWC3, TRA2A, MYADM, YPEL5, HRK, FOXN3, IQGAP1, CDC14A, C3orf38, PPP2R2D, PGM2L1, MAF1, ARL8A, ELF2, LYRM1, CBX4, GADD45A, COG3, MAPRE1, ASXL1, C16orf87, XIST, MTRF1L, RASSF1, SUCLA2, KIAA2013, ADARB1, ATP8A1, CYB5R4, ACSS1, UNK, HMCES, FKBP1A, MID1IP1, NTAN1, C16orf74, TBC1D1, ZNF280D, LETM1, REPIN1, TFDP2, MIR181A1HG, CDKN2D, CD38, CD9, CDC25B, NSMCE1, CMTM7, CD24, MME, SOX4, H1-10, ABHD15, APBB2, PCDH9, REXO2, CCDC112, ACSM3, DTX1, H1-2, CCDC191, BCL7A, KIN, MTHFD2L, UHRF2, TMEM243, SYF2, UBE2J1, CXXC5, MAP1LC3B, CYFIP2, TAPT1, OGFRL1, CSNK1E, TIFA, TCF3, PTP4A2, TLE5, CD81, H2AC6, IGLC3, ITM2C, MZB1, CST3, LYZ, S100A8, S100A9

```{r fig.height=4, fig.width=10}
markers1 <- b_cells.de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC > 1) %>% 
                            dplyr::arrange(desc(avg_log2FC)) %>% 
                            rownames() %>% 
  .[heatmap_bcells$tree_row$order] %>% 
  .[1:11]

markers2 <- b_cells.de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC > 1) %>% 
                            dplyr::arrange(desc(avg_log2FC)) %>% 
                            rownames() %>% 
  .[heatmap_bcells$tree_row$order] %>% 
  .[12:109]

b_cells <- AddModuleScore(b_cells,
                  features = list(markers1, markers2),
                  name="gene_signature")


FeaturePlot(b_cells, c('gene_signature1', 'gene_signature2'))
```
