---
title: "cellchat"
author: "Kate Petrenko"
date: "2024-09-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
library(data.table)
library(Seurat)
library(scDblFinder)
library(DoubletFinder)
library(dplyr)
library(remotes)
# remotes::install_github("carmonalab/STACAS")
library(STACAS)
library(SingleR)
# install.packages("scGate")
library(scGate)
library(ggplot2)
library(BiocParallel)
library(harmony)
library(RColorBrewer)
library(stringr)
library(cowplot)
library(scuttle)
# install.packages("devtools")
# BiocManager::install("Homo.sapiens")
# BiocManager::install("jpmam1/scalop") #forked copy of a really good guy 
# devtools::install_github("jlaffy/infercna")
library(infercna)
# library(infercnv)
library(parallel)
library(future)
library(clustree)
library(dittoSeq)
# install.packages("ggh4x")
library(ggh4x)
#BiocManager::install('EnhancedVolcano')
library(EnhancedVolcano)
library(clusterProfiler)
# BiocManager::install('org.Mm.eg.db')
library("org.Mm.eg.db")
library(ReactomePA)
library(msigdbr)
# install.packages('qs')
library(qs)
# devtools::install_github("ncborcherding/escape")
# devtools::install_github("rcastelo/GSVA")
library(escape)
# install.packages('gprofiler2')
library(gprofiler2)
# remotes::install_github("carmonalab/ProjecTILs")
library(ProjecTILs)
library(gridExtra)
library(scCustomize)
library(enrichR)
# devtools::install_github("jinworks/CellChat")
library(CellChat)
library(patchwork)
options(stringsAsFactors = FALSE)
# install.packages('pryr')
# library(pryr)
# reticulate::py_install(packages = c("umap-learn==0.5.4","numpy<2"))
library(ComplexHeatmap)
```

```{r global variables server, include=FALSE}
set.seed(100)
DATADIR = '~/neuroblastoma/data/'
WORKDIR = '~/neuroblastoma/'
PREPRDATADIR = '~/neuroblastoma/preprocessing_data/'

my_colors <- c("#9ACD32FF", "#4682B4FF",  "#DDA0DDFF", "#FFA07AFF", "#8FBC8BFF",
               "#40E0D0FF",  "#F0E68CFF", "#5F9EA0FF", "#D2B48CFF",  
               "#6495EDFF", "#FF69B4FF", "#BA55D3FF",  "#F08080FF", "#32CD32FF",  
               "#FFDAB9FF",  "#87CEEBFF")
```

```{r eval=FALSE, include=FALSE}
seu <- qread(paste0(PREPRDATADIR, 'ATLAS_object.qs'))
sample_metadata <- read.csv(paste0(DATADIR, 'samples_metadata_v5.csv'))
sample_metadata[sample_metadata==""] <- NA

rownames(sample_metadata) <- sample_metadata$Sample_dataset



clean_sample_id <- function(id) {
  # First remove _rep1 and _rep2
  id <- gsub("_rep[12]$", "", id)
  
  # Handle the special bonine cases
  id <- gsub("_sn$", "", id)      # Remove _sn suffix
  id <- gsub("_[12]$", "", id)    # Remove _1 or _2 suffix
  
  return(id)
}

seu$less18M <- sample_metadata[clean_sample_id(seu$SampleID),]$less18M
seu$Sex <- sample_metadata[clean_sample_id(seu$SampleID),]$Sex
seu$MYCN_status <- sample_metadata[clean_sample_id(seu$SampleID),]$MYCN
seu$method <- sample_metadata[clean_sample_id(seu$SampleID),]$Method
seu <- subset(seu, Study == 'patel')
```

# CellChat

I decided to apply CellChat to identify major signaling changes across ages.

```{r eval=FALSE, include=FALSE}
seu$samples <- seu$less18M
cellchat_young <- createCellChat(object = subset(seu, samples == 'TRUE'), group.by = "Annotation", assay = "RNA")
```

I used CellChatDB v2 that contains ~3,300 validated molecular interactions, including ~40% of secrete autocrine/paracrine signaling interactions, ~17% of extracellular matrix (ECM)-receptor interactions, ~13% of cell-cell contact interactions and ~30% non-protein signaling. But I used all CellChatDB except for "Non-protein Signaling".

```{r echo=FALSE}
CellChatDB <- CellChatDB.human
showDatabaseCategory(CellChatDB)
```

```{r eval=FALSE, include=FALSE}
# use all CellChatDB except for "Non-protein Signaling" for cell-cell communication analysis "Non-protein Signaling" (i.e., metabolic and synaptic signaling)
CellChatDB.use <- subsetDB(CellChatDB) 
cellchat_young@DB <- CellChatDB.use
```

```{r eval=FALSE, include=FALSE}
cellchat_young <- subsetData(cellchat_young)
future::plan("multisession", workers = 8) # do parallel
options(future.globals.maxSize = 4 * 1024^3)
cellchat_young <- identifyOverExpressedGenes(cellchat_young)
cellchat_young <- identifyOverExpressedInteractions(cellchat_young)
```

```{r eval=FALSE, include=FALSE}
cellchat_young <- computeCommunProb(cellchat_young, type = "triMean")
cellchat_young <- filterCommunication(cellchat_young, min.cells = 10)
```

```{r eval=FALSE, include=FALSE}
df.net.young <- subsetCommunication(cellchat_young)
```

```{r eval=FALSE, include=FALSE}
cellchat_young <- computeCommunProbPathway(cellchat_young)
cellchat_young <- aggregateNet(cellchat_young)
cellchat_young <- netAnalysis_computeCentrality(cellchat_young)

groupSize <- as.numeric(table(cellchat_young@idents))
par(mfrow = c(1,2), xpd=TRUE)
netVisual_circle(cellchat_young@net$count, vertex.weight = groupSize, weight.scale = T, label.edge= F, title.name = "Number of interactions")
netVisual_circle(cellchat_young@net$weight, vertex.weight = groupSize, weight.scale = T, label.edge= F, title.name = "Interaction weights/strength")
```



```{r eval=FALSE, include=FALSE}
seu$samples <- seu$less18M
cellchat_adult <- createCellChat(object = subset(seu, samples == 'FALSE'), group.by = "Annotation", assay = "RNA")
```

```{r eval=FALSE, include=FALSE}
cellchat_adult@DB <- CellChatDB.use
```

```{r eval=FALSE, include=FALSE}
cellchat_adult <- subsetData(cellchat_adult)
future::plan("multisession", workers = 8) # do parallel
cellchat_adult <- identifyOverExpressedGenes(cellchat_adult)
cellchat_adult <- identifyOverExpressedInteractions(cellchat_adult)
```

```{r eval=FALSE, include=FALSE}
cellchat_adult <- computeCommunProb(cellchat_adult, type = "triMean")
cellchat_adult <- filterCommunication(cellchat_adult, min.cells = 10)
```

```{r eval=FALSE, include=FALSE}
df.net.adult <- subsetCommunication(cellchat_adult)
```

```{r eval=FALSE, include=FALSE}
cellchat_adult <- computeCommunProbPathway(cellchat_adult)
cellchat_adult <- aggregateNet(cellchat_adult)
cellchat_adult <- netAnalysis_computeCentrality(cellchat_adult)

groupSize <- as.numeric(table(cellchat_adult@idents))
par(mfrow = c(1,2), xpd=TRUE)
netVisual_circle(cellchat_adult@net$count, vertex.weight = groupSize, weight.scale = T, label.edge= F, title.name = "Number of interactions")
netVisual_circle(cellchat_adult@net$weight, vertex.weight = groupSize, weight.scale = T, label.edge= F, title.name = "Interaction weights/strength")
```

```{r eval=FALSE, include=FALSE}
object.list <- list(young = cellchat_young, adult = cellchat_adult)
qsave(object.list, paste0(PREPRDATADIR, 'cellchat_list_ATLAS.qs'))
```

```{r include=FALSE}
object.list <- qread(paste0(PREPRDATADIR, 'cellchat_list_ATLAS.qs'))
```

```{r echo=FALSE}
weight.max <- getMaxWeight(object.list, attribute = c("idents","count"))
par(mfrow = c(1,2), xpd=TRUE)
for (i in 1:length(object.list)) {
  netVisual_circle(object.list[[i]]@net$count, weight.scale = T, label.edge= F, edge.weight.max = weight.max[2], edge.width.max = 12, title.name = paste0("Number of interactions - ", names(object.list)[i]))
}
```

## Compare the total number of interactions and interaction strength

```{r eval=FALSE, include=FALSE}
object.list <- list(young = cellchat_young, adult = cellchat_adult)
cellchat <- mergeCellChat(object.list, add.names = names(object.list))
```

```{r eval=FALSE, include=FALSE}
qsave(cellchat, paste0(PREPRDATADIR, 'cellchat_ATLAS.qs'))
```

```{r include=FALSE}
cellchat <- qread(paste0(PREPRDATADIR, 'cellchat_ATLAS.qs'))
```

```{r echo=FALSE}
gg1 <- compareInteractions(cellchat, show.legend = F, group = c(1,2))
gg2 <- compareInteractions(cellchat, show.legend = F, group = c(1,2), measure = "weight")
gg1 + gg2
```

The differential number of interactions or interaction strength in the cell-cell communication network between two datasets can be visualized using circle plot, where red (or blue) colored edges represent increased (or decreased) signaling in the second dataset (adult) compared to the first one (young). We observe that although the overall interaction strength is higher in young mice, certain cell type interactions are stronger in adults (indicated in red on the plot). For example, interactions between fibroblasts and malignant cells, malignant cells and T cells, as well as myeloid cells and T cells, show higher strength in adult mice.

```{r, fig.show='hold', fig.width=12, fig.height=6}
# Suppress automatic printing with invisible()
invisible(par(mfrow = c(1,2), xpd=TRUE))
invisible(netVisual_diffInteraction(cellchat, weight.scale = T))
invisible(netVisual_diffInteraction(cellchat, weight.scale = T, measure = "weight"))
```

# Compare the major sources and targets in a 2D space

Identifying cell populations with significant changes in sending or receiving signals between different age.

```{r}
num.link <- sapply(object.list, function(x) {rowSums(x@net$count) + colSums(x@net$count)-diag(x@net$count)})
weight.MinMax <- c(min(num.link), max(num.link)) # control the dot size in the different datasets
gg <- list()
for (i in 1:length(object.list)) {
  gg[[i]] <- netAnalysis_signalingRole_scatter(object.list[[i]], title = names(object.list)[i], weight.MinMax = weight.MinMax) 
}

patchwork::wrap_plots(plots = gg)
```
Young tumors demonstrated high incoming signals in B cells. Adult tumors demonstrated high outgoing signals in Stromal cells, probably indicating a more tumor-supportive microenvironment, and low outgoing signals in malignant cells.

Identifying the signaling changes of specific cell populations

```{r fig.height=12, fig.width=16}
gg1 <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "Neuroendocrine")

gg2 <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "Myeloid_cells")

gg3 <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "Stromal_cells")

gg4 <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "Endothelial_cells")

gg5 <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "B_cells")

gg6 <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "T_cells")

patchwork::wrap_plots(plots = list(gg1,gg2,gg3,gg4,gg5,gg6), nrow=2)
```

```{r include=FALSE}

cellchat <- computeNetSimilarityPairwise(cellchat, type = "functional")
#> Compute signaling network similarity for datasets 1 2
cellchat <- netEmbedding(cellchat, type = "functional")
#> Manifold learning of the signaling networks for datasets 1 2
cellchat <- netClustering(cellchat, type = "functional")
#> Classification learning of the signaling networks for datasets 1 2
# Visualization in 2D-space
netVisual_embeddingPairwise(cellchat, type = "functional", label.size = 3.5)
#> 2D visualization of signaling networks from datasets 1 2
```

This plot illustrates the distance, representing the differences in communication networks between the two datasets in terms of functional similarity. A larger distance indicates a greater disparity. It is important to note that the distance calculation is based solely on the signaling pathways that are shared between the two datasets.

```{r fig.height=8, fig.width=6}
rankSimilarity(cellchat, type = "functional")
```

# Identify altered signaling with distinct interaction strength

By comparing the information flow/interaction strength of each signaling pathway, CellChat identifies signaling pathways that: (i) turn off, (ii) decrease, (iii) turn on, or (iv) increase, by changing their information flow at one condition as compared to another condition. Identifying the altered signaling pathways or ligand-receptor pairs based on the overall information. Significant differencw

### All pathways 

```{r fig.height=10, fig.width=8}
gg1 <- rankNet(cellchat, mode = "comparison", measure = "weight", sources.use = NULL, targets.use = NULL, stacked = T, do.stat = TRUE)
gg2 <- rankNet(cellchat, mode = "comparison", measure = "weight", sources.use = NULL, targets.use = NULL, stacked = F, do.stat = TRUE)

gg1 + gg2
```

### Drop outliers

```{r fig.height=10, fig.width=8}
# Get all pathways except the outliers
all_paths <- union(cellchat@netP$young$pathways, cellchat@netP$adult$pathways)
filtered_paths <- setdiff(all_paths, c("PTN", "LAMININ", "APP", "MIF", "COLLAGEN", "FN1"))

# Create plot with filtered pathways
gg1 <- rankNet(cellchat, mode = "comparison", measure = "weight", signaling = filtered_paths, sources.use = NULL, targets.use = NULL, stacked = T, do.stat = TRUE)
gg2 <- rankNet(cellchat, mode = "comparison", measure = "weight", signaling = filtered_paths, sources.use = NULL, targets.use = NULL, stacked = F, do.stat = TRUE)

gg1 + gg2
```

Our CellChat analysis revealed distinct signaling patterns between young and adult neuroblastoma microenvironments, providing insights into the accelerated tumor progression observed in adult mice:

**Adult-Specific Signaling Profile**

Demonstrated significantly enhanced activity in pathways associated with tumor progression:  

- Elevated CCL signaling
- Upregulated FGF pathway activity. Inappropriate expression of FGF and improper activation of FGFRs are associated with various pathologic conditions, unregulated cell growth, and tumorigenesis.
- Higher FLRT signaling, potentially facilitating cell migration and invasion  
- Enhanced ECM-related signaling (CLEC, MMP), suggesting active tissue remodeling. 
- PARs - All of these agents (TF, thrombin, PARs—mainly PAR-1 and PAR-2) are thought to promote cancer invasion and metastasis at least in part by facilitating tumor cell migration, angiogenesis, and interactions with host vascular cells, including platelets, fibroblasts, and endothelial cells lining blood vessels.

**Young-Specific Signaling Profile**

Exhibited stronger activity in regulatory and immune-related pathways:

- Enhanced MHC-I signaling, indicating stronger immune surveillance. 
- Elevated BMP signaling, suggesting better growth regulation. 
- Higher CD6, CD200 and TNF pathways activity, pointing to more controlled immune modulation. 

# Outgoing and incoming signalings

Identify the altered signaling pathways or ligand-receptor pairs based on the outgoing (or incoming) signaling patterns.

In this heatmap, colobar represents the relative signaling strength of a signaling pathway across cell groups (Note that values are row-scaled). The top colored bar plot shows the total signaling strength of a cell group by summarizing all signaling pathways displayed in the heatmap. The right grey bar plot shows the total signaling strength of a signaling pathway by summarizing all cell groups displayed in the heatmap.

```{r fig.height=11, fig.width=8}
i = 1

pathway.union <- union(object.list[[i]]@netP$pathways, object.list[[i+1]]@netP$pathways)
ht1 = netAnalysis_signalingRole_heatmap(object.list[[i]], pattern = "outgoing", signaling =  filtered_paths, title = names(object.list)[i], width = 5, height = 22)
ht2 = netAnalysis_signalingRole_heatmap(object.list[[i+1]], pattern = "outgoing", signaling =  filtered_paths, title = names(object.list)[i+1], width = 5, height = 22)
draw(ht1 + ht2, ht_gap = unit(0.5, "cm"))
```

**Malignant cells:**

- THBS - At the early stage of carcinogenesis, THBS1 and THBS2 secreted by normal tissue act as an antiangiogenic wall. However, with tumor progression, they may switch to enhance angiogenic phenotype in order to support tumor development and metastasis.  
- NOTCH - The Notch signaling pathway has been shown to inhibit neuroblastoma tumor cell differentiation [8–11], and Notch1 receptor expression has been shown to be associated with high-risk tumor features and poor prognosis in a cohort of children with neuroblastoma [11].    
- ANGPTL - Angiopoietin-like protein 1 (ANGPTL1) is a member of the ANGPTL family that suppresses angiogenesis, cancer invasion, metastasis, and cancer progression.    
- EPHA - ectopic high-level expression of EPHA2 in N-type neuroblastoma cell lines resulted in significant growth suppression. 
- ANGPT - angiogenesis. 
- IL1 - IL-1 has been shown to promote tumor growth. 
- CD137 - CD137 is a kind of costimulatory molecule that mediates T cell activation, which regulates the activity of immune cells in a variety of physiological and pathological processes. 

**T cells:**

- MHC-1  
- CD 52 - Studies clearly showed that CD52 can modulate T-cell activation either by its intracellular signal pathways or by the interaction of soluble CD52 and Siglec-10 expressing on T cells.  
- SELL - CD62L/L-selectin is a marker found on naïve T cells. 
- CCL. 
- CD6 - CD6 mainly functions as an adhesion and an intracellular signal-transducing receptor and participates in the formation and stabilization of T-cell contacts with antigen-presenting cells.   


```{r fig.height=11, fig.width=8}
ht1 = netAnalysis_signalingRole_heatmap(object.list[[i]], pattern = "incoming", signaling = filtered_paths, title = names(object.list)[i], width = 5, height = 22, color.heatmap = "GnBu")
ht2 = netAnalysis_signalingRole_heatmap(object.list[[i+1]], pattern = "incoming", signaling = filtered_paths, title = names(object.list)[i+1], width = 5, height = 22, color.heatmap = "GnBu")
draw(ht1 + ht2, ht_gap = unit(0.5, "cm"))
```

MK incoming signaling exists in young mice only in all cell types, but outgoing MK signsling is in young Fibroblasts only

# Dif expression

## TGFb signaling

### Young is coral

```{r fig.height=6, fig.width=8}
cellchat@meta$datasets = factor(cellchat@meta$datasets, levels = c("young", "adult")) # set factor level
plotGeneExpression(cellchat, signaling = "TGFb", split.by = "datasets", colors.ggplot = T, type = "violin") 

# seu$age <- factor(seu$age, levels = c("young", "adult"))
# extractEnrichedLR(cellchat, signaling = "TGFb", geneLR.return = T)$geneLR
# VlnPlot(seu, features = 'Tgfb1', split.by = "age")
```

## PECAM2 signaling

```{r fig.height=4, fig.width=6}
plotGeneExpression(cellchat, signaling = "PECAM2", split.by = "datasets", colors.ggplot = T, type = "violin") 
```

## CCL signaling

Increased CCL2, CCL9, CCL6 in adult myeloid cells:

CCL2: Known to recruit immunosuppressive cells like MDSCs (Myeloid-Derived Suppressor Cells) and M2 macrophages
High expression suggests more immunosuppressive myeloid cell recruitment in adult tumors

Higher CCL5 in adult T cells:

CCL5 can paradoxically promote tumor growth by:

- Recruiting Tregs (regulatory T cells)
- Supporting immunosuppressive microenvironment
- Enhancing tumor cell survival

The elevated chemokine expression in adult tumors likely creates a more immunosuppressive microenvironment through enhanced recruitment of regulatory immune cells, potentially explaining the more aggressive tumor growth in adult mice. 

```{r fig.height=7, fig.width=12}
# Chord diagram
pathways.show <- c("CCL") 
par(mfrow = c(1,2), xpd=TRUE)
for (i in 1:length(object.list)) {
  netVisual_aggregate(object.list[[i]], signaling = pathways.show, layout = "chord", signaling.name = paste(pathways.show, names(object.list)[i]))
}
```

```{r fig.height=10, fig.width=8}
plotGeneExpression(cellchat, signaling = "CCL", split.by = "datasets", colors.ggplot = T, type = "violin") 
```

## MHC_I signaling

The critical observation is that malignant cells lack both MHC-I and MHC-II expression in both age groups, which is highly significant as MHC-I should normally be expressed by all nucleated cells. This dual absence of MHC molecules on tumor cells suggests a strong immune evasion mechanism through "hiding" from both CD4+ and CD8+ T cell recognition, potentially contributing to tumor progression. The presence of CD8a and CD8b1 exclusively in T cells from young mice (coral) indicates that while cytotoxic T cells are present in young tumors, they may be unable to recognize tumor cells due to missing MHC-I, creating a "missing self" scenario. The higher expression of various H2 genes in immune and stromal cells in adult condition (teal) might represent a compensatory mechanism, but without MHC expression on tumor cells, effective anti-tumor immunity would be severely compromised.

```{r fig.height=8, fig.width=8}
plotGeneExpression(cellchat, features = c('H2-M3', "H2-Q7", "H2-T22", "H2-T23", "H2-D1", "H2-K1", "H2-Q4", "Cd8a", "Cd8b1"), split.by = "datasets", colors.ggplot = T, type = "violin") 
```

## SELL signaling

```{r fig.height=6, fig.width=8}
plotGeneExpression(cellchat, signaling = "SELL", split.by = "datasets", colors.ggplot = T, type = "violin") 
```

# Conclusion



```{r include=FALSE}
# define a positive dataset, i.e., the dataset with positive fold change against the other dataset
pos.dataset = "young"
# define a char name used for storing the results of differential expression analysis
features.name = paste0(pos.dataset, ".merged")

# perform differential expression analysis 
# Of note, compared to CellChat version < v2, CellChat v2 now performs an ultra-fast Wilcoxon test using the presto package, which gives smaller values of logFC. Thus we here set a smaller value of thresh.fc compared to the original one (thresh.fc = 0.1). Users can also provide a vector and dataframe of customized DEGs by modifying the cellchat@var.features$LS.merged and cellchat@var.features$LS.merged.info. 

cellchat <- identifyOverExpressedGenes(cellchat, group.dataset = "datasets", pos.dataset = pos.dataset, features.name = features.name, only.pos = FALSE, thresh.pc = 0.1, thresh.fc = 0.05,thresh.p = 0.05, group.DE.combined = FALSE) 
#> Use the joint cell labels from the merged CellChat object

# map the results of differential expression analysis onto the inferred cell-cell communications to easily manage/subset the ligand-receptor pairs of interest
net <- netMappingDEG(cellchat, features.name = features.name, variable.all = TRUE)
# extract the ligand-receptor pairs with upregulated ligands in LS
net.up <- subsetCommunication(cellchat, net = net, datasets = "young",ligand.logFC = 0.05, receptor.logFC = NULL)
# extract the ligand-receptor pairs with upregulated ligands and upregulated receptors in NL, i.e.,downregulated in LS
net.down <- subsetCommunication(cellchat, net = net, datasets = "adult",ligand.logFC = -0.05, receptor.logFC = NULL)


```

```{r include=FALSE}
gene.up <- extractGeneSubsetFromPair(net.up, cellchat)
gene.down <- extractGeneSubsetFromPair(net.down, cellchat)

# df <- findEnrichedSignaling(object.list[[2]], features = c("CCL19", "CXCL12"), idents = c("Inflam. FIB", "COL11A1+ FIB"), pattern ="outgoing")
```

```{r echo=FALSE, fig.height=10, fig.width=10}
pairLR.use.up = net.up %>% 
  filter(abs(ligand.logFC) > 0.2 | abs(receptor.logFC) > 0.2) %>% 
  .[, "interaction_name", drop = F]
gg1 <- netVisual_bubble(cellchat, pairLR.use = pairLR.use.up, sources.use = 1, targets.use = 1, comparison = c(1, 2),  angle.x = 90, remove.isolate = T,title.name = paste0("Down-regulated signaling in ", names(object.list)[1]))
#> Comparing communications on a merged object
pairLR.use.down = net.down %>% 
  filter(abs(ligand.logFC) > 0.2 | abs(receptor.logFC) > 0.2) %>% 
  .[, "interaction_name", drop = F]
gg2 <- netVisual_bubble(cellchat, pairLR.use = pairLR.use.down, sources.use = 1, targets.use = 1, comparison = c(1, 2),  angle.x = 90, remove.isolate = T,title.name = paste0("Up-regulated signaling in ", names(object.list)[1]))
#> Comparing communications on a merged object
gg1 + gg2
```



```{r eval=FALSE, include=FALSE}
# Chord diagram
par(mfrow = c(1,2), xpd=TRUE)
netVisual_chord_gene(object.list[[2]], sources.use = 4, targets.use = c(5:11), slot.name = 'net', net = net.up, lab.cex = 0.8, small.gap = 3.5, title.name = paste0("Up-regulated signaling in ", names(object.list)[2]))
netVisual_chord_gene(object.list[[1]], sources.use = 4, targets.use = c(5:11), slot.name = 'net', net = net.down, lab.cex = 0.8, small.gap = 3.5, title.name = paste0("Down-regulated signaling in ", names(object.list)[2]))
#> You may try the function `netVisual_chord_cell` for visualizing individual signaling pathway
```

# Fibroblasts

It isn't full plot. Full plot was presented above

```{r echo=FALSE, fig.height=7, fig.width=6}
fibr_pl <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "Fibroblasts", ylims = c(-0.03, 0.03))
fibr_pl
```

### Adult specific pathways in Fibroblasts

```{r}
fibr_pl$data %>% 
  filter(specificity == 'adult specific') 
```

```{r fig.height=7, fig.width=12}
# Chord diagram
pathways.show <- c("VEGF") 
par(mfrow = c(1,2), xpd=TRUE)
for (i in 1:length(object.list)) {
  netVisual_aggregate(object.list[[i]], signaling = pathways.show, layout = "chord", signaling.name = paste(pathways.show, names(object.list)[i]))
}
```

```{r fig.height=4, fig.width=8}
plotGeneExpression(cellchat, features = c('Vegfd','Flt4'), split.by = "datasets", colors.ggplot = T, type = "violin") 
```

EphB signaling supports the formation of epithelial cell junctions through an interaction with E-cadherin

```{r fig.height=7, fig.width=12}
# Chord diagram
pathways.show <- c("EPHB") 
par(mfrow = c(1,2), xpd=TRUE)
for (i in 1:length(object.list)) {
  netVisual_aggregate(object.list[[i]], signaling = pathways.show, layout = "chord", signaling.name = paste(pathways.show, names(object.list)[i]))
}
```

```{r fig.height=3, fig.width=8}
plotGeneExpression(cellchat, features = c('Efnb1'), split.by = "datasets", colors.ggplot = T, type = "violin") 
```


# Malignant cells

```{r echo=FALSE, fig.height=7, fig.width=6}
malig_pl <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "Malignant cells", ylims = c(-0.03, 0.02))
malig_pl
```

### Adult specific pathways in Fibroblasts

```{r}
malig_pl$data %>% 
  filter(specificity == 'adult specific') 
```

```{r fig.height=7, fig.width=12}
# Chord diagram
pathways.show <- c("FGF") 
invisible(netVisual_aggregate(object.list[[2]], signaling = pathways.show, layout = "chord", signaling.name = paste(pathways.show, names(object.list)[2])))

```

```{r fig.height=4, fig.width=8}
plotGeneExpression(cellchat, features = c('Fgf1', 'Fgf2', 'Fgfr1', 'Fgfr2'), split.by = "datasets", colors.ggplot = T, type = "violin") 
```

```{r fig.height=7, fig.width=12}
# Chord diagram
pathways.show <- c("PARs") 
invisible(netVisual_aggregate(object.list[[2]], signaling = pathways.show, layout = "chord", signaling.name = paste(pathways.show, names(object.list)[2])))

```

```{r fig.height=6, fig.width=8}
plotGeneExpression(cellchat, signaling = 'PARs', split.by = "datasets", colors.ggplot = T, type = "violin") 
```






