---
title: "Cytof"
author: "Kate Petrenko"
date: "2024-10-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set()
# BiocManager::install("CATALYST")
library(CATALYST)    # For CyTOF data preprocessing
library(flowCore)    # Basic flow cytometry data handling
library(flowWorkspace) # Additional flow cytometry tools
library(openCyto)    # Automated gating methods
library(flowAI)
library(ggplot2)     
library(gridExtra)   
library(dplyr)      
library(readr)     
library(tidyr) 
library(cowplot)
# BiocManager::install("diffcyt")
library(diffcyt)
```

```{r global variables, include=FALSE}
set.seed(100)
DATADIR = '~/neuroblastoma/data/'
WORKDIR = '~/neuroblastoma/'
PREPRDATADIR = '~/neuroblastoma/preprocessing_data/'

my_colors <- c("#9ACD32FF", "#4682B4FF",  "#DDA0DDFF", "#FFA07AFF", "#8FBC8BFF",
               "#40E0D0FF",  "#F0E68CFF", "#5F9EA0FF", "#D2B48CFF",  
               "#6495EDFF", "#FF69B4FF", "#BA55D3FF",  "#F08080FF", "#32CD32FF",  
               "#FFDAB9FF",  "#87CEEBFF")
```

# Read Data

```{r include=FALSE}
fcs_files_fs1 <- list.files(path = paste0(DATADIR, "cytof_data"), pattern = "LIC.fcs", full.names = TRUE)
fcs_files_fs2 <- list.files(path = paste0(DATADIR, "cytof_data"), pattern = "cells.fcs", full.names = TRUE)
```


```{r message=FALSE, warning=FALSE, include=FALSE}
markers <- read_table('cytof_markers.tsv') %>% as.data.frame()
markers$Name <- sapply(markers$Metal, function(x) {
  # Extract the number and element
  num <- gsub("[A-Za-z]", "", x)  # Extract numbers
  element <- gsub("[0-9]", "", x)  # Extract letters
  
  # Format it like fs1's column names (ElementNumDi)
  paste0(tools::toTitleCase(tolower(element)), num, "Di")
})
markers$marker_class = "type"

# Add technical markers to your existing markers data frame
technical_markers <- data.frame(
    Metal = c("190BCKG", "191Ir", "193Ir", "140Ce", "103Rh"),
    Marker = c("Background", "DNA1", "DNA2", "Beads", "Rhodium"),
    Name = c("BCKG190Di", "Ir191Di", "Ir193Di", "Ce140Di", "Rh103Di"),
    marker_class = 'none',
    stringsAsFactors = FALSE
)

# Combine with existing markers
markers <- rbind(markers, technical_markers)

# Map names to markers
panel <- data.frame(fcs_colname = markers$Name, antigen = markers$Marker, marker_class = markers$marker_class)
```

# Metadata

```{r include=FALSE}
# Define metadata for fs1
md_fs1 <- data.frame(
    file_name = basename(fcs_files_fs1),
    sample_id = gsub("^\\d+_(.*?)_QC_.*\\.fcs$", "\\1", basename(fcs_files_fs1)),
    tumor = ifelse(grepl("NTB", basename(fcs_files_fs1)), "Non-tumor", "Tumor"),
    condition = ifelse(grepl("_P_", basename(fcs_files_fs1)), "Pups", "Adult"),
    tissue = case_when(
        grepl("_B_", basename(fcs_files_fs1)) ~ "Bone marrow",
        grepl("_BM_", basename(fcs_files_fs1)) ~ "Bone marrow",
        grepl("_T_", basename(fcs_files_fs1)) ~ "Tumor",
        grepl("_S_", basename(fcs_files_fs1)) ~ "Spleen"
    ),
    stringsAsFactors = FALSE
)

# Define metadata for fs2
md_fs2 <- data.frame(
    file_name = basename(fcs_files_fs2),
    sample_id = gsub("^export_\\d+_(.*?)_.*\\.fcs$", "\\1", basename(fcs_files_fs2)),
    condition = case_when(
        grepl("ABM|AS", basename(fcs_files_fs2)) ~ "Adult",
        grepl("PBM|PS", basename(fcs_files_fs2)) ~ "Pups"
    ),
    tissue = case_when(
        grepl("ABM|PBM", basename(fcs_files_fs2)) ~ "Bone marrow",
        grepl("AS|PS", basename(fcs_files_fs2)) ~ "Spleen"
    ),
    stringsAsFactors = FALSE
)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
fs1 <- read.flowSet(fcs_files_fs1)
fs2 <- read.flowSet(fcs_files_fs2)
```

```{r include=FALSE}
# Create SingleCellExperiment objects
sce_fs1 <- prepData(fcs_files_fs1, panel = panel %>% dplyr::filter(fcs_colname %in% colnames(fs1)), md = md_fs1, md_cols = list(factors = c("condition", "tissue", "tumor")), normalize = FALSE)
sce_fs2 <- prepData(fcs_files_fs2, panel = panel %>% dplyr::filter(fcs_colname %in% colnames(fs2)), md = md_fs2, md_cols = list(factors = c("condition", "tissue")), normalize = FALSE)
```

```{r eval=FALSE, include=FALSE}
plotExprs(sce_fs1, features = c('CD4', 'CD3', 'CD8', 'CD45'))
```

# Tumor dataset

```{r echo=FALSE}
plotCounts(sce_fs1, group_by = "sample_id", color_by = "condition")
```

The MDS plot reveals clear patterns of separation in the data. We can see that the main differenci in our dataset is the tissue.

```{r echo=FALSE}
pbMDS(sce_fs1, color_by = "condition", label_by = "sample_id", shape_by = 'tumor')
```

The heatmap shows median marker intensities with clustered columns (markers) and rows (samples).

```{r echo=FALSE, fig.height=4, fig.width=9}
plotExprHeatmap(sce_fs1, scale = "last",
    hm_pal = rev(hcl.colors(10, "YlGnBu")))
```

In this step, we identify the ability of markers to explain the variance observed in each sample. In particular, we calculate the PCA-based non-redundancy score (NRS). Markers with higher score explain a larger portion of variability present in a given sample.

```{r}
plotNRS(sce_fs1, features = "type", color_by = "condition")
```


## Clustering

```{r include=FALSE}
sce_fs1 <- cluster(sce_fs1, features = "type",
    xdim = 10, ydim = 10, maxK = 20, seed = 100)
```

```{r eval=FALSE, fig.height=9, fig.width=10, include=FALSE}
plotExprHeatmap(sce_fs1, features = "type", 
    by = "cluster_id", k = "meta20", 
    bars = TRUE, perc = TRUE)
```



```{r echo=FALSE}
sce_fs1 <- runDR(sce_fs1, "UMAP", cells = 1e3, features = "type")

plotDR(sce_fs1, "UMAP", color_by = "meta20")
plotDR(sce_fs1, "UMAP", color_by = "meta20", facet_by = "condition")
plotDR(sce_fs1, "UMAP", color_by = "meta20", facet_by = "tumor")
```

## Annotation

```{r include=FALSE}
merging_table1 <- data.frame(
  original_cluster = seq(1, 20),
  new_cluster = c('Myeloid cells',
                  'Myeloid cells',
                  'Myeloid cells',
                  'T cells CD4+',
                  'Non-immune cells',
                  'Myeloid cells',
                  'T cells CD8+',
                  'Myeloid cells',
                  'Myeloid cells',
                  'Unknown',
                  'Myeloid cells',
                  'Unknown',
                  'Myeloid cells',
                  'Myeloid cells',
                  'B cells',
                  'Myeloid cells',
                  'Myeloid cells',
                  'Unknown',
                  'B cells',
                  'B cells'
                  )
)

sce_fs1 <- mergeClusters(sce_fs1, k = "meta20", 
    table = merging_table1, id = "merging1", overwrite = T)
```

```{r echo=FALSE}
plotDR(sce_fs1, "UMAP", color_by = "merging1")
plotDR(sce_fs1, "UMAP", color_by = "merging1", facet_by = 'condition')
```

This heatmap provides a visual representation of marker expression across different clusters. Each column corresponds to a marker gene, while each row represents a cluster identified in the dataset. The intensity of the colors indicates the median scaled expression of each marker within a given cluster, with red signifying high expression and blue signifying low expression.

The heatmap allows for the identification of cluster-specific markers, enabling the characterization of clusters based on their unique expression patterns. I used them to annotate clusters so you can use it to check me.

```{r echo=FALSE, fig.height=9, fig.width=10}
plotExprHeatmap(sce_fs1, features = "type", 
    by = "cluster_id", k = "meta20", m = "merging1")
```

That's combined heatmap by celltypes.

```{r echo=FALSE, fig.height=7, fig.width=11}
plotExprHeatmap(sce_fs1, features = "type", 
    by = "cluster_id", k = "merging1",
    bars = TRUE, perc = TRUE)
```

## Analysis

### Differential Abundances Analysis

```{r echo=FALSE}
plotAbundances(sce_fs1, k = "merging1", by = "sample_id", group_by = 'tissue')
```

It may be quite hard to see the differences in cluster abundances in the plot above, especially for clusters with very low frequency. And, since boxplots cannot represent multimodal distributions, we show boxplots with jittered points of the sample-level cluster proportions overlaid

```{r}
plotAbundances(sce_fs1, k = "merging1", by = "cluster_id", shape_by = "sample_id")
```

```{r message=FALSE, warning=FALSE, include=FALSE}
ei <- metadata(sce_fs1)$experiment_info
(da_formula1 <- createFormula(ei, 
    cols_fixed = "condition", 
    cols_random = "sample_id"))

(da_formula2 <- createFormula(ei, 
    cols_fixed = "condition", 
    cols_random = c("sample_id", "tissue")))

contrast <- createContrast(c(0, 1))
```

```{r message=FALSE, warning=FALSE, include=FALSE}
da_res2 <- diffcyt(sce_fs1, 
    formula = da_formula2, contrast = contrast,
    analysis_type = "DA", method_DA = "diffcyt-DA-GLMM",
    clustering_to_use = "merging1", verbose = FALSE)
```

```{r include=FALSE}
FDR_cutoff = 0.05
rowData(da_res2$res) 
table(rowData(da_res2$res)$p_adj < FDR_cutoff)
```

```{r echo=FALSE}
sce_fs1$sample_id <- factor(sce_fs1$sample_id, levels = c('TB_A_BM', 'TB_P_BM', 'NTB_P_B',  
                                                          'TB_A_S', 'TB_P_S', 'NTB_P_S',
                                                          'TB_A_T', 'TB_P_T'))
plotDiffHeatmap(sce_fs1, rowData(da_res2$res), all = TRUE, fdr = FDR_cutoff)
```

# Naive dataset

The MDS plot reveals clear patterns of separation in the data. Dim1 predominantly distinguishes samples based on tissue type, indicating a strong tissue-specific effect. In contrast, Dim2 differentiates samples by age, reflecting age-associated variation. This separation along Dim2 highlights notable differences between age groups in the dataset.

```{r echo=FALSE}
pbMDS(sce_fs2, color_by = "condition", label_by = "sample_id", shape_by = 'tissue')
```

The heatmap shows median marker intensities with clustered columns (markers) and rows (samples).

```{r echo=FALSE, fig.height=4, fig.width=9}
plotExprHeatmap(sce_fs2, scale = "last",
    hm_pal = rev(hcl.colors(10, "YlGnBu")))
```

In this step, we identify the ability of markers to explain the variance observed in each sample. In particular, we calculate the PCA-based non-redundancy score (NRS). Markers with higher score explain a larger portion of variability present in a given sample.

```{r echo=FALSE, fig.height=4, fig.width=7}
plotNRS(sce_fs2, features = "type", color_by = "condition")
```

## Clustering

```{r message=FALSE, warning=FALSE, include=FALSE}
sce_fs2 <- cluster(sce_fs2, features = "type",
    xdim = 10, ydim = 10, maxK = 30, seed = 100)
```

```{r eval=FALSE, fig.height=6, fig.width=10, include=FALSE}
plotExprHeatmap(sce_fs2, features = "type", 
    by = "cluster_id", k = "meta30", 
    bars = TRUE, perc = TRUE)
```

```{r fig.height=6, fig.width=10}
# 1. Get all type markers
type_markers <- type_markers(sce_fs2)

# 2. Reorder them as you wish (example)
my_ordered_markers <- c("CD45", "CD44", "CD43", "PD-L1", "Sca-1", "CD127", 
                        "CD5", "CD3", "CD62L", "CD8", "CD4", "Ia-Ie", "CD45R",
                        "CD19", "CD21", "IgM", "CD79b", "IgD", "CD24", "Ly6c",
                        "CD138", "TCRyd", "CD11b", "Ly6g", "CD64", "F4/80",
                        "CD69", "NK1.1", "IFN-gr", "SiglecF", "CD115", "CD25", "PDL2",
                        "CD48", "CD28", "CCR4", "CD93", "CD11c", "CD117")  # <- your preferred order

# 3. Plot with explicit feature order and disable column clustering
plotExprHeatmap(sce_fs2,
    features = my_ordered_markers,
    by = "cluster_id", 
    k = "meta30", 
    col_clust = FALSE,   # disables automatic column reordering
    bars = TRUE, 
    perc = TRUE)
```

```{r echo=FALSE}
sce_fs2 <- runDR(sce_fs2, "UMAP", cells = 1e3, features = "type")

plotDR(sce_fs2, "UMAP", color_by = "meta30")
plotDR(sce_fs2, "UMAP", color_by = "meta30", facet_by = "condition")
plotDR(sce_fs2, "UMAP", color_by = "meta30", facet_by = "tissue")
```

## Annotation

```{r include=FALSE}
merging_table <- data.frame(
  original_cluster = seq(1, 30),
  new_cluster = c('Cytotoxic T-cells (CD8)', # 1
                  'Plasmacytoid DCs',
                  'Unknown',
                  'Unknown',
                  'Myeloid cells', # 5 Plasmablast
                  'Unknown',
                  'Unknown',
                  'Myeloid cells',
                  'T helper cells (CD4)',
                  'T helper cells (CD4)', # 10
                  'Plasmacytoid DCs',
                  'Unknown',
                  'Unknown', # HSC
                  'Myeloid cells',
                  'Unknown', # 15
                  'Unknown',
                  'LS-K',
                  'Unknown',
                  'Myeloid cells',
                  'T helper cells (CD4)', # 20 
                  'Activated B cells', # NK
                  'Cytotoxic T-cells (CD8)',
                  'Myeloid cells',
                  'Activated B cells',
                  'Unknown', # 25
                  'Unknown',
                  'NK',
                  'gd T cells', #
                  'Activated B cells',
                  'Activated B cells' # 30 
                  )
)

sce_fs2 <- mergeClusters(sce_fs2, k = "meta30", 
    table = merging_table, id = "merging1", overwrite = T)
```

```{r echo=FALSE, fig.height=5, fig.width=10}
plotDR(sce_fs2, "UMAP", color_by = "merging1")
plotDR(sce_fs2, "UMAP", color_by = "merging1", facet_by = 'condition')
```

This heatmap provides a visual representation of marker expression across different clusters. Each column corresponds to a marker gene, while each row represents a cluster identified in the dataset. The intensity of the colors indicates the median scaled expression of each marker within a given cluster, with red signifying high expression and blue signifying low expression.

The heatmap allows for the identification of cluster-specific markers, enabling the characterization of clusters based on their unique expression patterns. I used them to annotate clusters so you can use it to check me.

```{r echo=FALSE, fig.height=9, fig.width=10}
plotExprHeatmap(sce_fs2, features = "type", 
    by = "cluster_id", k = "meta30", m = "merging1")
```

That's combined heatmap by celltypes.

```{r echo=FALSE, fig.height=7, fig.width=11}
plotExprHeatmap(sce_fs2, features = "type", 
    by = "cluster_id", k = "merging1",
    bars = TRUE, perc = TRUE)
```

## Analysis

### Differential Abundances Analysis

```{r eval=FALSE, include=FALSE}
qsave(sce_fs2, paste0(PREPRDATADIR, 'cytof_naive.qs'))
```

```{r}
sce_fs2 <- qread(paste0(PREPRDATADIR, 'cytof_naive.qs'))
```

```{r echo=FALSE}
plotAbundances(sce_fs2, k = "merging1", by = "sample_id", group_by = 'tissue')
```

```{r echo=FALSE}
plotAbundances(sce_fs2, k = "merging1", by = "cluster_id", shape_by = "sample_id")
```

```{r message=FALSE, warning=FALSE, include=FALSE}
ei <- metadata(sce_fs2)$experiment_info
(da_formula1 <- createFormula(ei, 
    cols_fixed = "condition", 
    cols_random = "sample_id"))

(da_formula2 <- createFormula(ei, 
    cols_fixed = "condition", 
    cols_random = c("sample_id", "tissue")))

contrast <- createContrast(c(0, 1))
```

```{r message=FALSE, warning=FALSE, include=FALSE}
da_res2 <- diffcyt(sce_fs2, 
    formula = da_formula2, contrast = contrast,
    analysis_type = "DA", method_DA = "diffcyt-DA-GLMM",
    clustering_to_use = "merging1", verbose = FALSE)
```

```{r include=FALSE}
FDR_cutoff = 0.05
rowData(da_res2$res) 
table(rowData(da_res2$res)$p_adj < FDR_cutoff)
```

```{r echo=FALSE}
sce_fs2$sample_id <- factor(sce_fs2$sample_id, levels = c('ABM', 'PBM', 'AS', 'PS'))
plotDiffHeatmap(sce_fs2, rowData(da_res2$res), all = TRUE, fdr = FDR_cutoff)
```



```{r eval=FALSE, include=FALSE}
# Let's fix marker classifications
all_markers <- rownames(sce_fs2)

marker_classes <- list(
    # Technical markers/channels
    technical = c("Background", "DNA1", "DNA2", "Beads", "Rhodium",
                 "Time", "Event_length", "Center", "Offset", "Width", "Residual"),
    
    # Markers used primarily for cell type identification/annotation
    type = c("CD3", "CD4", "CD8", "CD19", "CD45", "CD45R",
             "CD11c", "F4/80", "Ly6c", "Ly6g", "NK1.1", "TCRyd",
             "CD64", "CD11b", "CD24", "CD138", "CD21", "CD127",
             "Sca-1", "CD115", "SiglecF", "CD79b"),
    
    # Markers for activation/function - these are what we want to test
    state = c("CCR4", "PDL2", "CD25", "IgM", "PD-L1", "IFN-gr",
              "CD28", "CD5", "CD93", "CD117", "CD62L", "CD44",
              "CD43", "CD69", "Ia-Ie", "IgD", "CD48")
)

# Update marker classes in rowData
rowData(sce_fs2)$marker_class <- "type"  # Default as type
rowData(sce_fs2)$marker_class[rownames(sce_fs2) %in% marker_classes$technical] <- "none"
rowData(sce_fs2)$marker_class[rownames(sce_fs2) %in% marker_classes$state] <- "state"


ds_res1 <- diffcyt(sce_fs2, 
    formula = ds_formula2, contrast = contrast,
    analysis_type = "DS", method_DS = "diffcyt-DS-LMM",
    clustering_to_use = "merging1", verbose = FALSE)

table(rowData(ds_res1$res)$p_adj < FDR_cutoff)

topTable(ds_res1, top_n = 5, order_by = "cluster_id", 
    show_meds = TRUE, format_vals = TRUE, digits = 3)

table(rowData(ds_res1$res)$p_adj < FDR_cutoff)
```

```{r eval=FALSE, fig.height=11, fig.width=7, include=FALSE}
df1 <-  rowData(ds_res1$res)
df1$p_val <- 1
df1$p_adj <- 1
plotDiffHeatmap(sce_fs2, df1, top_n = 60, all = T)
```

