---
title: "ECM"
author: "Kate Petrenko"
date: "2024-11-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
library(data.table)
library(Seurat)
library(scDblFinder)
library(DoubletFinder)
library(dplyr)
library(remotes)
# remotes::install_github("carmonalab/STACAS")
library(STACAS)
library(SingleR)
# install.packages("scGate")
library(scGate)
library(ggplot2)
library(BiocParallel)
library(harmony)
library(RColorBrewer)
library(stringr)
library(cowplot)
library(scuttle)
# install.packages("devtools")
# BiocManager::install("Homo.sapiens")
# BiocManager::install("jpmam1/scalop") #forked copy of a really good guy 
# devtools::install_github("jlaffy/infercna")
library(infercna)
# library(infercnv)
library(parallel)
library(future)
library(clustree)
library(dittoSeq)
# install.packages("ggh4x")
library(ggh4x)
#BiocManager::install('EnhancedVolcano')
library(EnhancedVolcano)
library(clusterProfiler)
# BiocManager::install('org.Mm.eg.db')
library("org.Mm.eg.db")
library(ReactomePA)
library(msigdbr)
# install.packages('qs')
library(qs)
# devtools::install_github("ncborcherding/escape")
# devtools::install_github("rcastelo/GSVA")
library(escape)
library(biomaRt)
library(tidyverse)
```

```{r global variables, include=FALSE}
set.seed(100)
DATADIR = '~/neuroblastoma/data/'
WORKDIR = '~/neuroblastoma/'
PREPRDATADIR = '~/neuroblastoma/preprocessing_data/'

my_colors <- c("#9ACD32FF", "#4682B4FF",  "#DDA0DDFF", "#FFA07AFF", "#8FBC8BFF",
               "#40E0D0FF",  "#F0E68CFF", "#5F9EA0FF", "#D2B48CFF",  
               "#6495EDFF", "#FF69B4FF", "#BA55D3FF",  "#F08080FF", "#32CD32FF",  
               "#FFDAB9FF",  "#87CEEBFF")
```

```{r download references, message=FALSE, warning=FALSE, include=FALSE}
mouse_ref <- celldex::MouseRNAseqData()
mouse_ref_imm <- celldex::ImmGenData()
scGate_models_DB <- get_scGateDB()
```

Let me remind you what our data looks like :)

```{r echo=FALSE, fig.height=5, fig.width=12}
seu <- qread(paste0(PREPRDATADIR, 'seu_mouse.qs'))

Idents(object = seu) <- "RNA_snn_res.0.1"

new.cluster.ids <- c('0'='Malignant cells',
                     '1'='Malignant cells',
                     '2'='T cells',
                     '3'='B cells',
                     '4'='Myeloid cells',
                     '5'='Fibroblasts',
                     '6'='Malignant cells',
                     '6'='Malignant cells',
                     '7'='Endothelial cells',
                     '8'='Fibroblasts'
)

seu <- RenameIdents(seu, new.cluster.ids)
seu$annotation <- Idents(seu)

plot_grid(DimPlot(seu, cols = c("#FF69B4FF", "#40E0D0FF", "#6495EDFF",  "#9ACD32FF", "#BA55D3FF", "#DDA0DDFF")) +
  ggtitle("Annotation"), 
  DimPlot(seu, group.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) +
  ggtitle("Age"),
  nrow = 1, rel_widths = c(1, 0.9))
```

# ECM check

My goal was to check all ECM genes that Yuval sent us in our data. There 1027 genes at all. 710 genes are represented in our data. All genes were devided in 5 groups as it was in the original table:
1. ECM Glycoproteins       
2. Collagens               
3. Proteoglycans           
4. ECM-affiliated Proteins
5. ECM Regulators          
6. Secreted Factors


```{r}
esm_genes_df <- read.csv(paste0(WORKDIR, 'matrisome.csv'), sep = ',', header = T)
```

```{r}
convert_human_to_mouse <- function(gene_list) {
    output = c()
    mouse_human_genes = read.csv("https://www.informatics.jax.org/downloads/reports/HOM_MouseHumanSequence.rpt",sep="\t")

    for(gene in gene_list) {
          class_key = (mouse_human_genes %>% filter(Symbol == gene & Common.Organism.Name == "human"))[['DB.Class.Key']]
          if( !identical(class_key, integer(0)) ) {
            human_genes = (mouse_human_genes %>% filter(DB.Class.Key == class_key & Common.Organism.Name=="mouse, laboratory"))[,"Symbol"]
            for(human_gene in human_genes) {
                output = rbind(c(gene, human_gene), output)
            }
          }
     }
     return (output)
}

esm_genes_df.mouse <- convert_human_to_mouse(esm_genes_df$Gene.Symbol)[,2]
```

## Enrichment 

Firstly, I decided to calculate module score for for each cell, representing the aggregate expression of a predefined set of genes (a module - each groupd of genes in our case) relative to the background gene expression.

```{r fig.height=5, fig.width=8}
# Add a module score
seu <- AddModuleScore(seu, features = list(esm_genes_df.mouse), name = "ECM_Module")

# Plot module score by condition
# VlnPlot(seu, features = "ECM_Module1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM all genes Score by Condition") +
#   theme_minimal()

dittoBoxPlot(seu, "ECM_Module1", group.by = "age", split.by = 'annotation', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             jitter.size = NA, boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5) +
  ggtitle('ECM_Module - all genes')
```

```{r fig.height=5, fig.width=8}
# Add a module score
seu <- AddModuleScore(seu, features = list(convert_human_to_mouse(esm_genes_df %>% 
                                                                    dplyr::filter(Category == 'ECM Glycoproteins') %>% .$Gene.Symbol)[,2]), 
                      name = "ECM_Glycoproteins")

# Plot module score by condition
# VlnPlot(seu, features = "ECM_Glycoproteins1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_Glycoproteins Module Score by Condition") +
#   theme_minimal()

dittoBoxPlot(seu, "ECM_Glycoproteins1", group.by = "age", split.by = 'annotation', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             jitter.size = NA, boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

```{r fig.height=5, fig.width=8}
# Add a module score
seu <- AddModuleScore(seu, features = list(convert_human_to_mouse(esm_genes_df %>% 
                                                                    dplyr::filter(Category == 'Collagens')%>% .$Gene.Symbol)[,2]),
                      name = "ECM_Collagens")

# Plot module score by condition
# VlnPlot(seu, features = "ECM_Collagens1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_Collagens Module Score by Condition") +
#   theme_minimal()

dittoBoxPlot(seu, "ECM_Collagens1", group.by = "age", split.by = 'annotation', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             jitter.size = NA, boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

```{r fig.height=5, fig.width=8}
# Add a module score
seu <- AddModuleScore(seu, features = list(convert_human_to_mouse(esm_genes_df %>% 
                                                                    dplyr::filter(Category == 'Proteoglycans')%>% .$Gene.Symbol)[,2]),
                      name = "ECM_Proteoglycans")

# Plot module score by condition
# VlnPlot(seu, features = "ECM_Proteoglycans1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_Proteoglycans Module Score by Condition") +
#   theme_minimal()

dittoBoxPlot(seu, "ECM_Proteoglycans1", group.by = "age", split.by = 'annotation', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             jitter.size = NA, boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

```{r fig.height=5, fig.width=8}
# Add a module score
seu <- AddModuleScore(seu, features = list(convert_human_to_mouse(esm_genes_df %>% 
                                                                    dplyr::filter(Category == 'ECM-affiliated Proteins')%>% .$Gene.Symbol)[,2]),
                      name = "ECM_affiliated_Proteins")

# Plot module score by condition
# VlnPlot(seu, features = "ECM_affiliated_Proteins1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_affiliated_Proteins Module Score by Condition") +
#   theme_minimal()

dittoBoxPlot(seu, "ECM_affiliated_Proteins1", group.by = "age", split.by = 'annotation', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             jitter.size = NA, boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

```{r fig.height=5, fig.width=8}
# Add a module score
seu <- AddModuleScore(seu, features = list(convert_human_to_mouse(esm_genes_df %>% 
                                                                    dplyr::filter(Category == 'ECM Regulators')%>% .$Gene.Symbol)[,2]),
                      name = "ECM_Regulators")

# Plot module score by condition
# VlnPlot(seu, features = "ECM_Regulators1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_Regulators Module Score by Condition") +
#   theme_minimal()

dittoBoxPlot(seu, "ECM_Regulators1", group.by = "age", split.by = 'annotation', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             jitter.size = NA, boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

```{r fig.height=5, fig.width=8}
# Add a module score
seu <- AddModuleScore(seu, features = list(convert_human_to_mouse(esm_genes_df %>% 
                                                                    dplyr::filter(Category == 'Secreted Factors')%>% .$Gene.Symbol)[,2]),
                      name = "ECM_Secreted_Factors")

# Plot module score by condition
# VlnPlot(seu, features = "ECM_Secreted_Factors1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_Secreted_Factors Module Score by Condition") +
#   theme_minimal()

dittoBoxPlot(seu, "ECM_Secreted_Factors1", group.by = "age", split.by = 'annotation', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             jitter.size = NA, boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

We can see that average module score is higher for adult in approximately all cases
. 
# DEG

Let's look at specific genes that are differently express. 

```{r}
de_results <- list()
esm_genes_df.mouse <- intersect(rownames(seu), esm_genes_df.mouse)
for (cell_type in seu$annotation %>% unique) {
  # Subset cells of the current cell type
  subset_obj <- subset(seu, idents = cell_type)
  
  # Run DE analysis
  Idents(subset_obj) <- "age" # Assuming 'condition' is the metadata column
  de_results[[cell_type]] <- FindMarkers(subset_obj, ident.1 = "young", ident.2 = "adult", features = esm_genes_df.mouse)
}

# Combine DE results into a single dataframe for easy handling
de_results_df <- bind_rows(
  lapply(names(de_results), function(cell_type) {
    de_results[[cell_type]] %>% 
      mutate(CellType = cell_type) %>% 
      rownames_to_column("Gene")
  })
)
```

```{r}
top_genes <- de_results_df %>% 
  dplyr::filter(p_val_adj < 0.05) %>% 
  dplyr::filter(pct.1 > 0.25 | pct.2 > 0.25) %>% 
  dplyr::filter(abs(avg_log2FC) > 0.5) 
```

## T cells

```{r fig.height=3, fig.width=7}
dittoHeatmap(seu, genes = top_genes %>% 
               dplyr::filter(CellType == 'T cells') %>% .$Gene, cells.use = seu$annotation == 'T cells', 
             annot.by = c('age', 'annotation'), 
             annot.colors = c("#6495EDFF", "#FF69B4FF", "#BA55D3FF", "#9ACD32FF", 
                              "#4682B4FF",  "#DDA0DDFF", "#FFA07AFF", "#8FBC8BFF"), 
             cluster_cols = F, breaks = seq(-2, 2, 0.08)) 
```

## Myeloid cells

```{r fig.height=3, fig.width=7}
dittoHeatmap(seu, genes = top_genes %>% 
               dplyr::filter(CellType == 'Myeloid cells') %>% .$Gene, cells.use = seu$annotation == 'Myeloid cells', 
             annot.by = c('age', 'annotation'), 
             annot.colors = c("#6495EDFF", "#FF69B4FF", "#BA55D3FF", "#9ACD32FF", 
                              "#4682B4FF",  "#DDA0DDFF", "#FFA07AFF", "#8FBC8BFF"), 
             cluster_cols = F, breaks = seq(-2, 2, 0.08)) 
```

## Fibroblasts

```{r fig.height=3, fig.width=7}
dittoHeatmap(seu, genes = top_genes %>% 
               dplyr::filter(CellType == 'Fibroblasts') %>% .$Gene, cells.use = seu$annotation == 'Fibroblasts', 
             annot.by = c('age', 'annotation'), 
             annot.colors = c("#6495EDFF", "#FF69B4FF", "#BA55D3FF", "#9ACD32FF", 
                              "#4682B4FF",  "#DDA0DDFF", "#FFA07AFF", "#8FBC8BFF"), 
             cluster_cols = F, breaks = seq(-2, 2, 0.08)) 
```

## Malignant cells

```{r fig.height=6, fig.width=6}
dittoHeatmap(seu, genes = top_genes %>% 
               dplyr::filter(CellType == 'Malignant cells') %>% .$Gene, cells.use = seu$annotation == 'Malignant cells', 
             annot.by = c('age', 'annotation'), 
             annot.colors = c("#6495EDFF", "#FF69B4FF", "#BA55D3FF", "#9ACD32FF", 
                              "#4682B4FF",  "#DDA0DDFF", "#FFA07AFF", "#8FBC8BFF"), 
             cluster_cols = F, breaks = seq(-2, 2, 0.08)) 
```

Based on the heatmap we can see that ECM genes in malignant cells are distributed unevenly. So I decided to look at it closer to probably find some intereting subset of genes.

```{r fig.height=5, fig.width=12,}
malig_cells <- qread(paste0(PREPRDATADIR, 'malig_cells_mouse_gsvas3_ssGSEA.qs'))

Idents(malig_cells) <- malig_cells$RNA_snn_res.0.2
plot_grid(DimPlot(malig_cells, group.by = 'RNA_snn_res.0.2', cols = my_colors) +
  ggtitle("Clusters"), 
  DimPlot(malig_cells, group.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) +
  ggtitle("Age"),
  nrow = 1)
```

Third cluster express genes of ECM more than other clusters. So we found some ECM specific subset of malignant cells. Intersting that in all clusters module score is higher for adult except of this 3 cluster.

```{r}
malig_cells <- AddModuleScore(malig_cells, features = list(esm_genes_df.mouse), name = "ECM_Module")

FeaturePlot(malig_cells, "ECM_Module1")
dittoBoxPlot(malig_cells, "ECM_Module1", group.by = "age", split.by = 'RNA_snn_res.0.2', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             jitter.size = NA, boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

Moreover, we know that adult have much more cells of 3 sluster.

```{r}
plot_grid(dittoBarPlot(malig_cells, 'age', group.by = 'RNA_snn_res.0.2', color.panel = my_colors, scale = 'count') + ggtitle('Age'), dittoBarPlot(malig_cells, 'RNA_snn_res.0.2', group.by = 'age', color.panel = my_colors) + ggtitle('Clusters'),
          rel_widths = c(2, 0.8),
          nrow = 1)
```

### Cluster's markers

```{r fig.height=7, fig.width=12}
Idents(malig_cells) <- 'RNA_snn_res.0.2'

maligcells.markers <- FindAllMarkers(malig_cells, only.pos = TRUE, min.pct = 0.25)
top5 <- maligcells.markers %>%
  dplyr::filter(p_val_adj < 0.05) %>% 
  group_by(cluster) %>%
  top_n(n = 7, wt = avg_log2FC)

DotPlot(malig_cells, c('Mycn', top5$gene), cols = c('lightgrey', "#BA55D3FF")) +
  RotatedAxis()
```

```{r}
# Find markers for cluster 3 vs. all other clusters
cluster3_markers <- FindMarkers(
  object = malig_cells,
  ident.1 = "3",                
  only.pos = TRUE,              
  min.pct = 0.25     
)

# Filter significant markers
cluster3_markers <- cluster3_markers %>%
  dplyr::filter(p_val_adj < 0.05) %>% 
  dplyr::filter(abs(avg_log2FC) > 1)

# View the top markers
cluster3_markers <- cluster3_markers %>% 
  top_n(n = 50, wt = avg_log2FC)
```

```{r fig.height=7, fig.width=7}
dittoHeatmap(malig_cells, genes = rownames(cluster3_markers),
             annot.by = c('RNA_snn_res.0.2'), 
             annot.colors = c("#6495EDFF", "#FF69B4FF", "#BA55D3FF", "#9ACD32FF", 
                              "#4682B4FF",  "#DDA0DDFF", "#FFA07AFF", "#8FBC8BFF"), 
             cluster_cols = F, breaks = seq(-2, 2, 0.08)) 
```

# Human

## single cell

I decided to check if the same picture in human samples too.

```{r}
sc <- qread(paste0(PREPRDATADIR, 'sc_v1.qs'))
sn <- qread(paste0(PREPRDATADIR, 'sn_v1.qs'))
```

```{r eval=FALSE, include=FALSE}
seu_h <- qread(paste0(PREPRDATADIR, 'seu_naivetreatment_v2.qs'))
sn <- subset(seu, Study %in% c('grossman', 'jansky'))

sc <- subset(seu_h, Study %in% c('dong', 'verhoeven', 'wienke'))
sc <- sc %>% 
  NormalizeData() %>% 
  FindVariableFeatures() %>% 
  ScaleData() %>% 
  RunPCA(verbose=FALSE)

sc <- sc %>% 
  RunHarmony(., 'SampleID',
             lambda = 1, verbose = FALSE) %>% 
  RunUMAP(reduction = "harmony", dims = 1:20, verbose=F) 

qsave(sc, paste0(PREPRDATADIR, 'sc_v1.qs'))
```

```{r fig.height=4, fig.width=12}

plot_grid(DimPlot(sc, reduction = "umap", group.by = 'ANNOTATION_FINAL', cols = my_colors),
DimPlot(sc, reduction = "umap", group.by = 'Study', cols = my_colors), nrow = 1)
```

```{r fig.height=4, fig.width=12}

plot_grid(DimPlot(sn, reduction = "umap", group.by = 'ANNOTATION_FINAL', cols = my_colors),
DimPlot(sn, reduction = "umap", group.by = 'Study', cols = my_colors), nrow = 1)
```

```{r fig.height=5, fig.width=8}
# Add a module score
sc <- AddModuleScore(sc, features = list(esm_genes_df$Gene.Symbol), name = "ECM_Module")

# Plot module score by condition
# VlnPlot(seu, features = "ECM_Module1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM all genes Score by Condition") +
#   theme_minimal()

dittoBoxPlot(sc, "ECM_Module1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA, 
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5) +
  ggtitle('ECM_Module - all genes')
```

```{r fig.height=5, fig.width=8}
# Add a module score
sc <- AddModuleScore(sc, features = list(esm_genes_df %>% dplyr::filter(Category == 'ECM Glycoproteins') %>% .$Gene.Symbol), 
                      name = "ECM_Glycoproteins")

# Plot module score by condition
# VlnPlot(seu, features = "ECM_Glycoproteins1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_Glycoproteins Module Score by Condition") +
#   theme_minimal()

dittoBoxPlot(sc, "ECM_Glycoproteins1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA,
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

```{r fig.height=5, fig.width=8}
# Add a module score
sc <- AddModuleScore(sc, features = list(esm_genes_df %>% dplyr::filter(Category == 'Collagens')%>% .$Gene.Symbol),
                      name = "ECM_Collagens")

# Plot module score by condition
# VlnPlot(seu, features = "ECM_Collagens1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_Collagens Module Score by Condition") +
#   theme_minimal()

dittoBoxPlot(sc, "ECM_Collagens1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA, 
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

```{r fig.height=5, fig.width=8}
# Add a module score
sc <- AddModuleScore(sc, features = list(esm_genes_df %>% dplyr::filter(Category == 'Proteoglycans')
                                         %>% .$Gene.Symbol),
                      name = "ECM_Proteoglycans")

# Plot module score by condition
# VlnPlot(seu, features = "ECM_Proteoglycans1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_Proteoglycans Module Score by Condition") +
#   theme_minimal()

dittoBoxPlot(sc, "ECM_Proteoglycans1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA, 
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

```{r fig.height=5, fig.width=8}
# Add a module score
sc <- AddModuleScore(sc, features = list(esm_genes_df %>% 
                                           dplyr::filter(Category == 'ECM-affiliated Proteins') %>%
                                           .$Gene.Symbol),
                      name = "ECM_affiliated_Proteins")

# Plot module score by condition
# VlnPlot(seu, features = "ECM_affiliated_Proteins1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_affiliated_Proteins Module Score by Condition") +
#   theme_minimal()

dittoBoxPlot(sc, "ECM_affiliated_Proteins1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA, 
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

```{r fig.height=5, fig.width=8}
# Add a module score
sc <- AddModuleScore(sc, features = list(esm_genes_df %>% 
                                           dplyr::filter(Category == 'ECM Regulators') %>%
                                           .$Gene.Symbol),
                      name = "ECM_Regulators")

# Plot module score by condition
# VlnPlot(seu, features = "ECM_Regulators1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_Regulators Module Score by Condition") +
#   theme_minimal()

dittoBoxPlot(sc, "ECM_Regulators1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA, 
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

```{r fig.height=5, fig.width=8}
# Add a module score
sc <- AddModuleScore(sc, features = list(esm_genes_df %>% 
                                           dplyr::filter(Category == 'Secreted Factors') %>%
                                           .$Gene.Symbol),
                      name = "ECM_Secreted_Factors")

# Plot module score by condition
# VlnPlot(seu, features = "ECM_Secreted_Factors1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_Secreted_Factors Module Score by Condition") +
#   theme_minimal()

dittoBoxPlot(sc, "ECM_Secreted_Factors1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA, 
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```


## single-nuclei

```{r eval=FALSE, fig.height=4, fig.width=12, include=FALSE}
sn <- sn %>% 
  NormalizeData() %>% 
  FindVariableFeatures() %>% 
  ScaleData() %>% 
  RunPCA(verbose=FALSE)

sn <- sn %>% 
  RunHarmony(., 'SampleID',
             lambda = 1, verbose = FALSE) %>% 
  RunUMAP(reduction = "harmony", dims = 1:20, verbose=F) 


qsave(sn, paste0(PREPRDATADIR, 'sn_v1.qs'))
```

```{r echo=FALSE, fig.height=4, fig.width=12}
plot_grid(DimPlot(sn, reduction = "umap", group.by = 'ANNOTATION_FINAL', cols = my_colors),
DimPlot(sn, reduction = "umap", group.by = 'Study', cols = my_colors), nrow = 1)
```

```{r fig.height=5, fig.width=8}
# Add a module snore
sn <- AddModuleScore(sn, features = list(esm_genes_df$Gene.Symbol), name = "ECM_Module")

# Plot module snore by condition
# VlnPlot(seu, features = "ECM_Module1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM all genes snore by Condition") +
#   theme_minimal()

dittoBoxPlot(sn, "ECM_Module1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA, 
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5) +
  ggtitle('ECM_Module - all genes')
```

```{r fig.height=5, fig.width=8}
# Add a module snore
sn <- AddModuleScore(sn, features = list(esm_genes_df %>% dplyr::filter(Category == 'ECM Glycoproteins') %>% .$Gene.Symbol), 
                     name = "ECM_Glycoproteins")

# Plot module snore by condition
# VlnPlot(seu, features = "ECM_Glycoproteins1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_Glycoproteins Module snore by Condition") +
#   theme_minimal()

dittoBoxPlot(sn, "ECM_Glycoproteins1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA,
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

```{r fig.height=5, fig.width=8}
# Add a module snore
sn <- AddModuleScore(sn, features = list(esm_genes_df %>% dplyr::filter(Category == 'Collagens')%>% .$Gene.Symbol),
                     name = "ECM_Collagens")

# Plot module snore by condition
# VlnPlot(seu, features = "ECM_Collagens1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_Collagens Module snore by Condition") +
#   theme_minimal()

dittoBoxPlot(sn, "ECM_Collagens1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA, 
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

```{r fig.height=5, fig.width=8}
# Add a module snore
sn <- AddModuleScore(sn, features = list(esm_genes_df %>% dplyr::filter(Category == 'Proteoglycans')
                                         %>% .$Gene.Symbol),
                     name = "ECM_Proteoglycans")

# Plot module snore by condition
# VlnPlot(seu, features = "ECM_Proteoglycans1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_Proteoglycans Module snore by Condition") +
#   theme_minimal()

dittoBoxPlot(sn, "ECM_Proteoglycans1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA, 
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

```{r fig.height=5, fig.width=8}
# Add a module snore
sn <- AddModuleScore(sn, features = list(esm_genes_df %>% 
                                           dplyr::filter(Category == 'ECM-affiliated Proteins') %>%
                                           .$Gene.Symbol),
                     name = "ECM_affiliated_Proteins")

# Plot module snore by condition
# VlnPlot(seu, features = "ECM_affiliated_Proteins1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_affiliated_Proteins Module snore by Condition") +
#   theme_minimal()

dittoBoxPlot(sn, "ECM_affiliated_Proteins1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA, 
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

```{r fig.height=5, fig.width=8}
# Add a module snore
sn <- AddModuleScore(sn, features = list(esm_genes_df %>% 
                                           dplyr::filter(Category == 'ECM Regulators') %>%
                                           .$Gene.Symbol),
                     name = "ECM_Regulators")

# Plot module snore by condition
# VlnPlot(seu, features = "ECM_Regulators1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_Regulators Module snore by Condition") +
#   theme_minimal()

dittoBoxPlot(sn, "ECM_Regulators1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA, 
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```

```{r fig.height=5, fig.width=8}
# Add a module snore
sn <- AddModuleScore(sn, features = list(esm_genes_df %>% 
                                           dplyr::filter(Category == 'Secreted Factors') %>%
                                           .$Gene.Symbol),
                     name = "ECM_Secreted_Factors")

# Plot module snore by condition
# VlnPlot(seu, features = "ECM_Secreted_Factors1", group.by = "annotation", split.by = 'age', cols = c("#6495EDFF", "#FF69B4FF")) + 
#   ggtitle("ECM_Secreted_Factors Module snore by Condition") +
#   theme_minimal()

dittoBoxPlot(sn, "ECM_Secreted_Factors1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA, 
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5)
```


# Malignant cells check

```{r}
malig_cells <- subset(sc, ANNOTATION_FINAL == 'malignant_cells')

malig_cells <- malig_cells %>% 
  NormalizeData() %>% 
  FindVariableFeatures() %>% 
  ScaleData(vars.to.regress = c("S.Score", "G2M.Score")) %>% 
  RunPCA(verbose=FALSE) %>% 
  RunHarmony(., 'Study',
             lambda = 1, verbose = FALSE) %>% 
  RunUMAP(reduction = "harmony", dims = 1:20, verbose=F) %>% 
  FindNeighbors(dims = 1:20) %>%
  FindClusters(resolution = seq(0.1, 1, 0.1))
```

```{r}
clustree(malig_cells)
```

```{r fig.height=5, fig.width=16,}
Idents(malig_cells) <- malig_cells$RNA_snn_res.0.1
plot_grid(DimPlot(malig_cells, group.by = 'RNA_snn_res.0.1') +
  ggtitle("Clusters"), 
  DimPlot(malig_cells, group.by = 'less18M', cols = c("#6495EDFF", "#FF69B4FF")) +
  ggtitle("Age"),
  DimPlot(malig_cells, group.by = 'SampleID') + NoLegend() +
  ggtitle("Sample"),
  nrow = 1)
```

```{r}
FeaturePlot(malig_cells, 'MYCN')
```

# Angiogenesis check

mouse genes

```{r}
hallmark_angio_m <- msigdbr(species = "Mus musculus", 
                           category = "H") %>%
    dplyr::filter(gs_name == "HALLMARK_ANGIOGENESIS") %>%
    dplyr::pull(gene_symbol)

hallmark_angio_m
```

human genes

```{r}
hallmark_angio <- msigdbr(species = "Homo sapiens", 
                           category = "H") %>%
    dplyr::filter(gs_name == "HALLMARK_ANGIOGENESIS") %>%
    dplyr::pull(gene_symbol)

hallmark_angio
```

## Mouse

```{r}
  
# Calculate module score
seu <- AddModuleScore(
  object = seu,
  features = list(hallmark_angio_m),
  name = "angiogenesis_score"
)

dittoBoxPlot(seu, "angiogenesis_score1", group.by = "age", split.by = 'annotation', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA, 
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5) +
  ggtitle('Angiogenesis score - mouse')
```

### DEG

Let's look at specific genes that are differently express. 

```{r}
de_results <- list()
hallmark_angio_m <- intersect(rownames(seu), hallmark_angio_m)
for (cell_type in seu$annotation %>% unique) {
  # Subset cells of the current cell type
  subset_obj <- subset(seu, idents = cell_type)
  
  # Run DE analysis
  Idents(subset_obj) <- "age" # Assuming 'condition' is the metadata column
  de_results[[cell_type]] <- FindMarkers(subset_obj, ident.1 = "young", ident.2 = "adult", features = hallmark_angio_m)
}

# Combine DE results into a single dataframe for easy handling
de_results_df <- bind_rows(
  lapply(names(de_results), function(cell_type) {
    de_results[[cell_type]] %>% 
      mutate(CellType = cell_type) %>% 
      rownames_to_column("Gene")
  })
)
```

```{r}
top_genes <- de_results_df %>% 
  dplyr::filter(p_val_adj < 0.05) %>% 
  dplyr::filter(pct.1 > 0.25 | pct.2 > 0.25) %>% 
  dplyr::filter(abs(avg_log2FC) > 0.5) 
```

#### T cells

```{r fig.height=3, fig.width=7}
dittoHeatmap(seu, genes = top_genes %>% 
               dplyr::filter(CellType == 'T cells') %>% .$Gene, cells.use = seu$annotation == 'T cells', 
             annot.by = c('age', 'annotation'), 
             annot.colors = c("#6495EDFF", "#FF69B4FF", "#BA55D3FF", "#9ACD32FF", 
                              "#4682B4FF",  "#DDA0DDFF", "#FFA07AFF", "#8FBC8BFF"), 
             cluster_cols = F, breaks = seq(-2, 2, 0.08)) 
```

#### Myeloid cells

there are not DEG

#### Fibroblasts

there are not DEG

#### Malignant cells

```{r fig.height=3, fig.width=6}
dittoHeatmap(seu, genes = top_genes %>% 
               dplyr::filter(CellType == 'Malignant cells') %>% .$Gene, cells.use = seu$annotation == 'Malignant cells', 
             annot.by = c('age', 'annotation'), 
             annot.colors = c("#6495EDFF", "#FF69B4FF", "#BA55D3FF", "#9ACD32FF", 
                              "#4682B4FF",  "#DDA0DDFF", "#FFA07AFF", "#8FBC8BFF"), 
             cluster_cols = F, breaks = seq(-2, 2, 0.08)) 
```

## Human

### sc

```{r}

# Calculate module score
sc <- AddModuleScore(
  object = sc,
  features = list(hallmark_angio),
  name = "angiogenesis_score"
)

sn <- AddModuleScore(
  object = sn,
  features = list(hallmark_angio),
  name = "angiogenesis_score"
)

dittoBoxPlot(sc, "angiogenesis_score1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA, 
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5) +
  ggtitle('Angiogenesis score - single cell')
```


### sn

```{r}
sn$less18M <- as.character(sn$less18M)
dittoBoxPlot(sn, "angiogenesis_score1", group.by = "less18M", split.by = 'ANNOTATION_FINAL', color.panel = c("#6495EDFF", "#FF69B4FF"), 
             # jitter.size = NA, 
             boxplot.width = 0.4, split.nrow = 1, boxplot.lineweight = 0.5) +
  ggtitle('Angiogenesis score - single nuclei')
```




