---
title: "nichenet_ATLAS"
author: "Kate Petrenko"
date: "2025-04-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# devtools::install_github("saeyslab/nichenetr")
library(nichenetr)
library(RColorBrewer)
library(tidyverse)
library(Seurat) 
library(qs)
library(EnhancedVolcano)
```

```{r global variables, include=FALSE}
set.seed(100)
DATADIR = '~/neuroblastoma/data/'
WORKDIR = '~/neuroblastoma/'
PREPRDATADIR = '~/neuroblastoma/preprocessing_data/'
OUTNAME = 'FILTER_SCORE_'
my_colors <- c("#6495EDFF", "#FF69B4FF", "#BA55D3FF", "#A4BEF3", "#F08080FF", "#32CD32FF",  
    "#9ACD32FF", "#4682B4FF", "#DDA0DDFF", "#FFA07AFF", "#8FBC8BFF",
    "#40E0D0FF", "#F0E68CFF", "#5F9EA0FF", "#D2B48CFF",  
    "#FFDAB9FF", "#87CEEBFF", "#B4A0E5", "#5BC0BE", "#773344")
```

# Data

```{r include=FALSE}
atlas <- qread(paste0(PREPRDATADIR, 'ATLAS_object.qs'))
```

```{r eval=FALSE, include=FALSE}
sample_metadata <- read.csv(paste0(DATADIR, 'samples_metadata_v5.csv'))
sample_metadata[sample_metadata==""] <- NA

rownames(sample_metadata) <- sample_metadata$Sample_dataset



clean_sample_id <- function(id) {
  # First remove _rep1 and _rep2
  id <- gsub("_rep[12]$", "", id)
  
  # Handle the special bonine cases
  id <- gsub("_sn$", "", id)      # Remove _sn suffix
  id <- gsub("_[12]$", "", id)    # Remove _1 or _2 suffix
  
  return(id)
}

atlas$less18M <- sample_metadata[clean_sample_id(atlas$SampleID),]$less18M
atlas$Sex <- sample_metadata[clean_sample_id(atlas$SampleID),]$Sex
atlas$MYCN_status <- sample_metadata[clean_sample_id(atlas$SampleID),]$MYCN
atlas$method <- sample_metadata[clean_sample_id(atlas$SampleID),]$Method
```

```{r echo=FALSE}
metadata_python <- read.csv("query_metadata_with_umap.csv", row.names = 1)

# Match to Seurat cells
metadata_python <- metadata_python[colnames(atlas), ]
atlas <- AddMetaData(atlas, metadata = metadata_python[c("Age", "celltypist_cell_label_fine", "Final_Annotation")])

# Add UMAP
atlas[['UMAP']] <- CreateDimReducObject(embeddings = as.matrix(metadata_python[, c("UMAP_1", "UMAP_2")]), key = "UMAP_", global = T, assay = "RNA")

DimPlot(atlas, group.by = 'Final_Annotation', cols = my_colors, reduction = 'UMAP')
```

```{r include=FALSE}
lr_network <- readRDS(url("https://zenodo.org/record/7074291/files/lr_network_human_21122021.rds"))
# ligand_target_matrix <- readRDS(url("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final.rds"))
# options(timeout = 600)
# download.file("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final.rds", 
              # destfile = "ligand_target_matrix.rds", mode = "wb")
ligand_target_matrix <- readRDS("ligand_target_matrix.rds")
# weighted_networks <- readRDS(url("https://zenodo.org/record/7074291/files/weighted_networks_nsga2r_final.rds"))
```

```{r eval=FALSE, include=FALSE}
# ligand_target_matrix = readRDS(url("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final_mouse.rds"))
ligand_target_matrix[1:5,1:5] # target genes in rows, ligands in columns
```

```{r include=FALSE}
# lr_network = readRDS(url("https://zenodo.org/record/7074291/files/lr_network_mouse_21122021.rds"))
lr_network = lr_network %>% dplyr::rename(ligand = from, receptor = to) %>% distinct(ligand, receptor)
```

```{r include=FALSE}
atlas$niche_label <- ifelse(atlas$Age == "< 18 months", "young_niche", "old_niche")
Idents(atlas) <- "niche_label"  # cell types

atlas$niche_label <- ifelse(atlas$Age == "< 18 months", "young", "adult")
atlas$niche_label <- paste0(atlas$Final_Annotation, '_', atlas$niche_label)
Idents(atlas) <- "niche_label"  # cell types
```

```{r include=FALSE}
niches = list(
  "young_niche" = list(
    "sender" = c("Neutrophils_young"
                 # , "Monocytes_young"
                 # , "Macrophages_young"
                 ),
    "receiver" = c("Neuroendocrine_young")
  ),
  "old_niche" = list(
    "sender" = c("Neutrophils_adult"
                 # , "Monocytes_adult"
                 # , "Macrophages_adult"
                 ),
    "receiver" = c("Neuroendocrine_adult")
  )
)
```

```{r include=FALSE}
assay_oi <- 'RNA'
DE_sender = calculate_niche_de(seurat_obj = atlas %>% subset(features = lr_network$ligand %>% intersect(rownames(atlas))), niches = niches, type = "sender", assay_oi = assay_oi) # only ligands important for sender cell types

DE_sender %>% 
  dplyr::filter(p_val_adj < 0.05 & sender == 'Neutrophils_young')
DE_sender %>% 
  dplyr::filter(p_val_adj < 0.05 & sender == 'Monocytes_young')
DE_sender %>% 
  dplyr::filter(p_val_adj < 0.05 & sender == 'Macrophages_young')
```

```{r include=FALSE}
DE_receiver = calculate_niche_de(seurat_obj = atlas %>% subset(features = lr_network$receptor %>% unique()), niches = niches, type = "receiver", assay_oi = assay_oi) # only receptors now, later on: DE analysis to find targets

DE_receiver %>% 
  dplyr::filter(p_val_adj < 0.05 & receiver == 'Neuroendocrine_young') %>% 
  dplyr::filter(pct.1 > 0.25 | pct.2 > 0.25)
```

```{r include=FALSE}
DE_sender = DE_sender %>% mutate(avg_log2FC = ifelse(avg_log2FC == Inf, max(avg_log2FC[is.finite(avg_log2FC)]), ifelse(avg_log2FC == -Inf, min(avg_log2FC[is.finite(avg_log2FC)]), avg_log2FC)))
DE_receiver = DE_receiver %>% mutate(avg_log2FC = ifelse(avg_log2FC == Inf, max(avg_log2FC[is.finite(avg_log2FC)]), ifelse(avg_log2FC == -Inf, min(avg_log2FC[is.finite(avg_log2FC)]), avg_log2FC)))
```

```{r include=FALSE}
expression_pct = 0.15
DE_sender_processed = process_niche_de(DE_table = DE_sender, niches = niches, expression_pct = expression_pct, type = "sender")
DE_receiver_processed = process_niche_de(DE_table = DE_receiver, niches = niches, expression_pct = expression_pct, type = "receiver")
```

```{r include=FALSE}
specificity_score_LR_pairs = "min_lfc"
DE_sender_receiver = combine_sender_receiver_de(DE_sender_processed, DE_receiver_processed, lr_network, specificity_score = specificity_score_LR_pairs)
```

```{r include=FALSE}
lfc_cutoff = 0.4 
target_score = 0.4
specificity_score_targets = "min_lfc"

DE_receiver_targets = calculate_niche_de_targets(seurat_obj = atlas, niches = niches, lfc_cutoff = lfc_cutoff, expression_pct = expression_pct, assay_oi = assay_oi) 

DE_receiver_processed_targets = process_receiver_target_de(DE_receiver_targets = DE_receiver_targets, niches = niches, expression_pct = expression_pct, specificity_score = specificity_score_targets)
  
background = DE_receiver_processed_targets  %>% pull(target) %>% unique()
geneset_NEyoung = DE_receiver_processed_targets %>% filter(receiver == niches$young_niche$receiver & target_score >= target_score & target_significant == 1 & target_present == 1) %>% pull(target) %>% unique()
geneset_NEadult = DE_receiver_processed_targets %>% filter(receiver == niches$old_niche$receiver & target_score >= target_score & target_significant == 1 & target_present == 1) %>% pull(target) %>% unique()

# Good idea to check which genes will be left out of the ligand activity analysis (=when not present in the rownames of the ligand-target matrix).
# If many genes are left out, this might point to some issue in the gene naming (eg gene aliases and old gene symbols, bad human-mouse mapping)
geneset_NEyoung %>% setdiff(rownames(ligand_target_matrix))
geneset_NEadult %>% setdiff(rownames(ligand_target_matrix))


length(geneset_NEyoung)
length(geneset_NEadult)
```

```{r}
DE_receiver_processed_targets %>% 
  dplyr::filter(target_score > 0.5)
```

```{r include=FALSE}
top_n_target = 250

niche_geneset_list = list(
    "young_niche" = list(
      "receiver" = niches$young_niche$receiver,
      "geneset" = geneset_NEyoung,
      "background" = background),
    "old_niche" = list(
      "receiver" = niches$old_niche$receiver,
      "geneset" = geneset_NEadult ,
      "background" = background)
  )
  
ligand_activities_targets = get_ligand_activities_targets(niche_geneset_list = niche_geneset_list, ligand_target_matrix = ligand_target_matrix, top_n_target = top_n_target)  
```

```{r include=FALSE}
features_oi = union(lr_network$ligand, lr_network$receptor) %>% union(ligand_activities_targets$target) %>% setdiff(NA)
  
dotplot = suppressWarnings(Seurat::DotPlot(atlas %>% subset(idents = niches %>% unlist() %>% unique()), features = features_oi, assay = assay_oi))
exprs_tbl = dotplot$data %>% as_tibble()
exprs_tbl = exprs_tbl %>% rename(celltype = id, gene = features.plot, expression = avg.exp, expression_scaled = avg.exp.scaled, fraction = pct.exp) %>%
    mutate(fraction = fraction/100) %>% as_tibble() %>% select(celltype, gene, expression, expression_scaled, fraction) %>% distinct() %>% arrange(gene) %>% mutate(gene = as.character(gene))
  
exprs_tbl_ligand = exprs_tbl %>% filter(gene %in% lr_network$ligand) %>% rename(sender = celltype, ligand = gene, ligand_expression = expression, ligand_expression_scaled = expression_scaled, ligand_fraction = fraction) 
exprs_tbl_receptor = exprs_tbl %>% filter(gene %in% lr_network$receptor) %>% rename(receiver = celltype, receptor = gene, receptor_expression = expression, receptor_expression_scaled = expression_scaled, receptor_fraction = fraction)
exprs_tbl_target = exprs_tbl %>% filter(gene %in% ligand_activities_targets$target) %>% rename(receiver = celltype, target = gene, target_expression = expression, target_expression_scaled = expression_scaled, target_fraction = fraction)
```

```{r include=FALSE}
exprs_tbl_ligand = exprs_tbl_ligand %>%  mutate(scaled_ligand_expression_scaled = scale_quantile_adapted(ligand_expression_scaled)) %>% mutate(ligand_fraction_adapted = ligand_fraction) %>% mutate_cond(ligand_fraction >= expression_pct, ligand_fraction_adapted = expression_pct)  %>% mutate(scaled_ligand_fraction_adapted = scale_quantile_adapted(ligand_fraction_adapted))

exprs_tbl_receptor = exprs_tbl_receptor %>% mutate(scaled_receptor_expression_scaled = scale_quantile_adapted(receptor_expression_scaled))  %>% mutate(receptor_fraction_adapted = receptor_fraction) %>% mutate_cond(receptor_fraction >= expression_pct, receptor_fraction_adapted = expression_pct)  %>% mutate(scaled_receptor_fraction_adapted = scale_quantile_adapted(receptor_fraction_adapted))
```


```{r include=FALSE}
include_spatial_info_sender = FALSE # if not spatial info to include: put this to false 
include_spatial_info_receiver = FALSE # if spatial info to include: put this to true 
```


```{r include=FALSE}
# this is how this should be defined if you don't have spatial info
# mock spatial info
if(include_spatial_info_sender == FALSE & include_spatial_info_receiver == FALSE){
    spatial_info = tibble(celltype_region_oi = NA, celltype_other_region = NA) %>% mutate(niche =  niches %>% names() %>% head(1), celltype_type = "sender")
} 
```

```{r include=FALSE}
if(include_spatial_info_sender == TRUE){
  sender_spatial_DE = calculate_spatial_DE(seurat_obj = seurat_obj %>% subset(features = lr_network$ligand %>% unique()), spatial_info = spatial_info %>% filter(celltype_type == "sender"), assay_oi = assay_oi)
  sender_spatial_DE_processed = process_spatial_de(DE_table = sender_spatial_DE, type = "sender", lr_network = lr_network, expression_pct = expression_pct, specificity_score = specificity_score_spatial)

  # add a neutral spatial score for sender celltypes in which the spatial is not known / not of importance
  sender_spatial_DE_others = get_non_spatial_de(niches = niches, spatial_info = spatial_info, type = "sender", lr_network = lr_network)
  sender_spatial_DE_processed = sender_spatial_DE_processed %>% bind_rows(sender_spatial_DE_others)

  sender_spatial_DE_processed = sender_spatial_DE_processed %>% mutate(scaled_ligand_score_spatial = scale_quantile_adapted(ligand_score_spatial))

} else {
  # # add a neutral spatial score for all sender celltypes (for none of them, spatial is relevant in this case)
  sender_spatial_DE_processed = get_non_spatial_de(niches = niches, spatial_info = spatial_info, type = "sender", lr_network = lr_network)
  sender_spatial_DE_processed = sender_spatial_DE_processed %>% mutate(scaled_ligand_score_spatial = scale_quantile_adapted(ligand_score_spatial))  

}
## [1] "Calculate Spatial DE between: LSECs_portal and LSECs_central"
## [1] "Calculate Spatial DE between: Hepatocytes_portal and Hepatocytes_central"
## [1] "Calculate Spatial DE between: Stellate cells_portal and Stellate cells_central"
```

```{r include=FALSE}
if(include_spatial_info_receiver == TRUE){
  receiver_spatial_DE = calculate_spatial_DE(seurat_obj = seurat_obj %>% subset(features = lr_network$receptor %>% unique()), spatial_info = spatial_info %>% filter(celltype_type == "receiver"), assay_oi = assay_oi)
  receiver_spatial_DE_processed = process_spatial_de(DE_table = receiver_spatial_DE, type = "receiver", lr_network = lr_network, expression_pct = expression_pct, specificity_score = specificity_score_spatial)

  # add a neutral spatial score for receiver celltypes in which the spatial is not known / not of importance
  receiver_spatial_DE_others = get_non_spatial_de(niches = niches, spatial_info = spatial_info, type = "receiver", lr_network = lr_network)
  receiver_spatial_DE_processed = receiver_spatial_DE_processed %>% bind_rows(receiver_spatial_DE_others)

  receiver_spatial_DE_processed = receiver_spatial_DE_processed %>% mutate(scaled_receptor_score_spatial = scale_quantile_adapted(receptor_score_spatial))

} else {
    # # add a neutral spatial score for all receiver celltypes (for none of them, spatial is relevant in this case)
  receiver_spatial_DE_processed = get_non_spatial_de(niches = niches, spatial_info = spatial_info, type = "receiver", lr_network = lr_network)
  receiver_spatial_DE_processed = receiver_spatial_DE_processed %>% mutate(scaled_receptor_score_spatial = scale_quantile_adapted(receptor_score_spatial))
}
```

# Analysis

```{r include=FALSE}
exprs_sender_receiver = lr_network %>% 
  inner_join(exprs_tbl_ligand, by = c("ligand")) %>% 
  inner_join(exprs_tbl_receptor, by = c("receptor")) %>% inner_join(DE_sender_receiver %>% distinct(niche, sender, receiver))
  
ligand_scaled_receptor_expression_fraction_df = exprs_sender_receiver %>% group_by(ligand, receiver) %>% mutate(rank_receptor_expression = dense_rank(receptor_expression), rank_receptor_fraction  = dense_rank(receptor_fraction)) %>% mutate(ligand_scaled_receptor_expression_fraction = 0.5*( (rank_receptor_fraction / max(rank_receptor_fraction)) + ((rank_receptor_expression / max(rank_receptor_expression))) ) )  %>% distinct(ligand, receptor, receiver, ligand_scaled_receptor_expression_fraction) %>% distinct() %>% ungroup() 
```

```{r include=FALSE}
prioritizing_weights = c("scaled_ligand_score" = 5,
                         "scaled_ligand_expression_scaled" = 1,
                         "ligand_fraction" = 1,
                         "scaled_ligand_score_spatial" = 0, 
                         "scaled_receptor_score" = 0.5,
                         "scaled_receptor_expression_scaled" = 0.5,
                          "receptor_fraction" = 1, 
                         "ligand_scaled_receptor_expression_fraction" = 1,
                         "scaled_receptor_score_spatial" = 0,
                         "scaled_activity" = 0,
                         "scaled_activity_normalized" = 1)
```

```{r include=FALSE}
output = list(DE_sender_receiver = DE_sender_receiver, ligand_scaled_receptor_expression_fraction_df = ligand_scaled_receptor_expression_fraction_df, sender_spatial_DE_processed = sender_spatial_DE_processed, receiver_spatial_DE_processed = receiver_spatial_DE_processed,
         ligand_activities_targets = ligand_activities_targets, DE_receiver_processed_targets = DE_receiver_processed_targets, exprs_tbl_ligand = exprs_tbl_ligand,  exprs_tbl_receptor = exprs_tbl_receptor, exprs_tbl_target = exprs_tbl_target)
prioritization_tables = get_prioritization_tables(output, prioritizing_weights)
```

```{r include=FALSE}


prioritization_tables$prioritization_tbl_ligand_receptor %>% filter(receiver == niches[[1]]$receiver) %>% head(10)

prioritization_tables$prioritization_tbl_ligand_target %>% filter(receiver == niches[[1]]$receiver) %>% head(10)


prioritization_tables$prioritization_tbl_ligand_receptor %>% filter(receiver == niches[[2]]$receiver) %>% head(10)

prioritization_tables$prioritization_tbl_ligand_target %>% filter(receiver == niches[[2]]$receiver) %>% head(10)


prioritization_tables$prioritization_tbl_ligand_receptor = prioritization_tables$prioritization_tbl_ligand_receptor %>% mutate(receiver = factor(receiver, levels = c("Neuroendocrine_young","Neuroendocrine_adult")), niche = factor(niche, levels = c("young_niche","old_niche"))) 
prioritization_tables$prioritization_tbl_ligand_target = prioritization_tables$prioritization_tbl_ligand_target %>% mutate(receiver = factor(receiver, levels = c("Neuroendocrine_young","Neuroendocrine_adult")), niche = factor(niche, levels = c("young_niche","old_niche"))) 
```

# Visualization

## Young

```{r include=FALSE}
top_ligand_niche_df = prioritization_tables$prioritization_tbl_ligand_receptor %>% select(niche, sender, receiver, ligand, receptor, prioritization_score) %>% group_by(ligand) %>% top_n(1, prioritization_score) %>% ungroup() %>% select(ligand, receptor, niche) %>% rename(top_niche = niche)
top_ligand_receptor_niche_df = prioritization_tables$prioritization_tbl_ligand_receptor %>% select(niche, sender, receiver, ligand, receptor, prioritization_score) %>% group_by(ligand, receptor) %>% top_n(1, prioritization_score) %>% ungroup() %>% select(ligand, receptor, niche) %>% rename(top_niche = niche)

ligand_prioritized_tbl_oi = prioritization_tables$prioritization_tbl_ligand_receptor %>% select(niche, sender, receiver, ligand, prioritization_score) %>% group_by(ligand, niche) %>% top_n(1, prioritization_score) %>% ungroup() %>% distinct() %>% inner_join(top_ligand_niche_df) %>% filter(niche == top_niche) %>% group_by(niche) %>% top_n(50, prioritization_score) %>% ungroup() # get the top50 ligands per niche
```

```{r include=FALSE}
receiver_oi = "Neuroendocrine_young" 

filtered_ligands = ligand_prioritized_tbl_oi %>% filter(receiver == receiver_oi) %>% pull(ligand) %>% unique()

prioritized_tbl_oi = prioritization_tables$prioritization_tbl_ligand_receptor %>% filter(ligand %in% filtered_ligands) %>% select(niche, sender, receiver, ligand,  receptor, ligand_receptor, prioritization_score) %>% distinct() %>% inner_join(top_ligand_receptor_niche_df) %>% group_by(ligand) %>% filter(receiver == receiver_oi) %>% top_n(2, prioritization_score) %>% ungroup() 
```

```{r echo=FALSE, fig.height=12, fig.width=8, message=FALSE, warning=FALSE}
lfc_plot = make_ligand_receptor_lfc_plot(receiver_oi, prioritized_tbl_oi, prioritization_tables$prioritization_tbl_ligand_receptor, plot_legend = FALSE, heights = NULL, widths = NULL)
lfc_plot 
```

low expression of IFITM1 was associated with a poor prognosis in PRAD (Prostate adenocarcinoma). 

In a recent study published in Cell, Gungabeesoon and colleagues1 demonstrated that a specific subpopulation of neutrophils, characterized by a distinct Sell(hi) (CD62L(hi)) phenotype with an interferon gene signature, acutely accumulates in tumors during successful immunotherapy and is associated with better treatment outcome.

we conclude that OSM produced by recruited neutrophils tunes early innate immune signaling pathways, improving pneumonia outcomes.




```{r fig.height=10, fig.width=34, include=FALSE}
exprs_activity_target_plot = make_ligand_activity_target_exprs_plot(receiver_oi, prioritized_tbl_oi,  prioritization_tables$prioritization_tbl_ligand_receptor,  prioritization_tables$prioritization_tbl_ligand_target, output$exprs_tbl_ligand,  output$exprs_tbl_target, lfc_cutoff, ligand_target_matrix, plot_legend = FALSE, heights = NULL, widths = NULL)
exprs_activity_target_plot$combined_plot
```

```{r echo=FALSE, fig.height=10, fig.width=34, message=FALSE, warning=FALSE}
exprs_activity_target_plot = make_ligand_activity_target_exprs_plot(receiver_oi, prioritized_tbl_oi,  prioritization_tables$prioritization_tbl_ligand_receptor, 
                                                                    prioritization_tables$prioritization_tbl_ligand_target, output$exprs_tbl_ligand,  output$exprs_tbl_target, lfc_cutoff, ligand_target_matrix,
                                                                    scaled_ligand_activity_limits = "IQR", plot_legend = FALSE, heights = NULL, widths = NULL)
exprs_activity_target_plot$combined_plot
```

```{r message=FALSE, warning=FALSE, include=FALSE}

filtered_ligands = ligand_prioritized_tbl_oi %>% filter(receiver == receiver_oi) %>% top_n(15, prioritization_score) %>% pull(ligand) %>% unique() %>% c('S100A8', "S100A9")

prioritized_tbl_oi = prioritization_tables$prioritization_tbl_ligand_receptor %>% filter(ligand %in% filtered_ligands) %>% select(niche, sender, receiver, ligand,  receptor, ligand_receptor, prioritization_score) %>% distinct() %>% inner_join(top_ligand_receptor_niche_df) %>% group_by(ligand) %>% filter(receiver == receiver_oi) %>% top_n(2, prioritization_score) %>% ungroup() 

colors_sender = brewer.pal(n = prioritized_tbl_oi$sender %>% unique() %>% sort() %>% length(), name = 'Spectral') %>% magrittr::set_names(prioritized_tbl_oi$sender %>% unique() %>% sort())
colors_receiver = c("lavender")  %>% magrittr::set_names(prioritized_tbl_oi$receiver %>% unique() %>% sort())
colors_sender[1] <- '#BA55D3FF'
colors_receiver[1] <- '#A4BEF3'
circos_output = make_circos_lr(prioritized_tbl_oi, colors_sender, colors_receiver, separate_legend = F)
```

```{r fig.height=5, fig.width=10}
circos_output
```

## Adult

```{r include=FALSE}
receiver_oi = "Neuroendocrine_adult" 

filtered_ligands = ligand_prioritized_tbl_oi %>% filter(receiver == receiver_oi) %>% pull(ligand) %>% unique()

prioritized_tbl_oi = prioritization_tables$prioritization_tbl_ligand_receptor %>% filter(ligand %in% filtered_ligands) %>% select(niche, sender, receiver, ligand,  receptor, ligand_receptor, prioritization_score) %>% distinct() %>% inner_join(top_ligand_receptor_niche_df) %>% group_by(ligand) %>% filter(receiver == receiver_oi) %>% top_n(2, prioritization_score) %>% ungroup() 
```

```{r echo=FALSE, fig.height=12, fig.width=8, message=FALSE, warning=FALSE}
lfc_plot = make_ligand_receptor_lfc_plot(receiver_oi, prioritized_tbl_oi, prioritization_tables$prioritization_tbl_ligand_receptor, plot_legend = FALSE, heights = NULL, widths = NULL)
lfc_plot
```

Recent studies have demonstrated that epiregulin (EREG) is upregulated in various cancer types, which contributes to cancer progression by triggering the EGFR signaling pathway.

ADAM17 is contributory to the occurrence and development of cancers

ytokine CXCL12 (also known as stromal-derived factor 1α) and its receptor CXCR4 represent the most promising actionable targets for this strategy. Both are overexpressed in various cancer types, and this aberrant expression strongly promotes proliferation, migration and invasion through multiple signal pathways.


```{r fig.height=10, fig.width=34, include=FALSE}
exprs_activity_target_plot = make_ligand_activity_target_exprs_plot(receiver_oi, prioritized_tbl_oi,  prioritization_tables$prioritization_tbl_ligand_receptor,  prioritization_tables$prioritization_tbl_ligand_target, output$exprs_tbl_ligand,  output$exprs_tbl_target, lfc_cutoff, ligand_target_matrix, plot_legend = FALSE, heights = NULL, widths = NULL)
exprs_activity_target_plot$combined_plot
```

```{r echo=FALSE, fig.height=10, fig.width=34, message=FALSE, warning=FALSE}
exprs_activity_target_plot = make_ligand_activity_target_exprs_plot(receiver_oi, prioritized_tbl_oi,  prioritization_tables$prioritization_tbl_ligand_receptor, 
                                                                    prioritization_tables$prioritization_tbl_ligand_target, output$exprs_tbl_ligand,  output$exprs_tbl_target, lfc_cutoff, ligand_target_matrix,
                                                                    scaled_ligand_activity_limits = "IQR", plot_legend = FALSE, heights = NULL, widths = NULL)
exprs_activity_target_plot$combined_plot
```

```{r message=FALSE, warning=FALSE, include=FALSE}

filtered_ligands = ligand_prioritized_tbl_oi %>% filter(receiver == receiver_oi) %>% top_n(15, prioritization_score) %>% pull(ligand) %>% unique()

prioritized_tbl_oi = prioritization_tables$prioritization_tbl_ligand_receptor %>% filter(ligand %in% filtered_ligands) %>% select(niche, sender, receiver, ligand,  receptor, ligand_receptor, prioritization_score) %>% distinct() %>% inner_join(top_ligand_receptor_niche_df) %>% group_by(ligand) %>% filter(receiver == receiver_oi) %>% top_n(2, prioritization_score) %>% ungroup() 

colors_sender = brewer.pal(n = prioritized_tbl_oi$sender %>% unique() %>% sort() %>% length(), name = 'Spectral') %>% magrittr::set_names(prioritized_tbl_oi$sender %>% unique() %>% sort())
colors_receiver = c("lavender")  %>% magrittr::set_names(prioritized_tbl_oi$receiver %>% unique() %>% sort())

colors_sender[1] <- '#F08080FF'
colors_receiver[1] <- '#32CD32FF'
circos_output = make_circos_lr(prioritized_tbl_oi, colors_sender, colors_receiver)
```

```{r echo=FALSE, fig.height=5, fig.width=10}
circos_output
```


# Heatmaps

```{r include=FALSE}
# Human gene names
human_genes <- c(
  "BCL2", "MAPK1", "BAX", "NFKB1", "JUN", "TNF", "IL6", "AP1", "CXCL1", "CCL5", "CCL7", 
  "SLC39A10", "LCN2", "ZC3H12A", "ENPP2", "MMP10", "MMP1", "MMP2", "MMP9", 
  "PIK3CA", "RAC1",
  "CASP1", "CASP3", "CASP4", "CASP5", "CASP7", "CASP8", "CASP9", "CASP10", "CASP14",
  "XAF1", "ULK1", "XIAP", "H2AFX", "TP53BP1", "ATG13", "RB1CC1", "AMBRA1", "BECN1", 
  "PIK3C3", "ATG14", "UVRAG", "ATG5", "ATG12", "ATG16L1", "MAP1LC3A", "ATG7", "ATG3", 
  "ATG10", "RAB7A", "LAMP1", "LAMP2", "STX17", "MTOR", "PRKAA1", "TFEB", "BAK1", 
  "BNIP3", "BBC3", "PMAIP1", "FAS", "MCL1", "BIRC5", "DNM1L", "VDAC1", "MPTP", 
  "TFAM", "ATP5F1A", "TNFRSF1A", "DDIT3", "ATF4", "TLR4", "AGER", "P2RX7", "NLRP3", 
  "CAPN1", "NFE2L2", "CYCS", "GSK3B", "IL18", "AIFM1", "NOX1", "NOX2", "NOX3", "NOX4", "NOX5", 
  "SOD1", "SOD2", "SOD3", "CAT", "GPX1", "GPX4", "TXN", "TXNRD1", "PRDX1", "PRDX2", "PRDX3", 
  "XDH", "AHR", 
  # Fibroblast related
  "VIM", "STAT3", "SMAD2", "SMAD3", "IL33", "IL1B", "ACTA2", "CXCL8", "COL1A1", "PDGFB",
  # Macrophage related
  "ARG1", "P2RX7", "SOCS3", "NLRP3", "MYD88", "NOS2"
)

# Mouse gene names
mouse_genes <- c(
  "Bcl2", "Mapk1", "Bax", "Nfkb1", "Jun", "Tnf", "Il6", "Ap1", "Cxcl1", "Ccl5", "Ccl7", 
  "Slc39a10", "Lcn2", "Zc3h12a", "Enpp2", "Mmp10", "Mmp1", "Mmp2", "Mmp9", 
  "Pik3ca", "Rac1",
  "Casp1", "Casp3", "Casp4", "Casp5", "Casp7", "Casp8", "Casp9", "Casp10", "Casp14",
  "Xaf1", "Ulk1", "Xiap", "H2afx", "Trp53bp1", "Atg13", "Rb1cc1", "Ambra1", "Becn1", 
  "Pik3c3", "Atg14", "Uvrag", "Atg5", "Atg12", "Atg16l1", "Map1lc3a", "Atg7", "Atg3", 
  "Atg10", "Rab7a", "Lamp1", "Lamp2", "Stx17", "Mtor", "Prkaa1", "Tfeb", "Bak1", 
  "Bnip3", "Bbc3", "Pmaip1", "Fas", "Mcl1", "Birc5", "Dnm1l", "Vdac1", "Mptp", 
  "Tfam", "Atp5f1a", "Tnfrsf1a", "Ddit3", "Atf4", "Tlr4", "Ager", "P2rx7", "Nlrp3", 
  "Capn1", "Nfe2l2", "Cycs", "Gsk3b", "Il18", "Aifm1", "Nox1", "Nox2", "Nox3", "Nox4", "Nox5", 
  "Sod1", "Sod2", "Sod3", "Cat", "Gpx1", "Gpx4", "Txn", "Txnrd1", "Prdx1", "Prdx2", "Prdx3", 
  "Xdh", "Ahr", 
  # Fibroblast related
  "Vim", "Stat3", "Smad2", "Smad3", "Il33", "Il1b", "Acta2", "Cxcl8", "Col1a1", "Pdgfb",
  # Macrophage related
  "Arg1", "P2rx7", "Socs3", "Nlrp3", "Myd88", "Nos2"
)
```

```{r include=FALSE}
DEG_sc_mouse <- qread(paste0(PREPRDATADIR, 'DEG_sc_mouse.qs'))
DEG_bulk_human <- qread(paste0(PREPRDATADIR, 'DEG_ATLAS_pseudobulk_norep_results.qs'))
mouse <- qread(paste0(PREPRDATADIR, 'seu_mouse_neutrophils.qs'))
```

```{r echo=FALSE, fig.height=8, fig.width=5}
for (i in unique(DEG_sc_mouse$celltype)) {
highlight_genes <- unique(mouse_genes[mouse_genes %in% (DEG_sc_mouse %>% dplyr::filter(celltype == i & p_val_adj < 0.05 & abs(avg_log2FC) > 0.15) %>% pull(gene))])

# 3. Plot with EnhancedVolcano
print(EnhancedVolcano(
  DEG_sc_mouse %>% dplyr::filter(celltype == i),
  lab = DEG_sc_mouse %>% dplyr::filter(celltype == i) %>% pull(gene),
  selectLab = highlight_genes,
  x = 'avg_log2FC',
  y = 'p_val_adj',
  col = c("grey30", "#8FBC8BFF", "#6495EDFF", "#BA55D3FF"),
  drawConnectors = TRUE,
  pCutoff = 0.05,
  FCcutoff = 0.5,
  max.overlaps = Inf,  # allow all 17 labels to show
  titleLabSize = 14,
  subtitleLabSize = 12,
  captionLabSize = 12,
  legendLabSize = 12,
  legendIconSize = 3,
  title = paste(i, 'mouse')
) + labs(subtitle = "young vs adult"))
}
```

```{r fig.height=8, fig.width=5}
for (i in names(DEG_bulk_human$meta_results)) {
  highlight_genes <- unique(human_genes[human_genes %in% (DEG_bulk_human$meta_results[[i]] %>% dplyr::filter(FDR < 0.05 & abs(mean_logFC) > 0.15) %>% pull(gene))])

print(EnhancedVolcano(
  DEG_bulk_human$meta_results[[i]],
  lab = DEG_bulk_human$meta_results[[i]]$gene,
  selectLab = highlight_genes,
  x = 'mean_logFC',
  y = 'FDR',
  col = c("grey30", "#8FBC8BFF", "#6495EDFF", "#BA55D3FF"),
  drawConnectors = TRUE,
  pCutoff = 0.05,
  FCcutoff = 0.5,
  max.overlaps = Inf,  # allow all 17 labels to show
  titleLabSize = 14,
  subtitleLabSize = 12,
  captionLabSize = 12,
  legendLabSize = 12,
  legendIconSize = 3,
  title = paste(i, 'human')
) + labs(subtitle = "< 18 months vs > 18 months"))
}
```

```{r eval=FALSE, include=FALSE}
# =================================================================
# SAVE ONLY CORE NICHENET DATA - Calculate tops/filters in Python
# =================================================================

# 1. TARGET GENE LISTS (Core results)
write.csv(data.frame(gene = geneset_NEyoung), paste0(OUTNAME, "nichenet_targets_young.csv"), row.names = FALSE)
write.csv(data.frame(gene = geneset_NEadult), paste0(OUTNAME, "nichenet_targets_adult.csv"), row.names = FALSE)

# 2. MAIN PRIORITIZATION TABLES (All ligand-receptor data)
write.csv(prioritization_tables$prioritization_tbl_ligand_receptor, 
          paste0(OUTNAME, "nichenet_ligand_receptor_prioritization.csv"), row.names = FALSE)

# 3. LIGAND-TARGET RELATIONSHIPS (All ligand-target data)  
write.csv(prioritization_tables$prioritization_tbl_ligand_target, 
          paste0(OUTNAME, "nichenet_ligand_target_prioritization.csv"), row.names = FALSE)

# Print what was saved
cat("Saved 4 core files:\n")
cat("1. nichenet_targets_young.csv -", length(geneset_NEyoung), "target genes\n")
cat("2. nichenet_targets_adult.csv -", length(geneset_NEadult), "target genes\n") 
cat("3. nichenet_ligand_receptor_prioritization.csv - All L-R pairs\n")
cat("4. nichenet_ligand_target_prioritization.csv - All L-T relationships\n")
```


```{r}
# 1. Number of differentially expressed ligands and receptors
n_de_ligands_young <- DE_sender %>% filter(sender == "Neutrophils_young", p_val_adj < 0.05) %>% nrow()
n_de_ligands_adult <- DE_sender %>% filter(sender == "Neutrophils_adult", p_val_adj < 0.05) %>% nrow()

n_de_receptors_young <- DE_receiver %>% filter(receiver == "Neuroendocrine_young", p_val_adj < 0.05) %>% nrow()
n_de_receptors_adult <- DE_receiver %>% filter(receiver == "Neuroendocrine_adult", p_val_adj < 0.05) %>% nrow()

# 2. Number of significantly upregulated target genes
n_targets_young <- length(geneset_NEyoung)
n_targets_adult <- length(geneset_NEadult)

# 3. Number of ligand-receptor pairs with high prioritization scores
lr_top_young <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(receiver == "Neuroendocrine_young", prioritization_score > 0.4) %>% nrow()
lr_top_adult <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(receiver == "Neuroendocrine_adult", prioritization_score > 0.4) %>% nrow()

# 4. Number of unique ligands prioritized per niche
n_unique_ligands_young <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(receiver == "Neuroendocrine_young") %>%
  pull(ligand) %>% unique() %>% length()

n_unique_ligands_adult <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(receiver == "Neuroendocrine_adult") %>%
  pull(ligand) %>% unique() %>% length()

# 5. Top ligand names (for discussion)
top_ligands_young <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(receiver == "Neuroendocrine_young") %>%
  arrange(desc(prioritization_score)) %>%
  slice_head(n = 5) %>%
  pull(ligand) %>% unique()

top_ligands_adult <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(receiver == "Neuroendocrine_adult") %>%
  arrange(desc(prioritization_score)) %>%
  slice_head(n = 5) %>%
  pull(ligand) %>% unique()

# Output for thesis
cat("\nNicheNet Summary:\n")
cat("- Young niche: ", n_de_ligands_young, "DE ligands, ", n_de_receptors_young, "DE receptors, ", n_targets_young, "target genes\n")
cat("- Adult niche: ", n_de_ligands_adult, "DE ligands, ", n_de_receptors_adult, "DE receptors, ", n_targets_adult, "target genes\n")
cat("- Ligand-Receptor pairs with score > 0.4: ", lr_top_young, " (young), ", lr_top_adult, " (adult)\n")
cat("- Unique prioritized ligands: ", n_unique_ligands_young, " (young), ", n_unique_ligands_adult, " (adult)\n")
cat("- Top ligands (young): ", paste(top_ligands_young, collapse = ", "), "\n")
cat("- Top ligands (adult): ", paste(top_ligands_adult, collapse = ", "), "\n")
```

```{r}
# ===============================================================
# NICHENET ANALYSIS RESULTS QUANTIFICATION FOR THESIS
# ===============================================================

# Load required libraries
library(tidyverse)
library(knitr)

# ===============================================================
# 1. TARGET GENE ANALYSIS
# ===============================================================

# Basic target gene statistics
cat("=== TARGET GENE ANALYSIS ===\n")
cat("Young niche target genes:", length(geneset_NEyoung), "\n")
cat("Adult niche target genes:", length(geneset_NEadult), "\n")

# Overlap analysis
target_overlap <- intersect(geneset_NEyoung, geneset_NEadult)
target_young_specific <- setdiff(geneset_NEyoung, geneset_NEadult)
target_adult_specific <- setdiff(geneset_NEadult, geneset_NEyoung)

cat("Overlapping target genes:", length(target_overlap), "\n")
cat("Young-specific target genes:", length(target_young_specific), "\n")
cat("Adult-specific target genes:", length(target_adult_specific), "\n")
cat("Percent overlap (young):", round(length(target_overlap)/length(geneset_NEyoung)*100, 1), "%\n")
cat("Percent overlap (adult):", round(length(target_overlap)/length(geneset_NEadult)*100, 1), "%\n")

# ===============================================================
# 2. LIGAND-RECEPTOR PAIR ANALYSIS
# ===============================================================

cat("\n=== LIGAND-RECEPTOR PAIR ANALYSIS ===\n")

# Total L-R pairs identified
total_lr_pairs <- nrow(prioritization_tables$prioritization_tbl_ligand_receptor)
cat("Total ligand-receptor pairs analyzed:", total_lr_pairs, "\n")

# High-priority pairs by niche (top quartile)
# First check for missing values
cat("Missing values in prioritization scores:", sum(is.na(prioritization_tables$prioritization_tbl_ligand_receptor$prioritization_score)), "\n")

# Remove missing values for threshold calculation
valid_scores <- prioritization_tables$prioritization_tbl_ligand_receptor$prioritization_score[!is.na(prioritization_tables$prioritization_tbl_ligand_receptor$prioritization_score)]
high_priority_threshold <- quantile(valid_scores, 0.75)

cat("High priority threshold (75th percentile):", round(high_priority_threshold, 3), "\n")

young_high_priority <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(niche == "young_niche", !is.na(prioritization_score), prioritization_score >= high_priority_threshold) %>%
  nrow()

adult_high_priority <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(niche == "old_niche", !is.na(prioritization_score), prioritization_score >= high_priority_threshold) %>%
  nrow()

cat("High-priority L-R pairs (young niche):", young_high_priority, "\n")
cat("High-priority L-R pairs (adult niche):", adult_high_priority, "\n")

# Top ligands per niche
top_n_show <- 10
young_top_ligands <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(niche == "young_niche", !is.na(prioritization_score)) %>%
  group_by(ligand) %>%
  summarise(max_score = max(prioritization_score, na.rm = TRUE), .groups = 'drop') %>%
  top_n(top_n_show, max_score) %>%
  arrange(desc(max_score))

adult_top_ligands <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(niche == "old_niche", !is.na(prioritization_score)) %>%
  group_by(ligand) %>%
  summarise(max_score = max(prioritization_score, na.rm = TRUE), .groups = 'drop') %>%
  top_n(top_n_show, max_score) %>%
  arrange(desc(max_score))

cat("\nTop", top_n_show, "ligands in young niche:\n")
print(young_top_ligands)

cat("\nTop", top_n_show, "ligands in adult niche:\n")
print(adult_top_ligands)

# Unique ligands per niche
young_ligands <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(niche == "young_niche") %>%
  pull(ligand) %>%
  unique()

adult_ligands <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(niche == "old_niche") %>%
  pull(ligand) %>%
  unique()

ligand_overlap <- intersect(young_ligands, adult_ligands)
young_specific_ligands <- setdiff(young_ligands, adult_ligands)
adult_specific_ligands <- setdiff(adult_ligands, young_ligands)

cat("\nUnique ligands in young niche:", length(young_ligands), "\n")
cat("Unique ligands in adult niche:", length(adult_ligands), "\n")
cat("Overlapping ligands:", length(ligand_overlap), "\n")
cat("Young-specific ligands:", length(young_specific_ligands), "\n")
cat("Adult-specific ligands:", length(adult_specific_ligands), "\n")

# ===============================================================
# 3. LIGAND ACTIVITY ANALYSIS
# ===============================================================

cat("\n=== LIGAND ACTIVITY ANALYSIS ===\n")

# Ligand activity statistics
young_activities <- prioritization_tables$prioritization_tbl_ligand_target %>%
  filter(niche == "young_niche") %>%
  select(ligand, activity) %>%
  distinct()

adult_activities <- prioritization_tables$prioritization_tbl_ligand_target %>%
  filter(niche == "old_niche") %>%
  select(ligand, activity) %>%
  distinct()

cat("Ligands with activity scores (young):", nrow(young_activities), "\n")
cat("Ligands with activity scores (adult):", nrow(adult_activities), "\n")
cat("Mean activity score (young):", round(mean(young_activities$activity, na.rm=T), 3), "\n")
cat("Mean activity score (adult):", round(mean(adult_activities$activity, na.rm=T), 3), "\n")

# High-activity ligands (top 25%)
young_high_activity <- young_activities %>%
  filter(activity >= quantile(activity, 0.75, na.rm=T)) %>%
  nrow()

adult_high_activity <- adult_activities %>%
  filter(activity >= quantile(activity, 0.75, na.rm=T)) %>%
  nrow()

cat("High-activity ligands (young):", young_high_activity, "\n")
cat("High-activity ligands (adult):", adult_high_activity, "\n")

# ===============================================================
# 4. SENDER-RECEIVER ANALYSIS
# ===============================================================

cat("\n=== SENDER-RECEIVER ANALYSIS ===\n")

# Sender cell type analysis
sender_summary <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  group_by(niche, sender) %>%
  summarise(
    n_ligands = n_distinct(ligand),
    n_receptors = n_distinct(receptor),
    n_lr_pairs = n(),
    mean_score = mean(prioritization_score, na.rm=T),
    .groups = 'drop'
  )

print(sender_summary)

# Receiver analysis
receiver_summary <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  group_by(niche, receiver) %>%
  summarise(
    n_ligands = n_distinct(ligand),
    n_receptors = n_distinct(receptor),
    n_lr_pairs = n(),
    mean_score = mean(prioritization_score, na.rm=T),
    .groups = 'drop'
  )

print(receiver_summary)

# ===============================================================
# 5. KEY S100A8/S100A9 ANALYSIS
# ===============================================================

cat("\n=== S100A8/S100A9 SPECIFIC ANALYSIS ===\n")

# Check if S100A8/A9 are in the results
s100_genes <- c("S100A8", "S100A9")
s100_in_ligands <- s100_genes[s100_genes %in% prioritization_tables$prioritization_tbl_ligand_receptor$ligand]

if(length(s100_in_ligands) > 0) {
  s100_results <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
    filter(ligand %in% s100_in_ligands)
  
  cat("S100A8/A9 ligands found in results:", paste(s100_in_ligands, collapse=", "), "\n")
  cat("Number of S100A8/A9 L-R pairs:", nrow(s100_results), "\n")
  
  # Show S100 results by niche
  s100_summary <- s100_results %>%
    group_by(niche, ligand) %>%
    summarise(
      n_receptors = n(),
      max_score = max(prioritization_score),
      mean_score = mean(prioritization_score),
      .groups = 'drop'
    )
  
  print(s100_summary)
} else {
  cat("S100A8/A9 not found in ligand-receptor results\n")
}

# ===============================================================
# 6. DIFFERENTIAL EXPRESSION CONTEXT
# ===============================================================

cat("\n=== DIFFERENTIAL EXPRESSION CONTEXT ===\n")

# DE sender analysis
de_sender_sig <- DE_sender %>%
  filter(p_val_adj < 0.05)

cat("Significantly DE ligands in senders:", nrow(de_sender_sig), "\n")

# By sender type
de_by_sender <- de_sender_sig %>%
  group_by(sender) %>%
  summarise(
    n_de_ligands = n(),
    n_upregulated = sum(avg_log2FC > 0),
    n_downregulated = sum(avg_log2FC < 0),
    .groups = 'drop'
  )

print(de_by_sender)

# DE receiver analysis
de_receiver_sig <- DE_receiver %>%
  filter(p_val_adj < 0.05)

cat("\nSignificantly DE receptors in receivers:", nrow(de_receiver_sig), "\n")

# By receiver type
de_by_receiver <- de_receiver_sig %>%
  group_by(receiver) %>%
  summarise(
    n_de_receptors = n(),
    n_upregulated = sum(avg_log2FC > 0),
    n_downregulated = sum(avg_log2FC < 0),
    .groups = 'drop'
  )

print(de_by_receiver)

# ===============================================================
# 7. CREATE SUMMARY TABLE FOR THESIS
# ===============================================================

# Create a comprehensive summary table
summary_stats <- data.frame(
  Metric = c(
    "Target genes (young niche)",
    "Target genes (adult niche)",
    "Overlapping target genes",
    "Total L-R pairs analyzed",
    "High-priority L-R pairs (young)",
    "High-priority L-R pairs (adult)",
    "Unique ligands (young)",
    "Unique ligands (adult)",
    "Ligands with activity scores (young)",
    "Ligands with activity scores (adult)",
    "Significantly DE ligands",
    "Significantly DE receptors"
  ),
  Value = c(
    length(geneset_NEyoung),
    length(geneset_NEadult),
    length(target_overlap),
    total_lr_pairs,
    young_high_priority,
    adult_high_priority,
    length(young_ligands),
    length(adult_ligands),
    nrow(young_activities),
    nrow(adult_activities),
    nrow(de_sender_sig),
    nrow(de_receiver_sig)
  )
)

cat("\n=== SUMMARY TABLE FOR THESIS ===\n")
print(kable(summary_stats, format = "markdown"))

# ===============================================================
# 8. TOP FINDINGS FOR THESIS WRITING
# ===============================================================

cat("\n=== KEY FINDINGS FOR THESIS ===\n")
cat("1. Young niche shows", length(geneset_NEyoung), "target genes vs", length(geneset_NEadult), "in adult niche\n")
cat("2. ", round(length(target_overlap)/min(length(geneset_NEyoung), length(geneset_NEadult))*100, 1), "% target gene overlap between niches\n")
cat("3. ", total_lr_pairs, "total ligand-receptor pairs analyzed across both niches\n")
cat("4. Top young ligand:", young_top_ligands$ligand[1], "(score:", round(young_top_ligands$max_score[1], 3), ")\n")
cat("5. Top adult ligand:", adult_top_ligands$ligand[1], "(score:", round(adult_top_ligands$max_score[1], 3), ")\n")
if(length(s100_in_ligands) > 0) {
  cat("6. S100A8/A9 identified as active ligands in the analysis\n")
}

# # Save results for later use
# write.csv(summary_stats, "nichenet_summary_stats.csv", row.names = FALSE)
# write.csv(young_top_ligands, "nichenet_top_ligands_young.csv", row.names = FALSE)
# write.csv(adult_top_ligands, "nichenet_top_ligands_adult.csv", row.names = FALSE)
```

```{r}
# ===============================================================
# ADDITIONAL NICHENET ANALYSIS FOR DETAILED THESIS WRITING
# ===============================================================

# Get the top 5 ligands for each niche (for text insertion)
cat("=== TOP LIGANDS FOR THESIS TEXT ===\n")

# Young niche top ligands
young_top_5 <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(niche == "young_niche", !is.na(prioritization_score)) %>%
  group_by(ligand) %>%
  summarise(max_score = max(prioritization_score, na.rm = TRUE), .groups = 'drop') %>%
  top_n(5, max_score) %>%
  arrange(desc(max_score))

cat("Top 5 young niche ligands:\n")
for(i in 1:nrow(young_top_5)) {
  cat(paste0(i, ". ", young_top_5$ligand[i], " (score: ", round(young_top_5$max_score[i], 3), ")\n"))
}

# Adult niche top ligands  
adult_top_5 <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(niche == "old_niche", !is.na(prioritization_score)) %>%
  group_by(ligand) %>%
  summarise(max_score = max(prioritization_score, na.rm = TRUE), .groups = 'drop') %>%
  top_n(5, max_score) %>%
  arrange(desc(max_score))

cat("\nTop 5 adult niche ligands:\n")
for(i in 1:nrow(adult_top_5)) {
  cat(paste0(i, ". ", adult_top_5$ligand[i], " (score: ", round(adult_top_5$max_score[i], 3), ")\n"))
}

# ===============================================================
# S100A8/A9 DETAILED ANALYSIS
# ===============================================================

cat("\n=== S100A8/A9 DETAILED ANALYSIS ===\n")

s100_detailed <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(ligand %in% c("S100A8", "S100A9")) %>%
  select(niche, sender, receiver, ligand, receptor, prioritization_score) %>%
  arrange(desc(prioritization_score))

cat("S100A8/A9 ligand-receptor pairs:\n")
print(s100_detailed)

# S100A8/A9 receptors
s100_receptors <- s100_detailed %>%
  pull(receptor) %>%
  unique()

cat("\nReceptors for S100A8/A9:", paste(s100_receptors, collapse = ", "), "\n")

# S100A8/A9 by niche
s100_by_niche <- s100_detailed %>%
  group_by(niche, ligand) %>%
  summarise(
    n_receptors = n(),
    max_score = max(prioritization_score, na.rm = TRUE),
    mean_score = mean(prioritization_score, na.rm = TRUE),
    .groups = 'drop'
  )

print(s100_by_niche)

# ===============================================================
# LIGAND COMPARISON BETWEEN NICHES
# ===============================================================

cat("\n=== LIGAND COMPARISON ANALYSIS ===\n")

# Compare ligand rankings between niches
young_ligand_ranks <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(niche == "young_niche", !is.na(prioritization_score)) %>%
  group_by(ligand) %>%
  summarise(young_max_score = max(prioritization_score, na.rm = TRUE), .groups = 'drop') %>%
  mutate(young_rank = rank(-young_max_score))

adult_ligand_ranks <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  filter(niche == "old_niche", !is.na(prioritization_score)) %>%
  group_by(ligand) %>%
  summarise(adult_max_score = max(prioritization_score, na.rm = TRUE), .groups = 'drop') %>%
  mutate(adult_rank = rank(-adult_max_score))

# Merge and find ligands with biggest ranking differences
ligand_rank_comparison <- young_ligand_ranks %>%
  full_join(adult_ligand_ranks, by = "ligand") %>%
  mutate(
    young_max_score = ifelse(is.na(young_max_score), 0, young_max_score),
    adult_max_score = ifelse(is.na(adult_max_score), 0, adult_max_score),
    young_rank = ifelse(is.na(young_rank), max(young_rank, na.rm = TRUE) + 1, young_rank),
    adult_rank = ifelse(is.na(adult_rank), max(adult_rank, na.rm = TRUE) + 1, adult_rank),
    rank_difference = young_rank - adult_rank,
    score_difference = young_max_score - adult_max_score
  )

# Ligands much higher in young niche
young_enriched <- ligand_rank_comparison %>%
  filter(rank_difference < -50) %>%  # Much better rank in young
  arrange(rank_difference) %>%
  head(10)

cat("Ligands enriched in young niche (better ranking):\n")
print(young_enriched %>% select(ligand, young_rank, adult_rank, rank_difference))

# Ligands much higher in adult niche  
adult_enriched <- ligand_rank_comparison %>%
  filter(rank_difference > 50) %>%  # Much better rank in adult
  arrange(desc(rank_difference)) %>%
  head(10)

cat("\nLigands enriched in adult niche (better ranking):\n")
print(adult_enriched %>% select(ligand, young_rank, adult_rank, rank_difference))

# ===============================================================
# TARGET GENE ANALYSIS
# ===============================================================

cat("\n=== TARGET GENE OVERLAP ANALYSIS ===\n")

# Age-specific target genes
young_specific_targets <- setdiff(geneset_NEyoung, geneset_NEadult)
adult_specific_targets <- setdiff(geneset_NEadult, geneset_NEyoung)
shared_targets <- intersect(geneset_NEyoung, geneset_NEadult)

cat("Young-specific targets (first 10):", paste(head(young_specific_targets, 10), collapse = ", "), "\n")
cat("Adult-specific targets (first 10):", paste(head(adult_specific_targets, 10), collapse = ", "), "\n")
cat("Shared targets (first 10):", paste(head(shared_targets, 10), collapse = ", "), "\n")

# Save gene lists for pathway analysis
write.csv(data.frame(gene = young_specific_targets), "young_specific_targets.csv", row.names = FALSE)
write.csv(data.frame(gene = adult_specific_targets), "adult_specific_targets.csv", row.names = FALSE)
write.csv(data.frame(gene = shared_targets), "shared_targets.csv", row.names = FALSE)

# ===============================================================
# CREATE TEXT SNIPPETS FOR THESIS
# ===============================================================

cat("\n=== TEXT SNIPPETS FOR THESIS ===\n")

# Create formatted text for top ligands
young_ligand_text <- paste0(young_top_5$ligand[2:4], collapse = ", ")
adult_ligand_text <- paste0(adult_top_5$ligand[2:4], collapse = ", ")

cat("For 'Additional highly-ranked ligands' in young niche:", young_ligand_text, "\n")
cat("For 'Additional top-ranking ligands' in adult niche:", adult_ligand_text, "\n")

# Summary statistics formatted for text
cat("\nFormatted statistics for thesis:\n")
cat("- Target gene overlap: 756 out of", min(length(geneset_NEyoung), length(geneset_NEadult)), "genes (82.6%)\n")
cat("- Young-specific targets:", length(young_specific_targets), "genes\n") 
cat("- Adult-specific targets:", length(adult_specific_targets), "genes\n")
cat("- Total L-R pairs analyzed: 9,710\n")
cat("- High-priority pairs (young): 1,082\n")
cat("- High-priority pairs (adult): 1,330\n")
cat("- S100A8/A9 receptor targets:", paste(s100_receptors, collapse = ", "), "\n")
```

```{r}
# ===============================================================
# PUBLICATION-READY TABLES FOR NICHENET RESULTS
# ===============================================================

library(knitr)
library(kableExtra)

# ===============================================================
# TABLE 1: TOP LIGANDS BY NICHE
# ===============================================================

# Create comprehensive top ligands table
top_ligands_table <- rbind(
  young_top_5 %>% mutate(Niche = "Young") %>% 
    select(Niche, Ligand = ligand, Score = max_score),
  adult_top_5 %>% mutate(Niche = "Adult") %>% 
    select(Niche, Ligand = ligand, Score = max_score)
) %>%
  mutate(Score = round(Score, 3)) %>%
  arrange(Niche, desc(Score))

cat("=== TABLE 1: TOP LIGANDS BY NICHE ===\n")
print(kable(top_ligands_table, format = "markdown", 
            caption = "Top 5 prioritized ligands in neutrophil-neuroendocrine communication by age niche"))

# ===============================================================
# TABLE 2: S100A8/A9 ANALYSIS SUMMARY
# ===============================================================

s100_summary_table <- s100_by_niche %>%
  mutate(
    Niche = case_when(
      niche == "young_niche" ~ "Young",
      niche == "old_niche" ~ "Adult"
    ),
    Ligand = ligand,
    `Receptor Pairs` = n_receptors,
    `Max Score` = round(max_score, 3),
    `Mean Score` = round(mean_score, 3)
  ) %>%
  select(Niche, Ligand, `Receptor Pairs`, `Max Score`, `Mean Score`)

cat("\n=== TABLE 2: S100A8/A9 ANALYSIS SUMMARY ===\n")
print(kable(s100_summary_table, format = "markdown",
            caption = "S100A8/A9 ligand activity by age niche"))

# ===============================================================
# TABLE 3: TARGET GENE SUMMARY
# ===============================================================

target_summary_table <- data.frame(
  Category = c("Young-specific targets", "Adult-specific targets", "Shared targets", "Total young targets", "Total adult targets"),
  Count = c(length(young_specific_targets), length(adult_specific_targets), 
            length(shared_targets), length(geneset_NEyoung), length(geneset_NEadult)),
  Percentage = c(
    round(length(young_specific_targets)/length(geneset_NEyoung)*100, 1),
    round(length(adult_specific_targets)/length(geneset_NEadult)*100, 1),
    round(length(shared_targets)/min(length(geneset_NEyoung), length(geneset_NEadult))*100, 1),
    100, 100
  )
)

cat("\n=== TABLE 3: TARGET GENE SUMMARY ===\n")
print(kable(target_summary_table, format = "markdown",
            caption = "Target gene distribution by age niche"))

# ===============================================================
# TABLE 4: DIFFERENTIAL EXPRESSION SUMMARY
# ===============================================================

de_summary_table <- rbind(
  de_by_sender %>% mutate(Cell_Type = paste0(sender, " (Sender)")) %>%
    select(Cell_Type, `Total DE` = n_de_ligands, Upregulated = n_upregulated, Downregulated = n_downregulated),
  de_by_receiver %>% mutate(Cell_Type = paste0(receiver, " (Receiver)")) %>%
    select(Cell_Type, `Total DE` = n_de_receptors, Upregulated = n_upregulated, Downregulated = n_downregulated)
)

cat("\n=== TABLE 4: DIFFERENTIAL EXPRESSION SUMMARY ===\n")
print(kable(de_summary_table, format = "markdown",
            caption = "Differential expression of ligands and receptors by cell type and age"))

# ===============================================================
# TABLE 5: LIGAND ENRICHMENT EXTREMES
# ===============================================================

# Create table of most enriched ligands in each niche
enrichment_table <- rbind(
  young_enriched %>% head(5) %>% mutate(Enriched_In = "Young") %>%
    select(Enriched_In, Ligand = ligand, `Young Rank` = young_rank, `Adult Rank` = adult_rank, `Rank Difference` = rank_difference),
  adult_enriched %>% head(5) %>% mutate(Enriched_In = "Adult") %>%
    select(Enriched_In, Ligand = ligand, `Young Rank` = young_rank, `Adult Rank` = adult_rank, `Rank Difference` = rank_difference)
) %>%
  mutate(
    `Young Rank` = as.integer(`Young Rank`),
    `Adult Rank` = as.integer(`Adult Rank`),
    `Rank Difference` = as.integer(`Rank Difference`)
  )

cat("\n=== TABLE 5: MOST AGE-ENRICHED LIGANDS ===\n")
print(kable(enrichment_table, format = "markdown",
            caption = "Ligands showing strongest age-specific enrichment (top 5 per niche)"))

# ===============================================================
# FIGURE LEGENDS
# ===============================================================

cat("\n=== SUGGESTED FIGURE LEGENDS ===\n")

cat("Figure X: Age-dependent ligand-receptor communication networks in neuroblastoma\n")
cat("(A) Circos plot showing top 15 ligand-receptor pairs in young niche. ")
cat("(B) Circos plot showing top 15 ligand-receptor pairs in adult niche. ")
cat("Ribbon thickness indicates prioritization score strength. ")
cat("Colors represent sender (neutrophils) and receiver (neuroendocrine) cell types.\n\n")

cat("Figure Y: S100A8/A9 validation in human neuroblastoma\n")
cat("(A) Prioritization scores for S100A8/A9-receptor pairs by age niche. ")
cat("(B) Expression levels of S100A8/A9 receptors in neuroendocrine cells. ")
cat("(C) Pathway analysis of S100A8/A9 target genes. ")
cat("Statistical significance: *p<0.05, **p<0.01, ***p<0.001.\n\n")

cat("Figure Z: Age-specific target gene signatures\n")
cat("(A) Heatmap of young-specific target genes across samples. ")
cat("(B) Heatmap of adult-specific target genes. ")
cat("(C) Pathway enrichment analysis comparing young vs adult signatures. ")
cat("Genes clustered by expression pattern, samples ordered by age group.\n\n")

# ===============================================================
# EXPORT TABLES FOR MANUSCRIPT
# ===============================================================

# Save as CSV files for easy import into manuscripts
write.csv(top_ligands_table, "Table1_Top_Ligands_By_Niche.csv", row.names = FALSE)
write.csv(s100_summary_table, "Table2_S100A8A9_Summary.csv", row.names = FALSE)
write.csv(target_summary_table, "Table3_Target_Gene_Summary.csv", row.names = FALSE)
write.csv(de_summary_table, "Table4_DE_Summary.csv", row.names = FALSE)
write.csv(enrichment_table, "Table5_Age_Enriched_Ligands.csv", row.names = FALSE)

# Save detailed results for supplementary material
write.csv(young_enriched, "Supplementary_Young_Enriched_Ligands.csv", row.names = FALSE)
write.csv(adult_enriched, "Supplementary_Adult_Enriched_Ligands.csv", row.names = FALSE)
write.csv(s100_detailed, "Supplementary_S100A8A9_Detailed.csv", row.names = FALSE)

cat("\n=== FILES EXPORTED ===\n")
cat("Main tables: Table1-5 saved as CSV files\n")
cat("Supplementary data: Detailed ligand rankings and S100A8/A9 analysis\n")
cat("Target gene lists: young_specific_targets.csv, adult_specific_targets.csv, shared_targets.csv\n")

# ===============================================================
# STATISTICAL SUMMARY FOR RESULTS TEXT
# ===============================================================

cat("\n=== STATISTICAL SUMMARY FOR RESULTS SECTION ===\n")
cat("Use these exact numbers in your thesis:\n\n")

cat("• NicheNet identified", length(geneset_NEyoung), "target genes in young niche and", 
    length(geneset_NEadult), "in adult niche\n")
cat("• Target gene overlap:", length(shared_targets), "genes (", 
    round(length(shared_targets)/min(length(geneset_NEyoung), length(geneset_NEadult))*100, 1), "%)\n")
cat("• Young-specific targets:", length(young_specific_targets), "genes (",
    round(length(young_specific_targets)/length(geneset_NEyoung)*100, 1), "% of young targets)\n")
cat("• Adult-specific targets:", length(adult_specific_targets), "genes (",
    round(length(adult_specific_targets)/length(geneset_NEadult)*100, 1), "% of adult targets)\n")
cat("• Total ligand-receptor pairs analyzed: 9,710\n")
cat("• High-priority pairs (young niche):", 1082, "\n")
cat("• High-priority pairs (adult niche):", 1330, "\n")
cat("• S100A8/A9 receptor targets: 7 unique receptors (ALCAM, CD36, AGER, CD68, ITGB2, CD69, TLR4)\n")
cat("• Top young ligand: IFITM1 (score: 0.768)\n")
cat("• Top adult ligand: APOE (score: 0.797)\n")
cat("• Most young-enriched ligand: FCGR3B (rank difference: -1162)\n")
cat("• Most adult-enriched ligand: CCL19 (rank difference: +1178)\n")
```


# Enrichment

```{r}
# ===============================================================
# COMPREHENSIVE PATHWAY ENRICHMENT ANALYSIS FOR NICHENET RESULTS
# ===============================================================

# Load required libraries
library(clusterProfiler)
library(org.Hs.eg.db)
library(DOSE)
library(enrichplot)
library(ggplot2)
library(dplyr)
library(ReactomePA)
library(msigdbr)
library(tidyr)
library(RColorBrewer)
library(pheatmap)
library(VennDiagram)
library(UpSetR)

# Set up
set.seed(123)

# ===============================================================
# 1. PREPARE GENE LISTS AND BACKGROUND
# ===============================================================

cat("=== PREPARING GENE LISTS ===\n")

# Your gene lists (already created)
young_genes <- young_specific_targets
adult_genes <- adult_specific_targets
shared_genes <- shared_targets
background_genes <- background

# Convert to Entrez IDs for enrichment analysis
convert_to_entrez <- function(gene_symbols) {
  entrez_ids <- bitr(gene_symbols, fromType = "SYMBOL", toType = "ENTREZID", 
                     OrgDb = org.Hs.eg.db, drop = TRUE)
  return(entrez_ids$ENTREZID)
}

young_entrez <- convert_to_entrez(young_genes)
adult_entrez <- convert_to_entrez(adult_genes)
shared_entrez <- convert_to_entrez(shared_genes)
background_entrez <- convert_to_entrez(background_genes)

cat("Young genes converted:", length(young_entrez), "out of", length(young_genes), "\n")
cat("Adult genes converted:", length(adult_entrez), "out of", length(adult_genes), "\n")
cat("Shared genes converted:", length(shared_entrez), "out of", length(shared_genes), "\n")
cat("Background genes:", length(background_entrez), "\n")

# ===============================================================
# 2. GO ENRICHMENT ANALYSIS
# ===============================================================

cat("\n=== GO ENRICHMENT ANALYSIS ===\n")

# GO Biological Process
go_young_bp <- enrichGO(gene = young_entrez,
                        universe = background_entrez,
                        OrgDb = org.Hs.eg.db,
                        ont = "BP",
                        pAdjustMethod = "BH",
                        pvalueCutoff = 0.05,
                        qvalueCutoff = 0.2,
                        readable = TRUE)

go_adult_bp <- enrichGO(gene = adult_entrez,
                        universe = background_entrez,
                        OrgDb = org.Hs.eg.db,
                        ont = "BP",
                        pAdjustMethod = "BH",
                        pvalueCutoff = 0.05,
                        qvalueCutoff = 0.2,
                        readable = TRUE)

go_shared_bp <- enrichGO(gene = shared_entrez,
                         universe = background_entrez,
                         OrgDb = org.Hs.eg.db,
                         ont = "BP",
                         pAdjustMethod = "BH",
                         pvalueCutoff = 0.05,
                         qvalueCutoff = 0.2,
                         readable = TRUE)

# Print results
cat("Young-specific GO BP terms:", nrow(go_young_bp@result), "\n")
cat("Adult-specific GO BP terms:", nrow(go_adult_bp@result), "\n")
cat("Shared GO BP terms:", nrow(go_shared_bp@result), "\n")

if(nrow(go_young_bp@result) > 0) {
  cat("\nTop 10 Young-specific GO BP terms:\n")
  print(go_young_bp@result[1:min(10, nrow(go_young_bp@result)), c("Description", "pvalue", "p.adjust", "Count")])
}

if(nrow(go_adult_bp@result) > 0) {
  cat("\nTop 10 Adult-specific GO BP terms:\n")
  print(go_adult_bp@result[1:min(10, nrow(go_adult_bp@result)), c("Description", "pvalue", "p.adjust", "Count")])
}

# GO Molecular Function
go_young_mf <- enrichGO(gene = young_entrez,
                        universe = background_entrez,
                        OrgDb = org.Hs.eg.db,
                        ont = "MF",
                        pAdjustMethod = "BH",
                        pvalueCutoff = 0.05,
                        qvalueCutoff = 0.2,
                        readable = TRUE)

go_adult_mf <- enrichGO(gene = adult_entrez,
                        universe = background_entrez,
                        OrgDb = org.Hs.eg.db,
                        ont = "MF",
                        pAdjustMethod = "BH",
                        pvalueCutoff = 0.05,
                        qvalueCutoff = 0.2,
                        readable = TRUE)

# ===============================================================
# 3. KEGG PATHWAY ENRICHMENT
# ===============================================================

cat("\n=== KEGG PATHWAY ENRICHMENT ===\n")

kegg_young <- enrichKEGG(gene = young_entrez,
                         universe = background_entrez,
                         organism = 'hsa',
                         pvalueCutoff = 0.05,
                         pAdjustMethod = "BH",
                         qvalueCutoff = 0.2)

kegg_adult <- enrichKEGG(gene = adult_entrez,
                         universe = background_entrez,
                         organism = 'hsa',
                         pvalueCutoff = 0.05,
                         pAdjustMethod = "BH",
                         qvalueCutoff = 0.2)

kegg_shared <- enrichKEGG(gene = shared_entrez,
                          universe = background_entrez,
                          organism = 'hsa',
                          pvalueCutoff = 0.05,
                          pAdjustMethod = "BH",
                          qvalueCutoff = 0.2)

cat("Young-specific KEGG pathways:", nrow(kegg_young@result), "\n")
cat("Adult-specific KEGG pathways:", nrow(kegg_adult@result), "\n")
cat("Shared KEGG pathways:", nrow(kegg_shared@result), "\n")

if(nrow(kegg_young@result) > 0) {
  cat("\nTop Young-specific KEGG pathways:\n")
  print(kegg_young@result[, c("Description", "pvalue", "p.adjust", "Count")])
}

if(nrow(kegg_adult@result) > 0) {
  cat("\nTop Adult-specific KEGG pathways:\n")
  print(kegg_adult@result[, c("Description", "pvalue", "p.adjust", "Count")])
}

# ===============================================================
# 4. REACTOME PATHWAY ENRICHMENT
# ===============================================================

cat("\n=== REACTOME PATHWAY ENRICHMENT ===\n")

reactome_young <- enrichPathway(gene = young_entrez,
                                universe = background_entrez,
                                pvalueCutoff = 0.05,
                                pAdjustMethod = "BH",
                                qvalueCutoff = 0.2,
                                readable = TRUE)

reactome_adult <- enrichPathway(gene = adult_entrez,
                                universe = background_entrez,
                                pvalueCutoff = 0.05,
                                pAdjustMethod = "BH",
                                qvalueCutoff = 0.2,
                                readable = TRUE)

reactome_shared <- enrichPathway(gene = shared_entrez,
                                 universe = background_entrez,
                                 pvalueCutoff = 0.05,
                                 pAdjustMethod = "BH",
                                 qvalueCutoff = 0.2,
                                 readable = TRUE)

cat("Young-specific Reactome pathways:", nrow(reactome_young@result), "\n")
cat("Adult-specific Reactome pathways:", nrow(reactome_adult@result), "\n")
cat("Shared Reactome pathways:", nrow(reactome_shared@result), "\n")

if(nrow(reactome_young@result) > 0) {
  cat("\nTop Young-specific Reactome pathways:\n")
  print(reactome_young@result[1:min(10, nrow(reactome_young@result)), c("Description", "pvalue", "p.adjust", "Count")])
}

if(nrow(reactome_adult@result) > 0) {
  cat("\nTop Adult-specific Reactome pathways:\n")
  print(reactome_adult@result[1:min(10, nrow(reactome_adult@result)), c("Description", "pvalue", "p.adjust", "Count")])
}

# ===============================================================
# 5. HALLMARK GENE SETS (MSigDB)
# ===============================================================

cat("\n=== HALLMARK GENE SET ENRICHMENT ===\n")

# Get Hallmark gene sets
hallmark_sets <- msigdbr(species = "Homo sapiens", category = "H")

hallmark_young <- enricher(gene = young_genes,
                           universe = background_genes,
                           TERM2GENE = hallmark_sets[, c("gs_name", "gene_symbol")],
                           pvalueCutoff = 0.05,
                           pAdjustMethod = "BH",
                           qvalueCutoff = 0.2)

hallmark_adult <- enricher(gene = adult_genes,
                           universe = background_genes,
                           TERM2GENE = hallmark_sets[, c("gs_name", "gene_symbol")],
                           pvalueCutoff = 0.05,
                           pAdjustMethod = "BH",
                           qvalueCutoff = 0.2)

hallmark_shared <- enricher(gene = shared_genes,
                            universe = background_genes,
                            TERM2GENE = hallmark_sets[, c("gs_name", "gene_symbol")],
                            pvalueCutoff = 0.05,
                            pAdjustMethod = "BH",
                            qvalueCutoff = 0.2)

cat("Young-specific Hallmark pathways:", nrow(hallmark_young@result), "\n")
cat("Adult-specific Hallmark pathways:", nrow(hallmark_adult@result), "\n")
cat("Shared Hallmark pathways:", nrow(hallmark_shared@result), "\n")

if(nrow(hallmark_young@result) > 0) {
  cat("\nYoung-specific Hallmark pathways:\n")
  print(hallmark_young@result[, c("Description", "pvalue", "p.adjust", "Count")])
}

if(nrow(hallmark_adult@result) > 0) {
  cat("\nAdult-specific Hallmark pathways:\n")
  print(hallmark_adult@result[, c("Description", "pvalue", "p.adjust", "Count")])
}

# ===============================================================
# 6. CANCER-RELATED GENE SETS (C6 ONCOGENIC SIGNATURES)
# ===============================================================

cat("\n=== ONCOGENIC SIGNATURE ENRICHMENT ===\n")

# Get oncogenic signature gene sets
oncogenic_sets <- msigdbr(species = "Homo sapiens", category = "C6")

oncogenic_young <- enricher(gene = young_genes,
                            universe = background_genes,
                            TERM2GENE = oncogenic_sets[, c("gs_name", "gene_symbol")],
                            pvalueCutoff = 0.05,
                            pAdjustMethod = "BH",
                            qvalueCutoff = 0.2)

oncogenic_adult <- enricher(gene = adult_genes,
                            universe = background_genes,
                            TERM2GENE = oncogenic_sets[, c("gs_name", "gene_symbol")],
                            pvalueCutoff = 0.05,
                            pAdjustMethod = "BH",
                            qvalueCutoff = 0.2)

cat("Young-specific oncogenic signatures:", nrow(oncogenic_young@result), "\n")
cat("Adult-specific oncogenic signatures:", nrow(oncogenic_adult@result), "\n")

if(nrow(oncogenic_young@result) > 0) {
  cat("\nTop Young-specific oncogenic signatures:\n")
  print(oncogenic_young@result[1:min(5, nrow(oncogenic_young@result)), c("Description", "pvalue", "p.adjust", "Count")])
}

if(nrow(oncogenic_adult@result) > 0) {
  cat("\nTop Adult-specific oncogenic signatures:\n")
  print(oncogenic_adult@result[1:min(5, nrow(oncogenic_adult@result)), c("Description", "pvalue", "p.adjust", "Count")])
}

# ===============================================================
# 7. SAVE ENRICHMENT RESULTS
# ===============================================================

cat("\n=== SAVING ENRICHMENT RESULTS ===\n")

# Function to save enrichment results
save_enrichment_results <- function(enrich_obj, filename) {
  if(nrow(enrich_obj@result) > 0) {
    write.csv(enrich_obj@result, filename, row.names = FALSE)
    return(TRUE)
  } else {
    cat("No significant results for", filename, "\n")
    return(FALSE)
  }
}

# Save all results
save_enrichment_results(go_young_bp, "GO_BP_Young_Specific.csv")
save_enrichment_results(go_adult_bp, "GO_BP_Adult_Specific.csv")
save_enrichment_results(go_shared_bp, "GO_BP_Shared.csv")
save_enrichment_results(go_young_mf, "GO_MF_Young_Specific.csv")
save_enrichment_results(go_adult_mf, "GO_MF_Adult_Specific.csv")
save_enrichment_results(kegg_young, "KEGG_Young_Specific.csv")
save_enrichment_results(kegg_adult, "KEGG_Adult_Specific.csv")
save_enrichment_results(kegg_shared, "KEGG_Shared.csv")
save_enrichment_results(reactome_young, "Reactome_Young_Specific.csv")
save_enrichment_results(reactome_adult, "Reactome_Adult_Specific.csv")
save_enrichment_results(reactome_shared, "Reactome_Shared.csv")
save_enrichment_results(hallmark_young, "Hallmark_Young_Specific.csv")
save_enrichment_results(hallmark_adult, "Hallmark_Adult_Specific.csv")
save_enrichment_results(hallmark_shared, "Hallmark_Shared.csv")
save_enrichment_results(oncogenic_young, "Oncogenic_Young_Specific.csv")
save_enrichment_results(oncogenic_adult, "Oncogenic_Adult_Specific.csv")

# ===============================================================
# 8. CREATE SUMMARY STATISTICS
# ===============================================================

cat("\n=== ENRICHMENT SUMMARY STATISTICS ===\n")

enrichment_summary <- data.frame(
  Database = c("GO_BP", "GO_MF", "KEGG", "Reactome", "Hallmark", "Oncogenic"),
  Young_Pathways = c(
    nrow(go_young_bp@result[go_young_bp@result$p.adjust < 0.05,]),
    nrow(go_young_mf@result[go_young_mf@result$p.adjust < 0.05,]),
    nrow(kegg_young@result[kegg_young@result$p.adjust < 0.05,]),
    nrow(reactome_young@result[reactome_young@result$p.adjust < 0.05,]),
    nrow(hallmark_young@result[hallmark_young@result$p.adjust < 0.05,]),
    nrow(oncogenic_young@result[oncogenic_young@result$p.adjust < 0.05,])
  ),
  Adult_Pathways = c(
    nrow(go_adult_bp@result[go_adult_bp@result$p.adjust < 0.05,]),
    nrow(go_adult_mf@result[go_adult_mf@result$p.adjust < 0.05,]),
    nrow(kegg_adult@result[kegg_adult@result$p.adjust < 0.05,]),
    nrow(reactome_adult@result[reactome_adult@result$p.adjust < 0.05,]),
    nrow(hallmark_adult@result[hallmark_adult@result$p.adjust < 0.05,]),
    nrow(oncogenic_adult@result[oncogenic_adult@result$p.adjust < 0.05,])
  ),
  Shared_Pathways = c(
    nrow(go_shared_bp@result[go_shared_bp@result$p.adjust < 0.05,]),
    NA,  # No MF for shared
    nrow(kegg_shared@result[kegg_shared@result$p.adjust < 0.05,]),
    nrow(reactome_shared@result[reactome_shared@result$p.adjust < 0.05,]),
    nrow(hallmark_shared@result[hallmark_shared@result$p.adjust < 0.05,]),
    NA   # No oncogenic for shared
  )
)

print(enrichment_summary)
write.csv(enrichment_summary, "Enrichment_Summary_Statistics.csv", row.names = FALSE)

cat("\n=== ENRICHMENT ANALYSIS COMPLETE ===\n")
cat("All results saved as CSV files\n")
cat("Summary statistics saved as Enrichment_Summary_Statistics.csv\n")
```

```{r}
# ===============================================================
# ENRICHMENT ANALYSIS VISUALIZATIONS
# ===============================================================

# Load additional libraries for advanced visualizations
library(ggplot2)
library(enrichplot)
library(dplyr)
library(tidyr)
library(RColorBrewer)
library(cowplot)
library(ggrepel)
library(viridis)
library(ComplexHeatmap)
library(circlize)

# Set up color palettes
young_color <- "#6495EDFF"  # Blue
adult_color <- "#FF69B4FF"  # Pink  
shared_color <- "#32CD32FF" # Green

# ===============================================================
# 1. DOT PLOTS FOR PATHWAY ENRICHMENT
# ===============================================================

cat("=== CREATING DOT PLOTS ===\n")

# Function to create publication-ready dot plots
create_dotplot <- function(enrich_obj, title, color = "red", top_n = 15) {
  if(nrow(enrich_obj@result) == 0) {
    return(ggplot() + ggtitle(paste("No significant results for", title)) + theme_minimal())
  }
  
  dotplot(enrich_obj, showCategory = top_n, 
          title = title, font.size = 10) +
    scale_color_gradient(low = "blue", high = color) +
    theme(axis.text.y = element_text(size = 9),
          plot.title = element_text(size = 12, hjust = 0.5))
}

# Create dot plots for GO BP
p1 <- create_dotplot(go_young_bp, "Young-Specific GO Biological Process", young_color)
p2 <- create_dotplot(go_adult_bp, "Adult-Specific GO Biological Process", adult_color)
p3 <- create_dotplot(go_shared_bp, "Shared GO Biological Process", shared_color)

# Save GO BP plots
ggsave("GO_BP_Young_Dotplot.pdf", p1, width = 12, height = 8)
ggsave("GO_BP_Adult_Dotplot.pdf", p2, width = 12, height = 8)
ggsave("GO_BP_Shared_Dotplot.pdf", p3, width = 12, height = 8)

# Create dot plots for KEGG
p4 <- create_dotplot(kegg_young, "Young-Specific KEGG Pathways", young_color)
p5 <- create_dotplot(kegg_adult, "Adult-Specific KEGG Pathways", adult_color)
p6 <- create_dotplot(kegg_shared, "Shared KEGG Pathways", shared_color)

# Save KEGG plots
# ggsave("KEGG_Young_Dotplot.pdf", p4, width = 12, height = 8)
ggsave("KEGG_Adult_Dotplot.pdf", p5, width = 12, height = 8)
ggsave("KEGG_Shared_Dotplot.pdf", p6, width = 12, height = 8)

# Create dot plots for Reactome
p7 <- create_dotplot(reactome_young, "Young-Specific Reactome Pathways", young_color)
p8 <- create_dotplot(reactome_adult, "Adult-Specific Reactome Pathways", adult_color)
p9 <- create_dotplot(reactome_shared, "Shared Reactome Pathways", shared_color)

# Save Reactome plots
ggsave("Reactome_Young_Dotplot.pdf", p7, width = 12, height = 8)
ggsave("Reactome_Adult_Dotplot.pdf", p8, width = 12, height = 8)
ggsave("Reactome_Shared_Dotplot.pdf", p9, width = 12, height = 8)

# ===============================================================
# 2. BAR PLOTS FOR HALLMARK PATHWAYS
# ===============================================================

cat("=== CREATING BAR PLOTS FOR HALLMARK PATHWAYS ===\n")

# Function to create bar plots
create_barplot <- function(enrich_obj, title, color = "red", top_n = 10) {
  if(nrow(enrich_obj@result) == 0) {
    return(ggplot() + ggtitle(paste("No significant results for", title)) + theme_minimal())
  }
  
  barplot(enrich_obj, showCategory = top_n, title = title) +
    scale_fill_gradient(low = "lightblue", high = color) +
    theme(axis.text.y = element_text(size = 9),
          plot.title = element_text(size = 12, hjust = 0.5))
}

# Hallmark bar plots
h1 <- create_barplot(hallmark_young, "Young-Specific Hallmark Pathways", young_color)
h2 <- create_barplot(hallmark_adult, "Adult-Specific Hallmark Pathways", adult_color)
h3 <- create_barplot(hallmark_shared, "Shared Hallmark Pathways", shared_color)

ggsave("Hallmark_Young_Barplot.pdf", h1, width = 10, height = 6)
ggsave("Hallmark_Adult_Barplot.pdf", h2, width = 10, height = 6)
ggsave("Hallmark_Shared_Barplot.pdf", h3, width = 10, height = 6)

# ===============================================================
# 3. COMBINED COMPARISON PLOTS
# ===============================================================

cat("=== CREATING COMPARISON PLOTS ===\n")

# Function to extract top pathways for comparison
extract_top_pathways <- function(enrich_obj, label, top_n = 10) {
  if(nrow(enrich_obj@result) == 0) {
    return(data.frame())
  }
  
  result <- enrich_obj@result[1:min(top_n, nrow(enrich_obj@result)), ]
  result$Group <- label
  result$NegLogP <- -log10(result$p.adjust)
  return(result[, c("Description", "Group", "NegLogP", "Count", "p.adjust")])
}

# Extract top GO BP pathways for comparison
go_comparison <- rbind(
  extract_top_pathways(go_young_bp, "Young", 8),
  extract_top_pathways(go_adult_bp, "Adult", 8)
)

if(nrow(go_comparison) > 0) {
  # Create comparison plot
  comparison_plot <- ggplot(go_comparison, aes(x = reorder(Description, NegLogP), y = NegLogP, fill = Group)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_manual(values = c("Young" = young_color, "Adult" = adult_color)) +
    coord_flip() +
    labs(title = "Top GO Biological Process: Young vs Adult",
         x = "Pathway", y = "-log10(Adjusted P-value)",
         fill = "Age Group") +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 8),
          plot.title = element_text(size = 12, hjust = 0.5))
  
  ggsave("GO_BP_Young_vs_Adult_Comparison.pdf", comparison_plot, width = 14, height = 10)
}

# Similar comparison for KEGG
kegg_comparison <- rbind(
  extract_top_pathways(kegg_young, "Young", 8),
  extract_top_pathways(kegg_adult, "Adult", 8)
)

if(nrow(kegg_comparison) > 0) {
  kegg_comparison_plot <- ggplot(kegg_comparison, aes(x = reorder(Description, NegLogP), y = NegLogP, fill = Group)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_manual(values = c("Young" = young_color, "Adult" = adult_color)) +
    coord_flip() +
    labs(title = "Top KEGG Pathways: Young vs Adult",
         x = "Pathway", y = "-log10(Adjusted P-value)",
         fill = "Age Group") +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 8),
          plot.title = element_text(size = 12, hjust = 0.5))
  
  ggsave("KEGG_Young_vs_Adult_Comparison.pdf", kegg_comparison_plot, width = 14, height = 10)
}

# ===============================================================
# 4. NETWORK PLOTS FOR TOP PATHWAYS
# ===============================================================

cat("=== CREATING NETWORK PLOTS ===\n")

# Create network plots for enrichment results
create_network_plot <- function(enrich_obj, title, filename) {
  if(nrow(enrich_obj@result) == 0) {
    cat("No results for network plot:", title, "\n")
    return(NULL)
  }
  
  tryCatch({
    # Simplify terms to avoid overcrowding
    enrich_simplified <- simplify(enrich_obj, cutoff = 0.7, by = "p.adjust", select_fun = min)
    
    if(nrow(enrich_simplified@result) > 0) {
      p <- emapplot(enrich_simplified, showCategory = 15, cex_label_category = 0.6) +
        ggtitle(title)
      ggsave(filename, p, width = 12, height = 10)
      return(p)
    }
  }, error = function(e) {
    cat("Error creating network plot for", title, ":", e$message, "\n")
    return(NULL)
  })
}

# Create network plots (only for results with sufficient pathways)
if(nrow(go_young_bp@result) > 5) {
  create_network_plot(go_young_bp, "Young GO BP Network", "GO_BP_Young_Network.pdf")
}
if(nrow(go_adult_bp@result) > 5) {
  create_network_plot(go_adult_bp, "Adult GO BP Network", "GO_BP_Adult_Network.pdf")
}
if(nrow(reactome_young@result) > 5) {
  create_network_plot(reactome_young, "Young Reactome Network", "Reactome_Young_Network.pdf")
}
if(nrow(reactome_adult@result) > 5) {
  create_network_plot(reactome_adult, "Adult Reactome Network", "Reactome_Adult_Network.pdf")
}

# ===============================================================
# 5. HEATMAP OF ENRICHMENT SCORES
# ===============================================================

cat("=== CREATING ENRICHMENT HEATMAP ===\n")

# Function to create enrichment score matrix
create_enrichment_matrix <- function() {
  # Combine all enrichment results
  all_results <- list()
  
  # Helper function to extract scores
  extract_scores <- function(enrich_obj, db_name, group_name) {
    if(nrow(enrich_obj@result) == 0) return(data.frame())
    
    result <- enrich_obj@result
    result$Database <- db_name
    result$Group <- group_name
    result$NegLogP <- -log10(result$p.adjust)
    return(result[, c("Description", "Database", "Group", "NegLogP", "p.adjust")])
  }
  
  # Extract all results
  results_list <- list(
    extract_scores(go_young_bp, "GO_BP", "Young"),
    extract_scores(go_adult_bp, "GO_BP", "Adult"),
    extract_scores(kegg_young, "KEGG", "Young"),
    extract_scores(kegg_adult, "KEGG", "Adult"),
    extract_scores(reactome_young, "Reactome", "Young"),
    extract_scores(reactome_adult, "Reactome", "Adult"),
    extract_scores(hallmark_young, "Hallmark", "Young"),
    extract_scores(hallmark_adult, "Hallmark", "Adult")
  )
  
  # Combine and filter
  combined_results <- do.call(rbind, results_list[sapply(results_list, nrow) > 0])
  
  if(nrow(combined_results) == 0) {
    cat("No results for heatmap\n")
    return(NULL)
  }
  
  # Select top pathways per database per group
  top_pathways <- combined_results %>%
    group_by(Database, Group) %>%
    top_n(5, NegLogP) %>%
    ungroup()
  
  # Create matrix
  heatmap_data <- top_pathways %>%
    dplyr::select(Description, Group, NegLogP) %>%
    pivot_wider(names_from = Group, values_from = NegLogP, values_fill = 0) %>%
    column_to_rownames("Description")
  
  return(heatmap_data)
}

# Create and save heatmap
heatmap_matrix <- create_enrichment_matrix()

if(!is.null(heatmap_matrix) && nrow(heatmap_matrix) > 0) {
  # Create heatmap
  pdf("Enrichment_Heatmap.pdf", width = 10, height = 12)
  pheatmap(heatmap_matrix,
           color = colorRampPalette(c("white", "red"))(50),
           cluster_rows = TRUE,
           cluster_cols = FALSE,
           fontsize_row = 8,
           fontsize_col = 10,
           main = "Pathway Enrichment Scores (-log10 adj. p-value)",
           angle_col = 45)
  dev.off()
}

# ===============================================================
# 6. UPSET PLOT FOR PATHWAY OVERLAPS
# ===============================================================

cat("=== CREATING UPSET PLOT ===\n")

# Function to create pathway overlap data
create_pathway_overlap <- function() {
  # Extract pathway names from each category
  pathways_list <- list()
  
  if(nrow(go_young_bp@result) > 0) {
    pathways_list$GO_Young <- go_young_bp@result$Description[go_young_bp@result$p.adjust < 0.05]
  }
  if(nrow(go_adult_bp@result) > 0) {
    pathways_list$GO_Adult <- go_adult_bp@result$Description[go_adult_bp@result$p.adjust < 0.05]
  }
  if(nrow(kegg_young@result) > 0) {
    pathways_list$KEGG_Young <- kegg_young@result$Description[kegg_young@result$p.adjust < 0.05]
  }
  if(nrow(kegg_adult@result) > 0) {
    pathways_list$KEGG_Adult <- kegg_adult@result$Description[kegg_adult@result$p.adjust < 0.05]
  }
  
  return(pathways_list)
}

pathway_overlaps <- create_pathway_overlap()

if(length(pathway_overlaps) > 1) {
  # Create UpSet plot
  pdf("Pathway_Overlap_UpSet.pdf", width = 12, height = 8)
  upset(fromList(pathway_overlaps), 
        order.by = "freq",
        nsets = length(pathway_overlaps),
        text.scale = c(1.3, 1.3, 1, 1, 2, 0.75),
        mainbar.y.label = "Pathway Intersections",
        sets.x.label = "Pathways per Category")
  dev.off()
}

# ===============================================================
# 7. VOLCANO PLOTS FOR ENRICHMENT
# ===============================================================

cat("=== CREATING VOLCANO PLOTS ===\n")

# Function to create volcano plots
create_enrichment_volcano <- function(enrich_obj, title, filename) {
  if(nrow(enrich_obj@result) == 0) {
    cat("No data for volcano plot:", title, "\n")
    return(NULL)
  }
  
  result_df <- enrich_obj@result
  result_df$NegLogP <- -log10(result_df$p.adjust)
  result_df$Significant <- result_df$p.adjust < 0.05
  
  p <- ggplot(result_df, aes(x = Count, y = NegLogP, color = Significant)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "red")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
    labs(title = title,
         x = "Gene Count",
         y = "-log10(Adjusted P-value)") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
  
  # Add labels for top pathways
  if(sum(result_df$Significant) > 0) {
    top_pathways <- result_df[result_df$Significant, ][1:min(5, sum(result_df$Significant)), ]
    p <- p + geom_text_repel(data = top_pathways, 
                             aes(label = Description), 
                             color = "black", size = 3, max.overlaps = 10)
  }
  
  ggsave(filename, p, width = 12, height = 8)
  return(p)
}

# Create volcano plots
create_enrichment_volcano(go_young_bp, "Young GO BP Enrichment", "Volcano_GO_Young.pdf")
create_enrichment_volcano(go_adult_bp, "Adult GO BP Enrichment", "Volcano_GO_Adult.pdf")
create_enrichment_volcano(reactome_young, "Young Reactome Enrichment", "Volcano_Reactome_Young.pdf")
create_enrichment_volcano(reactome_adult, "Adult Reactome Enrichment", "Volcano_Reactome_Adult.pdf")

# ===============================================================
# 8. COMBINED SUMMARY VISUALIZATION
# ===============================================================

cat("=== CREATING SUMMARY VISUALIZATION ===\n")

# Create summary bar plot of enrichment counts
summary_data <- data.frame(
  Database = rep(c("GO_BP", "GO_MF", "KEGG", "Reactome", "Hallmark"), 2),
  Group = rep(c("Young", "Adult"), each = 5),
  Count = c(
    nrow(go_young_bp@result[go_young_bp@result$p.adjust < 0.05,]),
    nrow(go_young_mf@result[go_young_mf@result$p.adjust < 0.05,]),
    nrow(kegg_young@result[kegg_young@result$p.adjust < 0.05,]),
    nrow(reactome_young@result[reactome_young@result$p.adjust < 0.05,]),
    nrow(hallmark_young@result[hallmark_young@result$p.adjust < 0.05,]),
    nrow(go_adult_bp@result[go_adult_bp@result$p.adjust < 0.05,]),
    nrow(go_adult_mf@result[go_adult_mf@result$p.adjust < 0.05,]),
    nrow(kegg_adult@result[kegg_adult@result$p.adjust < 0.05,]),
    nrow(reactome_adult@result[reactome_adult@result$p.adjust < 0.05,]),
    nrow(hallmark_adult@result[hallmark_adult@result$p.adjust < 0.05,])
  )
)

summary_plot <- ggplot(summary_data, aes(x = Database, y = Count, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Young" = young_color, "Adult" = adult_color)) +
  labs(title = "Number of Significantly Enriched Pathways by Database",
       x = "Database", y = "Number of Pathways (p.adj < 0.05)",
       fill = "Age Group") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

ggsave("Enrichment_Summary_Barplot.pdf", summary_plot, width = 10, height = 6)

# ===============================================================
# 9. GENE-CONCEPT NETWORK (if data available)
# ===============================================================

cat("=== CREATING GENE-CONCEPT NETWORKS ===\n")

# Function to create gene-concept network
create_gene_concept_network <- function(enrich_obj, title, filename) {
  if(nrow(enrich_obj@result) == 0) {
    cat("No data for gene-concept network:", title, "\n")
    return(NULL)
  }
  
  tryCatch({
    p <- cnetplot(enrich_obj, showCategory = 10, cex_label_gene = 0.5, cex_label_category = 0.8) +
      ggtitle(title)
    ggsave(filename, p, width = 14, height = 12)
    return(p)
  }, error = function(e) {
    cat("Error creating gene-concept network for", title, ":", e$message, "\n")
    return(NULL)
  })
}

# Create gene-concept networks for top results
if(nrow(go_young_bp@result) > 0) {
  create_gene_concept_network(go_young_bp, "Young GO BP Gene-Concept Network", "GeneConcept_GO_Young.pdf")
}
if(nrow(go_adult_bp@result) > 0) {
  create_gene_concept_network(go_adult_bp, "Adult GO BP Gene-Concept Network", "GeneConcept_GO_Adult.pdf")
}

# ===============================================================
# 10. PUBLICATION-READY MULTI-PANEL FIGURE
# ===============================================================

cat("=== CREATING MULTI-PANEL FIGURE ===\n")

# Create a comprehensive multi-panel figure
create_multipanel_figure <- function() {
  # Panel A: Summary bar plot
  panel_a <- summary_plot + ggtitle("A) Pathway Enrichment Summary")
  
  # Panel B: Top GO terms comparison (if available)
  if(nrow(go_comparison) > 0) {
    panel_b <- ggplot(go_comparison[1:min(16, nrow(go_comparison)),], 
                      aes(x = reorder(Description, NegLogP), y = NegLogP, fill = Group)) +
      geom_bar(stat = "identity", position = "dodge") +
      scale_fill_manual(values = c("Young" = young_color, "Adult" = adult_color)) +
      coord_flip() +
      labs(title = "B) Top GO Biological Processes",
           x = "", y = "-log10(Adjusted P-value)") +
      theme_minimal() +
      theme(axis.text.y = element_text(size = 7),
            legend.position = "none")
  } else {
    panel_b <- ggplot() + ggtitle("B) No GO comparison data") + theme_minimal()
  }
  
  # Panel C: Hallmark pathways (if available)
  hallmark_combined <- rbind(
    extract_top_pathways(hallmark_young, "Young", 5),
    extract_top_pathways(hallmark_adult, "Adult", 5)
  )
  
  if(nrow(hallmark_combined) > 0) {
    panel_c <- ggplot(hallmark_combined, aes(x = reorder(Description, NegLogP), y = NegLogP, fill = Group)) +
      geom_bar(stat = "identity", position = "dodge") +
      scale_fill_manual(values = c("Young" = young_color, "Adult" = adult_color)) +
      coord_flip() +
      labs(title = "C) Hallmark Gene Sets",
           x = "", y = "-log10(Adjusted P-value)") +
      theme_minimal() +
      theme(axis.text.y = element_text(size = 7),
            legend.position = "bottom")
  } else {
    panel_c <- ggplot() + ggtitle("C) No Hallmark data") + theme_minimal()
  }
  
  # Combine panels
  if(nrow(go_comparison) > 0 && nrow(hallmark_combined) > 0) {
    combined_figure <- plot_grid(panel_a, panel_b, panel_c, 
                                 ncol = 1, rel_heights = c(1, 1.5, 1.2))
  } else {
    combined_figure <- panel_a
  }
  
  ggsave("Figure_Enrichment_MultiPanel.pdf", combined_figure, width = 14, height = 16)
  return(combined_figure)
}

multipanel_fig <- create_multipanel_figure()

# ===============================================================
# 11. SAVE VISUALIZATION SUMMARY
# ===============================================================

cat("\n=== VISUALIZATION SUMMARY ===\n")
cat("Created the following visualization files:\n")
cat("1. Individual dotplots for GO, KEGG, Reactome (PDF)\n")
cat("2. Bar plots for Hallmark pathways (PDF)\n")
cat("3. Comparison plots (Young vs Adult) (PDF)\n")
cat("4. Network plots for pathway relationships (PDF)\n")
cat("5. Enrichment heatmap (PDF)\n")
cat("6. UpSet plot for pathway overlaps (PDF)\n")
cat("7. Volcano plots for enrichment significance (PDF)\n")
cat("8. Summary bar plot of pathway counts (PDF)\n")
cat("9. Gene-concept networks (PDF)\n")
cat("10. Multi-panel publication figure (PDF)\n")

# Create a list of all generated files
visualization_files <- c(
  "GO_BP_Young_Dotplot.pdf", "GO_BP_Adult_Dotplot.pdf", "GO_BP_Shared_Dotplot.pdf",
  "KEGG_Young_Dotplot.pdf", "KEGG_Adult_Dotplot.pdf", "KEGG_Shared_Dotplot.pdf",
  "Reactome_Young_Dotplot.pdf", "Reactome_Adult_Dotplot.pdf", "Reactome_Shared_Dotplot.pdf",
  "Hallmark_Young_Barplot.pdf", "Hallmark_Adult_Barplot.pdf", "Hallmark_Shared_Barplot.pdf",
  "GO_BP_Young_vs_Adult_Comparison.pdf", "KEGG_Young_vs_Adult_Comparison.pdf",
  "Enrichment_Heatmap.pdf", "Pathway_Overlap_UpSet.pdf",
  "Volcano_GO_Young.pdf", "Volcano_GO_Adult.pdf", "Volcano_Reactome_Young.pdf", "Volcano_Reactome_Adult.pdf",
  "Enrichment_Summary_Barplot.pdf", "Figure_Enrichment_MultiPanel.pdf"
)

cat("\nTotal visualization files created:", length(visualization_files), "\n")
cat("All files saved in current working directory\n")

# ===============================================================
# 12. CREATE INTERPRETATION GUIDE
# ===============================================================

cat("\n=== INTERPRETATION GUIDE ===\n")
cat("How to interpret your enrichment visualizations:\n\n")

cat("DOT PLOTS:\n")
cat("- X-axis: Gene ratio (proportion of genes in pathway)\n")
cat("- Y-axis: Pathway names\n")
cat("- Color intensity: Statistical significance (-log10 p.adj)\n")
cat("- Dot size: Number of genes in pathway\n\n")

cat("BAR PLOTS:\n")
cat("- Height: Statistical significance (-log10 p.adj)\n")
cat("- Color intensity: Significance level\n")
cat("- Useful for comparing pathway strength\n\n")

cat("VOLCANO PLOTS:\n")
cat("- X-axis: Gene count in pathway\n")
cat("- Y-axis: Statistical significance (-log10 p.adj)\n")
cat("- Red points: Significantly enriched (p.adj < 0.05)\n")
cat("- Labels: Top significant pathways\n\n")

cat("NETWORK PLOTS:\n")
cat("- Nodes: Pathways (circles) and genes (smaller dots)\n")
cat("- Edges: Gene-pathway relationships\n")
cat("- Node size: Statistical significance\n")
cat("- Useful for seeing pathway relationships\n\n")

cat("For your thesis:\n")
cat("- Use dot plots for main pathway results\n")
cat("- Use comparison plots to highlight age differences\n")
cat("- Use multi-panel figure for comprehensive overview\n")
cat("- Include network plots in supplementary material\n")

cat("\n=== ENRICHMENT VISUALIZATION COMPLETE ===\n")
```

```{r}
# ===============================================================
# ENRICHMENT RESULTS ANALYSIS AND THESIS TEXT GENERATION
# ===============================================================

# ===============================================================
# 1. EXTRACT KEY FINDINGS FOR THESIS
# ===============================================================

cat("=== KEY ENRICHMENT FINDINGS FOR THESIS ===\n")

# Function to get top pathways with clean descriptions
get_top_pathways_clean <- function(enrich_obj, n = 5) {
  if(nrow(enrich_obj@result) == 0) return(character(0))
  
  result <- enrich_obj@result[enrich_obj@result$p.adjust < 0.05, ]
  if(nrow(result) == 0) return(character(0))
  
  top_n <- head(result, n)
  return(top_n$Description)
}

# Extract top pathways for each category
young_go_top <- get_top_pathways_clean(go_young_bp, 5)
adult_go_top <- get_top_pathways_clean(go_adult_bp, 5)
young_kegg_top <- get_top_pathways_clean(kegg_young, 3)
adult_kegg_top <- get_top_pathways_clean(kegg_adult, 3)
young_reactome_top <- get_top_pathways_clean(reactome_young, 3)
adult_reactome_top <- get_top_pathways_clean(reactome_adult, 3)
young_hallmark_top <- get_top_pathways_clean(hallmark_young, 3)
adult_hallmark_top <- get_top_pathways_clean(hallmark_adult, 3)

# Print results for thesis
cat("YOUNG-SPECIFIC ENRICHED PATHWAYS:\n")
cat("GO Biological Process (top 5):\n")
if(length(young_go_top) > 0) {
  for(i in 1:length(young_go_top)) {
    cat(paste0(i, ". ", young_go_top[i], "\n"))
  }
} else {
  cat("No significant GO BP terms\n")
}

cat("\nKEGG Pathways (top 3):\n")
if(length(young_kegg_top) > 0) {
  for(i in 1:length(young_kegg_top)) {
    cat(paste0(i, ". ", young_kegg_top[i], "\n"))
  }
} else {
  cat("No significant KEGG pathways\n")
}

cat("\nReactome Pathways (top 3):\n")
if(length(young_reactome_top) > 0) {
  for(i in 1:length(young_reactome_top)) {
    cat(paste0(i, ". ", young_reactome_top[i], "\n"))
  }
} else {
  cat("No significant Reactome pathways\n")
}

cat("\nHallmark Gene Sets (top 3):\n")
if(length(young_hallmark_top) > 0) {
  for(i in 1:length(young_hallmark_top)) {
    cat(paste0(i, ". ", young_hallmark_top[i], "\n"))
  }
} else {
  cat("No significant Hallmark gene sets\n")
}

cat("\n", strrep("=", 50), "\n")
cat("ADULT-SPECIFIC ENRICHED PATHWAYS:\n")
cat("GO Biological Process (top 5):\n")
if(length(adult_go_top) > 0) {
  for(i in 1:length(adult_go_top)) {
    cat(paste0(i, ". ", adult_go_top[i], "\n"))
  }
} else {
  cat("No significant GO BP terms\n")
}

cat("\nKEGG Pathways (top 3):\n")
if(length(adult_kegg_top) > 0) {
  for(i in 1:length(adult_kegg_top)) {
    cat(paste0(i, ". ", adult_kegg_top[i], "\n"))
  }
} else {
  cat("No significant KEGG pathways\n")
}

cat("\nReactome Pathways (top 3):\n")
if(length(adult_reactome_top) > 0) {
  for(i in 1:length(adult_reactome_top)) {
    cat(paste0(i, ". ", adult_reactome_top[i], "\n"))
  }
} else {
  cat("No significant Reactome pathways\n")
}

cat("\nHallmark Gene Sets (top 3):\n")
if(length(adult_hallmark_top) > 0) {
  for(i in 1:length(adult_hallmark_top)) {
    cat(paste0(i, ". ", adult_hallmark_top[i], "\n"))
  }
} else {
  cat("No significant Hallmark gene sets\n")
}

# ===============================================================
# 2. STATISTICAL SUMMARY
# ===============================================================

cat("\n=== STATISTICAL SUMMARY FOR THESIS ===\n")

# Count significant pathways
count_significant <- function(enrich_obj) {
  return(nrow(enrich_obj@result[enrich_obj@result$p.adjust < 0.05, ]))
}

young_go_count <- count_significant(go_young_bp)
adult_go_count <- count_significant(go_adult_bp)
shared_go_count <- count_significant(go_shared_bp)
young_kegg_count <- count_significant(kegg_young)
adult_kegg_count <- count_significant(kegg_adult)
shared_kegg_count <- count_significant(kegg_shared)
young_reactome_count <- count_significant(reactome_young)
adult_reactome_count <- count_significant(reactome_adult)
shared_reactome_count <- count_significant(reactome_shared)
young_hallmark_count <- count_significant(hallmark_young)
adult_hallmark_count <- count_significant(hallmark_adult)
shared_hallmark_count <- count_significant(hallmark_shared)

cat("PATHWAY ENRICHMENT STATISTICS (p.adj < 0.05):\n")
cat("GO Biological Process:\n")
cat("  Young-specific:", young_go_count, "pathways\n")
cat("  Adult-specific:", adult_go_count, "pathways\n")
cat("  Shared:", shared_go_count, "pathways\n")

cat("KEGG Pathways:\n")
cat("  Young-specific:", young_kegg_count, "pathways\n")
cat("  Adult-specific:", adult_kegg_count, "pathways\n")
cat("  Shared:", shared_kegg_count, "pathways\n")

cat("Reactome Pathways:\n")
cat("  Young-specific:", young_reactome_count, "pathways\n")
cat("  Adult-specific:", adult_reactome_count, "pathways\n")
cat("  Shared:", shared_reactome_count, "pathways\n")

cat("Hallmark Gene Sets:\n")
cat("  Young-specific:", young_hallmark_count, "gene sets\n")
cat("  Adult-specific:", adult_hallmark_count, "gene sets\n")
cat("  Shared:", shared_hallmark_count, "gene sets\n")

# ===============================================================
# 3. BIOLOGICAL INTERPRETATION
# ===============================================================

cat("\n=== BIOLOGICAL INTERPRETATION ===\n")

# Analyze pathway themes
analyze_pathway_themes <- function(pathways, label) {
  if(length(pathways) == 0) {
    cat("No pathways to analyze for", label, "\n")
    return()
  }
  
  cat("Analysis for", label, ":\n")
  
  # Look for key themes
  immune_terms <- sum(grepl("immune|inflammation|cytokine|interferon|antigen", pathways, ignore.case = TRUE))
  development_terms <- sum(grepl("development|differentiation|morphogenesis", pathways, ignore.case = TRUE))
  metabolism_terms <- sum(grepl("metabol|synthesis|catabolism", pathways, ignore.case = TRUE))
  apoptosis_terms <- sum(grepl("apoptosis|death|survival", pathways, ignore.case = TRUE))
  proliferation_terms <- sum(grepl("proliferation|cycle|division|growth", pathways, ignore.case = TRUE))
  migration_terms <- sum(grepl("migration|motility|adhesion", pathways, ignore.case = TRUE))
  
  cat("  Immune/inflammation related:", immune_terms, "\n")
  cat("  Development/differentiation:", development_terms, "\n")
  cat("  Metabolism related:", metabolism_terms, "\n")
  cat("  Apoptosis/survival:", apoptosis_terms, "\n")
  cat("  Proliferation/growth:", proliferation_terms, "\n")
  cat("  Migration/adhesion:", migration_terms, "\n")
}

analyze_pathway_themes(young_go_top, "Young GO pathways")
analyze_pathway_themes(adult_go_top, "Adult GO pathways")

# ===============================================================
# 4. CREATE THESIS TEXT BLOCKS
# ===============================================================

cat("\n=== THESIS TEXT BLOCKS ===\n")

# Generate formatted text for thesis
generate_thesis_text <- function() {
  cat("COPY THIS TEXT FOR YOUR THESIS RESULTS SECTION:\n")
  cat("\n", strrep("=", 50), "\n")
  
  cat("## Pathway Enrichment Analysis Reveals Age-Dependent Functional Programs\n\n")
  
  cat("To understand the functional implications of age-specific target gene signatures, we performed comprehensive pathway enrichment analysis using multiple databases including Gene Ontology (GO), KEGG, Reactome, and Hallmark gene sets.\n\n")
  
  # GO Results
  if(young_go_count > 0 || adult_go_count > 0) {
    cat("Gene Ontology biological process analysis revealed", young_go_count, "significantly enriched pathways in young-specific targets and", adult_go_count, "in adult-specific targets (adjusted p < 0.05).")
    
    if(young_go_count > 0) {
      top_young_go <- paste(head(young_go_top, 3), collapse = ", ")
      cat(" Young-specific targets were enriched for pathways including", top_young_go, ".")
    }
    
    if(adult_go_count > 0) {
      top_adult_go <- paste(head(adult_go_top, 3), collapse = ", ")
      cat(" In contrast, adult-specific targets showed enrichment for", top_adult_go, ".")
    }
    cat("\n\n")
  }
  
  # KEGG Results
  if(young_kegg_count > 0 || adult_kegg_count > 0) {
    cat("KEGG pathway analysis identified", young_kegg_count, "significantly enriched pathways in young targets and", adult_kegg_count, "in adult targets.")
    
    if(young_kegg_count > 0 && length(young_kegg_top) > 0) {
      cat(" Young-specific enrichment included", paste(young_kegg_top, collapse = ", "), ".")
    }
    
    if(adult_kegg_count > 0 && length(adult_kegg_top) > 0) {
      cat(" Adult targets were enriched for", paste(adult_kegg_top, collapse = ", "), ".")
    }
    cat("\n\n")
  }
  
  # Reactome Results
  if(young_reactome_count > 0 || adult_reactome_count > 0) {
    cat("Reactome pathway analysis revealed", young_reactome_count, "pathways enriched in young targets and", adult_reactome_count, "in adult targets.")
    
    if(young_reactome_count > 0 && length(young_reactome_top) > 0) {
      cat(" Notable young-enriched pathways included", paste(head(young_reactome_top, 2), collapse = " and "), ".")
    }
    
    if(adult_reactome_count > 0 && length(adult_reactome_top) > 0) {
      cat(" Adult-enriched pathways included", paste(head(adult_reactome_top, 2), collapse = " and "), ".")
    }
    cat("\n\n")
  }
  
  # Hallmark Results
  if(young_hallmark_count > 0 || adult_hallmark_count > 0) {
    cat("Analysis of Hallmark gene sets, which represent well-defined biological states and processes, identified", young_hallmark_count, "sets enriched in young targets and", adult_hallmark_count, "in adult targets.")
    
    if(young_hallmark_count > 0 && length(young_hallmark_top) > 0) {
      cat(" Young-enriched Hallmark sets included", paste(young_hallmark_top, collapse = ", "), ".")
    }
    
    if(adult_hallmark_count > 0 && length(adult_hallmark_top) > 0) {
      cat(" Adult-enriched sets encompassed", paste(adult_hallmark_top, collapse = ", "), ".")
    }
    cat("\n\n")
  }
  
  # Shared pathways
  if(shared_go_count > 0 || shared_kegg_count > 0 || shared_reactome_count > 0) {
    cat("Shared target genes (", length(shared_genes), "genes) were enriched for", shared_go_count, "GO biological processes,", shared_kegg_count, "KEGG pathways, and", shared_reactome_count, "Reactome pathways, representing core neuroblastoma-associated processes conserved across age groups.\n\n")
  }
  
  cat("These pathway enrichment patterns provide mechanistic insights into age-dependent neuroblastoma progression, with young patients showing enrichment for [SUMMARIZE YOUNG THEMES] while adult patients exhibit activation of [SUMMARIZE ADULT THEMES]. The distinct functional programs support the hypothesis that age-dependent microenvironmental communication drives different cellular responses that ultimately influence clinical outcomes.\n")
  
  cat("\n", strrep("=", 50), "\n")
}

generate_thesis_text()

# ===============================================================
# 5. SAVE COMPREHENSIVE RESULTS SUMMARY
# ===============================================================

# Create comprehensive results table
create_comprehensive_results_table <- function() {
  results_summary <- data.frame(
    Analysis_Type = c("Young_GO_BP", "Adult_GO_BP", "Shared_GO_BP",
                     "Young_KEGG", "Adult_KEGG", "Shared_KEGG",
                     "Young_Reactome", "Adult_Reactome", "Shared_Reactome",
                     "Young_Hallmark", "Adult_Hallmark", "Shared_Hallmark"),
    Significant_Pathways = c(young_go_count, adult_go_count, shared_go_count,
                            young_kegg_count, adult_kegg_count, shared_kegg_count,
                            young_reactome_count, adult_reactome_count, shared_reactome_count,
                            young_hallmark_count, adult_hallmark_count, shared_hallmark_count),
    Top_Pathway = c(
      ifelse(length(young_go_top) > 0, young_go_top[1], "None"),
      ifelse(length(adult_go_top) > 0, adult_go_top[1], "None"),
      ifelse(length(get_top_pathways_clean(go_shared_bp, 1)) > 0, get_top_pathways_clean(go_shared_bp, 1)[1], "None"),
      ifelse(length(young_kegg_top) > 0, young_kegg_top[1], "None"),
      ifelse(length(adult_kegg_top) > 0, adult_kegg_top[1], "None"),
      ifelse(length(get_top_pathways_clean(kegg_shared, 1)) > 0, get_top_pathways_clean(kegg_shared, 1)[1], "None"),
      ifelse(length(young_reactome_top) > 0, young_reactome_top[1], "None"),
      ifelse(length(adult_reactome_top) > 0, adult_reactome_top[1], "None"),
      ifelse(length(get_top_pathways_clean(reactome_shared, 1)) > 0, get_top_pathways_clean(reactome_shared, 1)[1], "None"),
      ifelse(length(young_hallmark_top) > 0, young_hallmark_top[1], "None"),
      ifelse(length(adult_hallmark_top) > 0, adult_hallmark_top[1], "None"),
      ifelse(length(get_top_pathways_clean(hallmark_shared, 1)) > 0, get_top_pathways_clean(hallmark_shared, 1)[1], "None")
    )
  )
  return(results_summary)
}

comprehensive_results <- create_comprehensive_results_table()
write.csv(comprehensive_results, "Comprehensive_Enrichment_Results.csv", row.names = FALSE)

print(comprehensive_results)

# ===============================================================
# 6. DETAILED PATHWAY ANALYSIS FOR KEY FINDINGS
# ===============================================================

cat("\n=== DETAILED PATHWAY ANALYSIS ===\n")

# Function to extract detailed pathway information
extract_pathway_details <- function(enrich_obj, pathway_name) {
  if(nrow(enrich_obj@result) == 0) return(NULL)
  
  pathway_info <- enrich_obj@result[enrich_obj@result$Description == pathway_name, ]
  if(nrow(pathway_info) == 0) return(NULL)
  
  return(list(
    description = pathway_info$Description,
    pvalue = pathway_info$pvalue,
    p_adjust = pathway_info$p.adjust,
    gene_count = pathway_info$Count,
    genes = strsplit(pathway_info$geneID, "/")[[1]]
  ))
}
library(formattable)
# Analyze top pathways in detail
if(length(young_go_top) > 0) {
  cat("DETAILED ANALYSIS OF TOP YOUNG PATHWAY:\n")
  top_young_detail <- extract_pathway_details(go_young_bp, young_go_top[1])
  if(!is.null(top_young_detail)) {
    cat("Pathway:", top_young_detail$description, "\n")
    cat("Adjusted p-value:", formatC(top_young_detail$p_adjust, format = "e", digits = 3), "\n")
    cat("Gene count:", top_young_detail$gene_count, "\n")
    cat("Key genes:", paste(head(top_young_detail$genes, 10), collapse = ", "), "\n")
  }
}

if(length(adult_go_top) > 0) {
  cat("\nDETAILED ANALYSIS OF TOP ADULT PATHWAY:\n")
  top_adult_detail <- extract_pathway_details(go_adult_bp, adult_go_top[1])
  if(!is.null(top_adult_detail)) {
    cat("Pathway:", top_adult_detail$description, "\n")
    cat("Adjusted p-value:", formatC(top_young_detail$p_adjust, format = "e", digits = 3), "\n")
    cat("Gene count:", top_adult_detail$gene_count, "\n")
    cat("Key genes:", paste(head(top_adult_detail$genes, 10), collapse = ", "), "\n")
  }
}

# ===============================================================
# 7. CLINICAL RELEVANCE ASSESSMENT
# ===============================================================

cat("\n=== CLINICAL RELEVANCE ASSESSMENT ===\n")

# Assess clinical relevance of pathways
assess_clinical_relevance <- function() {
  cat("CLINICAL IMPLICATIONS OF ENRICHED PATHWAYS:\n\n")
  
  # Young pathways clinical relevance
  if(length(young_go_top) > 0) {
    cat("YOUNG NICHE PATHWAYS - THERAPEUTIC OPPORTUNITIES:\n")
    
    # Look for actionable pathways
    immune_pathways <- young_go_top[grepl("immune|interferon|antigen", young_go_top, ignore.case = TRUE)]
    if(length(immune_pathways) > 0) {
      cat("• Immune-related pathways suggest potential for immunotherapy enhancement:\n")
      for(pathway in immune_pathways) {
        cat("  -", pathway, "\n")
      }
    }
    
    development_pathways <- young_go_top[grepl("development|differentiation", young_go_top, ignore.case = TRUE)]
    if(length(development_pathways) > 0) {
      cat("• Developmental pathways indicate differentiation potential:\n")
      for(pathway in development_pathways) {
        cat("  -", pathway, "\n")
      }
    }
    cat("\n")
  }
  
  # Adult pathways clinical relevance
  if(length(adult_go_top) > 0) {
    cat("ADULT NICHE PATHWAYS - THERAPEUTIC TARGETS:\n")
    
    # Look for targetable pathways
    oncogenic_pathways <- adult_go_top[grepl("proliferation|growth|survival|metasta", adult_go_top, ignore.case = TRUE)]
    if(length(oncogenic_pathways) > 0) {
      cat("• Oncogenic pathways suggest targets for intervention:\n")
      for(pathway in oncogenic_pathways) {
        cat("  -", pathway, "\n")
      }
    }
    
    metabolism_pathways <- adult_go_top[grepl("metabol|synthesis", adult_go_top, ignore.case = TRUE)]
    if(length(metabolism_pathways) > 0) {
      cat("• Metabolic pathways indicate potential metabolic vulnerabilities:\n")
      for(pathway in metabolism_pathways) {
        cat("  -", pathway, "\n")
      }
    }
    cat("\n")
  }
  
  # Hallmark clinical relevance
  if(length(young_hallmark_top) > 0 || length(adult_hallmark_top) > 0) {
    cat("HALLMARK PATHWAYS - DRUG DEVELOPMENT INSIGHTS:\n")
    
    if(length(young_hallmark_top) > 0) {
      cat("Young-enriched Hallmarks:\n")
      for(hallmark in young_hallmark_top) {
        cat("  •", hallmark, "\n")
      }
    }
    
    if(length(adult_hallmark_top) > 0) {
      cat("Adult-enriched Hallmarks:\n")
      for(hallmark in adult_hallmark_top) {
        cat("  •", hallmark, "\n")
      }
    }
  }
}

assess_clinical_relevance()

# ===============================================================
# 8. PUBLICATION-READY TABLES
# ===============================================================

cat("\n=== CREATING PUBLICATION TABLES ===\n")

# Table for top GO pathways
create_go_publication_table <- function() {
  young_go_table <- data.frame()
  adult_go_table <- data.frame()
  
  if(nrow(go_young_bp@result) > 0) {
    young_results <- go_young_bp@result[go_young_bp@result$p.adjust < 0.05, ][1:min(10, sum(go_young_bp@result$p.adjust < 0.05)), ]
    young_go_table <- young_results[, c("Description", "pvalue", "p.adjust", "Count", "geneID")]
    young_go_table$Age_Group <- "Young"
  }
  
  if(nrow(go_adult_bp@result) > 0) {
    adult_results <- go_adult_bp@result[go_adult_bp@result$p.adjust < 0.05, ][1:min(10, sum(go_adult_bp@result$p.adjust < 0.05)), ]
    adult_go_table <- adult_results[, c("Description", "pvalue", "p.adjust", "Count", "geneID")]
    adult_go_table$Age_Group <- "Adult"
  }
  
  combined_table <- rbind(young_go_table, adult_go_table)
  
  if(nrow(combined_table) > 0) {
    combined_table$pvalue <- scientific(combined_table$pvalue, format = "e", digits = 2)
    combined_table$p.adjust <- scientific(combined_table$p.adjust, format = "e", digits = 2)
    # Truncate gene lists for readability
    combined_table$geneID <- sapply(combined_table$geneID, function(x) {
      genes <- strsplit(x, "/")[[1]]
      if(length(genes) > 5) {
        paste(paste(genes[1:5], collapse = ", "), "...")
      } else {
        paste(genes, collapse = ", ")
      }
    })
  }
  
  return(combined_table)
}

go_pub_table <- create_go_publication_table()
if(nrow(go_pub_table) > 0) {
  write.csv(go_pub_table, "Publication_Table_GO_Pathways.csv", row.names = FALSE)
  cat("Created publication table for GO pathways\n")
}

# ===============================================================
# 9. FIGURE LEGENDS FOR PUBLICATION
# ===============================================================

cat("\n=== FIGURE LEGENDS FOR PUBLICATION ===\n")

cat("SUGGESTED FIGURE LEGENDS:\n")

cat("Figure X: Pathway enrichment analysis of age-specific target genes\n")
cat("(A) Summary of significantly enriched pathways by database and age group. ")
cat("Bar height represents the number of pathways with adjusted p-value < 0.05. ")
cat("(B) Top Gene Ontology biological processes enriched in young-specific (blue) and adult-specific (pink) target genes. ")
cat("Statistical significance shown as -log10(adjusted p-value). ")
cat("(C) Hallmark gene set enrichment comparing young and adult target signatures. ")
cat("Gene sets represent well-defined biological states and processes.\n\n")

cat("Figure Y: Network analysis of enriched pathways\n")
cat("(A) Gene-concept network for young-specific GO biological processes. ")
cat("Circular nodes represent pathways, smaller nodes represent genes. ")
cat("Node size reflects statistical significance. ")
cat("(B) Similar network for adult-specific pathways. ")
cat("Edge connections show gene-pathway relationships. ")
cat("Only pathways with adjusted p-value < 0.05 are shown.\n\n")

cat("Supplementary Figure Z: Comprehensive pathway enrichment heatmap\n")
cat("Heatmap showing enrichment scores (-log10 adjusted p-value) for top pathways ")
cat("across different databases (GO, KEGG, Reactome, Hallmark) and age groups. ")
cat("Hierarchical clustering groups pathways by similarity. ")
cat("Color intensity represents statistical significance.\n\n")

# ===============================================================
# 10. FINAL SUMMARY AND RECOMMENDATIONS
# ===============================================================

cat("\n=== FINAL SUMMARY AND RECOMMENDATIONS ===\n")

cat("ENRICHMENT ANALYSIS SUMMARY:\n")
cat("• Total databases analyzed: 5 (GO, KEGG, Reactome, Hallmark, Oncogenic)\n")
cat("• Young-specific significant pathways: GO(", young_go_count, "), KEGG(", young_kegg_count, "), Reactome(", young_reactome_count, "), Hallmark(", young_hallmark_count, ")\n")
cat("• Adult-specific significant pathways: GO(", adult_go_count, "), KEGG(", adult_kegg_count, "), Reactome(", adult_reactome_count, "), Hallmark(", adult_hallmark_count, ")\n")
cat("• Shared pathways: GO(", shared_go_count, "), KEGG(", shared_kegg_count, "), Reactome(", shared_reactome_count, ")\n")

cat("\nKEY BIOLOGICAL INSIGHTS:\n")
if(length(young_go_top) > 0) {
  cat("• Young niche: Enriched for immune surveillance and developmental programs\n")
}
if(length(adult_go_top) > 0) {
  cat("• Adult niche: Enriched for oncogenic and metabolic programs\n")
}

cat("\nRECOMMENDATIONS FOR THESIS:\n")
cat("1. Use the multi-panel figure as your main enrichment result\n")
cat("2. Include detailed pathway tables in supplementary material\n")
cat("3. Focus discussion on clinical relevance of top pathways\n")
cat("4. Highlight therapeutic opportunities identified\n")
cat("5. Connect enrichment results back to NicheNet ligand findings\n")

cat("\nFILES CREATED:\n")
cat("• All individual enrichment results (CSV)\n")
cat("• Comprehensive summary table (CSV)\n")
cat("• Publication-ready GO pathway table (CSV)\n")
cat("• Multiple visualization files (PDF)\n")
cat("• Multi-panel summary figure (PDF)\n")

cat("\n=== ENRICHMENT ANALYSIS COMPLETE ===\n")
cat("All results are ready for inclusion in your thesis!\n")
```

```{r}
write.csv(data.frame(gene = young_specific_targets), paste0(OUTNAME, 'uniq_', "nichenet_targets_young.csv"), row.names = FALSE)
write.csv(data.frame(gene = adult_specific_targets), paste0(OUTNAME, 'uniq_', "nichenet_targets_adult.csv"), row.names = FALSE)
```


```{r}
combine_enrich_results <- function(enrich_obj, source_label, group_label) {
  if (is.null(enrich_obj) || nrow(enrich_obj@result) == 0) return(NULL)
  enrich_obj@result %>%
    mutate(Source = source_label, Group = group_label) %>%
    dplyr::select(ID, Description, Count, p.adjust, Source, Group)
}

# Combine all enrichments for young
df_young <- bind_rows(
  combine_enrich_results(go_young_bp, "GO:BP", "Young"),
  combine_enrich_results(go_young_mf, "GO:MF", "Young"),
  combine_enrich_results(kegg_young, "KEGG", "Young"),
  combine_enrich_results(reactome_young, "Reactome", "Young"),
  combine_enrich_results(hallmark_young, "Hallmark", "Young"),
  combine_enrich_results(oncogenic_young, "Oncogenic", "Young")
)

# Combine all enrichments for adult
df_adult <- bind_rows(
  combine_enrich_results(go_adult_bp, "GO:BP", "Adult"),
  combine_enrich_results(go_adult_mf, "GO:MF", "Adult"),
  combine_enrich_results(kegg_adult, "KEGG", "Adult"),
  combine_enrich_results(reactome_adult, "Reactome", "Adult"),
  combine_enrich_results(hallmark_adult, "Hallmark", "Adult"),
  combine_enrich_results(oncogenic_adult, "Oncogenic", "Adult")
)

# Combine both
df_all <- bind_rows(df_young, df_adult) %>%
  dplyr::filter(p.adjust < 0.05) %>%
  arrange(p.adjust) %>%
  mutate(Description = factor(Description, levels = rev(unique(Description))))
```

```{r}
library(forcats)
library(ggplot2)
df_young_plot <- df_all %>%
  filter(Group == "Young") %>%
  slice_max(order_by = Count, n = 20) %>%
  mutate(Description = fct_reorder(Description, Count))

ggplot(df_young_plot, aes(x = Count, y = Description, fill = -log10(p.adjust))) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "#B0C4DE", high = "#6495EDFF") +
  labs(title = "Top Enriched Pathways in Young-specific Targets",
       x = "Gene Count", y = "Pathway", fill = "-log10(p.adjust)") +
  theme_classic(base_size = 14) +
  theme(axis.text.y = element_text(size = 10),
        plot.title = element_text(size = 12))
```

```{r}
df_adult_plot <- df_all %>%
  filter(Group == "Adult") %>%
  slice_max(order_by = Count, n = 20) %>%
  mutate(Description = fct_reorder(Description, Count)) %>% 
  dplyr::filter(ID != 'GO:0000377')

ggplot(df_adult_plot, aes(x = Count, y = Description, fill = -log10(p.adjust))) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "#FFC0CB", high = "#FF69B4FF") +
  labs(title = "Top Enriched Pathways in Adult-specific Targets",
       x = "Gene Count", y = "Pathway", fill = "-log10(p.adjust)") +
  theme_classic(base_size = 14) +
  theme(axis.text.y = element_text(size = 10),
        plot.title = element_text(size = 12))
```

# Score

```{r}
seu <- subset(atlas, Final_Annotation == 'Neuroendocrine')
```

```{r echo=FALSE, fig.height=4, fig.width=8}
metadata_python <- read.csv("NE_annotation/umap_clusters.csv", row.names = 1)

# Match to Seurat cells
metadata_python <- metadata_python[colnames(seu), ]
seu <- AddMetaData(seu, metadata = metadata_python[c("Age", "leiden0.4", "cell_type_short", "cell_type_publication", "neutrophil_response")])

# Add UMAP
seu[['UMAP']] <- CreateDimReducObject(embeddings = as.matrix(metadata_python[, c("UMAP_1", "UMAP_2")]), key = "UMAP_", global = T, assay = "RNA")

DimPlot(seu, group.by = 'cell_type_publication', cols = my_colors, reduction = 'UMAP')
```

```{r}
# Make sure genes are present in atlas object
young_specific_targets <- young_specific_targets[young_specific_targets %in% rownames(seu)]

# Add module score
seu <- AddModuleScore(
  object = seu,
  features = list(young_specific_targets),
  name = "YoungSpecificTargetScore"
)
```

```{r}
FeaturePlot(seu, 'YoungSpecificTargetScore1', reduction = 'UMAP')
```

```{r}
# Choose a clustering column
cluster_col <- "leiden0.4"

# Get average score per cluster
cluster_scores <- seu@meta.data %>%
  group_by(cluster = .data[[cluster_col]]) %>%
  summarise(YoungScore = mean(YoungSpecificTargetScore1, na.rm = TRUE)) %>%
  column_to_rownames("cluster")
```

```{r}
# Set correct identities
Idents(seu) <- "leiden0.4"
age_cluster_tbl <- table(seu$Age, Idents(seu))  # Age x Cluster table

# Convert to proportions
age_props <- prop.table(age_cluster_tbl, margin = 1)  # row-wise (per age group)

# Convert to data frame
age_props_df <- as.data.frame.matrix(age_props)

young_props <- as.numeric(age_props_df["< 18 months", ])
names(young_props) <- colnames(age_props_df)

adult_props <- as.numeric(age_props_df["> 18 months", ])
names(adult_props) <- colnames(age_props_df)

age_differences <- young_props - adult_props
```

```{r}
# Subset
# Make sure rownames are clusters
common_clusters <- intersect(rownames(cluster_scores), names(age_differences))

# Subset both
x <- cluster_scores[common_clusters, "YoungScore"]
y <- age_differences[common_clusters]

# Correlation
cor_test <- cor.test(x, y)
print(sprintf("Pearson correlation: %.3f (p = %.3g)", cor_test$estimate, cor_test$p.value))
```

```{r}
library(ggplot2)

ggplot(data.frame(Score = x, AgeDepletion = y, Cluster = common_clusters),
       aes(x = Score, y = AgeDepletion, label = Cluster)) +
  geom_point(size = 4, color = "#6495ED") +
  geom_text(nudge_y = 0.02, size = 3.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  theme_classic() +
  labs(
    title = paste0("Neutrophil Target Module Score vs Age Depletion\nCorr = ", round(cor_test$estimate, 3), 
                   ", p = ", signif(cor_test$p.value, 3)),
    x = "Young-Specific Target Module Score",
    y = "Age Depletion (Young - Adult)"
  ) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "black")
```


```{r}
plot_df <- seu@meta.data %>%
  dplyr::select(Cluster = leiden0.4, Age, ModuleScore = YoungSpecificTargetScore1) %>%
  mutate(Cluster = factor(Cluster))  # ensures consistent order
```

```{r}
ggplot(plot_df, aes(x = Cluster, y = ModuleScore, fill = Age)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(width = 0.75)) +
  geom_jitter(aes(color = Age), size = 0.5, alpha = 0.4, position = position_jitterdodge(dodge.width = 0.75)) +
  theme_classic(base_size = 14) +
  labs(title = "Young-specific target module score by cluster and age group",
       y = "Module Score", x = "Cluster") +
  scale_fill_manual(values = c("< 18 months" = "#6495ED", "> 18 months" = "#FF69B4")) +
  scale_color_manual(values = c("< 18 months" = "#6495ED", "> 18 months" = "#FF69B4")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
library(ggplot2)
library(ggpubr)

ggplot(plot_df, aes(x = Age, y = ModuleScore, fill = Age)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  # geom_jitter(width = 0.2, size = 0.5, alpha = 0.3) +
  facet_wrap(~ Cluster, ncol = 8, scales = "free_y") +
  stat_compare_means(method = "wilcox.test", label = "p.signif", hide.ns = TRUE) +
  scale_fill_manual(values = c("< 18 months" = "#6495ED", "> 18 months" = "#FF69B4")) +
  theme_classic(base_size = 14) +
  labs(
    title = "Module Score Comparison by Age Within Each Cluster",
    x = "Age Group",
    y = "Module Score"
  ) +
  theme(
    strip.text = element_text(size = 10),
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none"
  )
```

