---
title: "nichenet_ATLAS"
author: "Kate Petrenko"
date: "2025-04-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# devtools::install_github("saeyslab/nichenetr")
library(nichenetr)
library(RColorBrewer)
library(tidyverse)
library(Seurat) 
library(qs)
library(EnhancedVolcano)
```

```{r global variables, include=FALSE}
set.seed(100)
DATADIR = '~/neuroblastoma/data/'
WORKDIR = '~/neuroblastoma/'
PREPRDATADIR = '~/neuroblastoma/preprocessing_data/'
OUTNAME = 'VALID_'
my_colors <- c("#6495EDFF", "#FF69B4FF", "#BA55D3FF", "#A4BEF3", "#F08080FF", "#32CD32FF",  
    "#9ACD32FF", "#4682B4FF", "#DDA0DDFF", "#FFA07AFF", "#8FBC8BFF",
    "#40E0D0FF", "#F0E68CFF", "#5F9EA0FF", "#D2B48CFF",  
    "#FFDAB9FF", "#87CEEBFF", "#B4A0E5", "#5BC0BE", "#773344")
```

# Data

```{r include=FALSE}
atlas <- qread(paste0(PREPRDATADIR, 'ATLAS_object.qs'))
```

```{r eval=FALSE, include=FALSE}
sample_metadata <- read.csv(paste0(DATADIR, 'samples_metadata_v5.csv'))
sample_metadata[sample_metadata==""] <- NA

rownames(sample_metadata) <- sample_metadata$Sample_dataset



clean_sample_id <- function(id) {
  # First remove _rep1 and _rep2
  id <- gsub("_rep[12]$", "", id)
  
  # Handle the special bonine cases
  id <- gsub("_sn$", "", id)      # Remove _sn suffix
  id <- gsub("_[12]$", "", id)    # Remove _1 or _2 suffix
  
  return(id)
}

atlas$less18M <- sample_metadata[clean_sample_id(atlas$SampleID),]$less18M
atlas$Sex <- sample_metadata[clean_sample_id(atlas$SampleID),]$Sex
atlas$MYCN_status <- sample_metadata[clean_sample_id(atlas$SampleID),]$MYCN
atlas$method <- sample_metadata[clean_sample_id(atlas$SampleID),]$Method
```

```{r echo=FALSE}
metadata_python <- read.csv("query_metadata_with_umap.csv", row.names = 1)

# Match to Seurat cells
metadata_python <- metadata_python[colnames(atlas), ]
atlas <- AddMetaData(atlas, metadata = metadata_python[c("Age", "celltypist_cell_label_fine", "Final_Annotation")])

# Add UMAP
atlas[['UMAP']] <- CreateDimReducObject(embeddings = as.matrix(metadata_python[, c("UMAP_1", "UMAP_2")]), key = "UMAP_", global = T, assay = "RNA")

DimPlot(atlas, group.by = 'Final_Annotation', cols = my_colors, reduction = 'UMAP')
```

```{r}
atlas <- subset(atlas, Study != 'dong')
```

```{r include=FALSE}
lr_network <- readRDS(url("https://zenodo.org/record/7074291/files/lr_network_human_21122021.rds"))
# ligand_target_matrix <- readRDS(url("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final.rds"))
# options(timeout = 600)
# download.file("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final.rds", 
              # destfile = "ligand_target_matrix.rds", mode = "wb")
ligand_target_matrix <- readRDS("ligand_target_matrix.rds")
# weighted_networks <- readRDS(url("https://zenodo.org/record/7074291/files/weighted_networks_nsga2r_final.rds"))
```

```{r eval=FALSE, include=FALSE}
# ligand_target_matrix = readRDS(url("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final_mouse.rds"))
ligand_target_matrix[1:5,1:5] # target genes in rows, ligands in columns
```

```{r include=FALSE}
# lr_network = readRDS(url("https://zenodo.org/record/7074291/files/lr_network_mouse_21122021.rds"))
lr_network = lr_network %>% dplyr::rename(ligand = from, receptor = to) %>% distinct(ligand, receptor)
```

```{r include=FALSE}
atlas$niche_label <- ifelse(atlas$Age == "< 18 months", "young_niche", "old_niche")
Idents(atlas) <- "niche_label"  # cell types

atlas$niche_label <- ifelse(atlas$Age == "< 18 months", "young", "adult")
atlas$niche_label <- paste0(atlas$Final_Annotation, '_', atlas$niche_label)
Idents(atlas) <- "niche_label"  # cell types
```

```{r include=FALSE}
niches = list(
  "young_niche" = list(
    "sender" = c("Neutrophils_young"
                 # , "Monocytes_young"
                 # , "Macrophages_young"
                 ),
    "receiver" = c("Neuroendocrine_young")
  ),
  "old_niche" = list(
    "sender" = c("Neutrophils_adult"
                 # , "Monocytes_adult"
                 # , "Macrophages_adult"
                 ),
    "receiver" = c("Neuroendocrine_adult")
  )
)
```

```{r include=FALSE}
assay_oi <- 'RNA'
DE_sender = calculate_niche_de(seurat_obj = atlas %>% subset(features = lr_network$ligand %>% intersect(rownames(atlas))), niches = niches, type = "sender", assay_oi = assay_oi) # only ligands important for sender cell types

DE_sender %>% 
  dplyr::filter(p_val_adj < 0.05 & sender == 'Neutrophils_young')
DE_sender %>% 
  dplyr::filter(p_val_adj < 0.05 & sender == 'Monocytes_young')
DE_sender %>% 
  dplyr::filter(p_val_adj < 0.05 & sender == 'Macrophages_young')
```

```{r include=FALSE}
DE_receiver = calculate_niche_de(seurat_obj = atlas %>% subset(features = lr_network$receptor %>% unique()), niches = niches, type = "receiver", assay_oi = assay_oi) # only receptors now, later on: DE analysis to find targets

DE_receiver %>% 
  dplyr::filter(p_val_adj < 0.05 & receiver == 'Neuroendocrine_young') %>% 
  dplyr::filter(pct.1 > 0.25 | pct.2 > 0.25)
```

```{r include=FALSE}
DE_sender = DE_sender %>% mutate(avg_log2FC = ifelse(avg_log2FC == Inf, max(avg_log2FC[is.finite(avg_log2FC)]), ifelse(avg_log2FC == -Inf, min(avg_log2FC[is.finite(avg_log2FC)]), avg_log2FC)))
DE_receiver = DE_receiver %>% mutate(avg_log2FC = ifelse(avg_log2FC == Inf, max(avg_log2FC[is.finite(avg_log2FC)]), ifelse(avg_log2FC == -Inf, min(avg_log2FC[is.finite(avg_log2FC)]), avg_log2FC)))
```

```{r include=FALSE}
expression_pct = 0.15
DE_sender_processed = process_niche_de(DE_table = DE_sender, niches = niches, expression_pct = expression_pct, type = "sender")
DE_receiver_processed = process_niche_de(DE_table = DE_receiver, niches = niches, expression_pct = expression_pct, type = "receiver")
```

```{r include=FALSE}
specificity_score_LR_pairs = "min_lfc"
DE_sender_receiver = combine_sender_receiver_de(DE_sender_processed, DE_receiver_processed, lr_network, specificity_score = specificity_score_LR_pairs)
```

```{r include=FALSE}
lfc_cutoff = 0.4 
target_score = 0.4
specificity_score_targets = "min_lfc"

DE_receiver_targets = calculate_niche_de_targets(seurat_obj = atlas, niches = niches, lfc_cutoff = lfc_cutoff, expression_pct = expression_pct, assay_oi = assay_oi) 

DE_receiver_processed_targets = process_receiver_target_de(DE_receiver_targets = DE_receiver_targets, niches = niches, expression_pct = expression_pct, specificity_score = specificity_score_targets)
  
background = DE_receiver_processed_targets  %>% pull(target) %>% unique()
geneset_NEyoung = DE_receiver_processed_targets %>% filter(receiver == niches$young_niche$receiver & target_score >= target_score & target_significant == 1 & target_present == 1) %>% pull(target) %>% unique()
geneset_NEadult = DE_receiver_processed_targets %>% filter(receiver == niches$old_niche$receiver & target_score >= target_score & target_significant == 1 & target_present == 1) %>% pull(target) %>% unique()

# Good idea to check which genes will be left out of the ligand activity analysis (=when not present in the rownames of the ligand-target matrix).
# If many genes are left out, this might point to some issue in the gene naming (eg gene aliases and old gene symbols, bad human-mouse mapping)
geneset_NEyoung %>% setdiff(rownames(ligand_target_matrix))
geneset_NEadult %>% setdiff(rownames(ligand_target_matrix))


length(geneset_NEyoung)
length(geneset_NEadult)
```

```{r}
DE_receiver_processed_targets %>% 
  dplyr::filter(target_score > 0.5)
```

```{r include=FALSE}
top_n_target = 250

niche_geneset_list = list(
    "young_niche" = list(
      "receiver" = niches$young_niche$receiver,
      "geneset" = geneset_NEyoung,
      "background" = background),
    "old_niche" = list(
      "receiver" = niches$old_niche$receiver,
      "geneset" = geneset_NEadult ,
      "background" = background)
  )
  
ligand_activities_targets = get_ligand_activities_targets(niche_geneset_list = niche_geneset_list, ligand_target_matrix = ligand_target_matrix, top_n_target = top_n_target)  
```

```{r include=FALSE}
features_oi = union(lr_network$ligand, lr_network$receptor) %>% union(ligand_activities_targets$target) %>% setdiff(NA)
  
dotplot = suppressWarnings(Seurat::DotPlot(atlas %>% subset(idents = niches %>% unlist() %>% unique()), features = features_oi, assay = assay_oi))
exprs_tbl = dotplot$data %>% as_tibble()
exprs_tbl = exprs_tbl %>% rename(celltype = id, gene = features.plot, expression = avg.exp, expression_scaled = avg.exp.scaled, fraction = pct.exp) %>%
    mutate(fraction = fraction/100) %>% as_tibble() %>% select(celltype, gene, expression, expression_scaled, fraction) %>% distinct() %>% arrange(gene) %>% mutate(gene = as.character(gene))
  
exprs_tbl_ligand = exprs_tbl %>% filter(gene %in% lr_network$ligand) %>% rename(sender = celltype, ligand = gene, ligand_expression = expression, ligand_expression_scaled = expression_scaled, ligand_fraction = fraction) 
exprs_tbl_receptor = exprs_tbl %>% filter(gene %in% lr_network$receptor) %>% rename(receiver = celltype, receptor = gene, receptor_expression = expression, receptor_expression_scaled = expression_scaled, receptor_fraction = fraction)
exprs_tbl_target = exprs_tbl %>% filter(gene %in% ligand_activities_targets$target) %>% rename(receiver = celltype, target = gene, target_expression = expression, target_expression_scaled = expression_scaled, target_fraction = fraction)
```

```{r include=FALSE}
exprs_tbl_ligand = exprs_tbl_ligand %>%  mutate(scaled_ligand_expression_scaled = scale_quantile_adapted(ligand_expression_scaled)) %>% mutate(ligand_fraction_adapted = ligand_fraction) %>% mutate_cond(ligand_fraction >= expression_pct, ligand_fraction_adapted = expression_pct)  %>% mutate(scaled_ligand_fraction_adapted = scale_quantile_adapted(ligand_fraction_adapted))

exprs_tbl_receptor = exprs_tbl_receptor %>% mutate(scaled_receptor_expression_scaled = scale_quantile_adapted(receptor_expression_scaled))  %>% mutate(receptor_fraction_adapted = receptor_fraction) %>% mutate_cond(receptor_fraction >= expression_pct, receptor_fraction_adapted = expression_pct)  %>% mutate(scaled_receptor_fraction_adapted = scale_quantile_adapted(receptor_fraction_adapted))
```


```{r include=FALSE}
include_spatial_info_sender = FALSE # if not spatial info to include: put this to false 
include_spatial_info_receiver = FALSE # if spatial info to include: put this to true 
```


```{r include=FALSE}
# this is how this should be defined if you don't have spatial info
# mock spatial info
if(include_spatial_info_sender == FALSE & include_spatial_info_receiver == FALSE){
    spatial_info = tibble(celltype_region_oi = NA, celltype_other_region = NA) %>% mutate(niche =  niches %>% names() %>% head(1), celltype_type = "sender")
} 
```

```{r include=FALSE}
if(include_spatial_info_sender == TRUE){
  sender_spatial_DE = calculate_spatial_DE(seurat_obj = seurat_obj %>% subset(features = lr_network$ligand %>% unique()), spatial_info = spatial_info %>% filter(celltype_type == "sender"), assay_oi = assay_oi)
  sender_spatial_DE_processed = process_spatial_de(DE_table = sender_spatial_DE, type = "sender", lr_network = lr_network, expression_pct = expression_pct, specificity_score = specificity_score_spatial)

  # add a neutral spatial score for sender celltypes in which the spatial is not known / not of importance
  sender_spatial_DE_others = get_non_spatial_de(niches = niches, spatial_info = spatial_info, type = "sender", lr_network = lr_network)
  sender_spatial_DE_processed = sender_spatial_DE_processed %>% bind_rows(sender_spatial_DE_others)

  sender_spatial_DE_processed = sender_spatial_DE_processed %>% mutate(scaled_ligand_score_spatial = scale_quantile_adapted(ligand_score_spatial))

} else {
  # # add a neutral spatial score for all sender celltypes (for none of them, spatial is relevant in this case)
  sender_spatial_DE_processed = get_non_spatial_de(niches = niches, spatial_info = spatial_info, type = "sender", lr_network = lr_network)
  sender_spatial_DE_processed = sender_spatial_DE_processed %>% mutate(scaled_ligand_score_spatial = scale_quantile_adapted(ligand_score_spatial))  

}
## [1] "Calculate Spatial DE between: LSECs_portal and LSECs_central"
## [1] "Calculate Spatial DE between: Hepatocytes_portal and Hepatocytes_central"
## [1] "Calculate Spatial DE between: Stellate cells_portal and Stellate cells_central"
```

```{r include=FALSE}
if(include_spatial_info_receiver == TRUE){
  receiver_spatial_DE = calculate_spatial_DE(seurat_obj = seurat_obj %>% subset(features = lr_network$receptor %>% unique()), spatial_info = spatial_info %>% filter(celltype_type == "receiver"), assay_oi = assay_oi)
  receiver_spatial_DE_processed = process_spatial_de(DE_table = receiver_spatial_DE, type = "receiver", lr_network = lr_network, expression_pct = expression_pct, specificity_score = specificity_score_spatial)

  # add a neutral spatial score for receiver celltypes in which the spatial is not known / not of importance
  receiver_spatial_DE_others = get_non_spatial_de(niches = niches, spatial_info = spatial_info, type = "receiver", lr_network = lr_network)
  receiver_spatial_DE_processed = receiver_spatial_DE_processed %>% bind_rows(receiver_spatial_DE_others)

  receiver_spatial_DE_processed = receiver_spatial_DE_processed %>% mutate(scaled_receptor_score_spatial = scale_quantile_adapted(receptor_score_spatial))

} else {
    # # add a neutral spatial score for all receiver celltypes (for none of them, spatial is relevant in this case)
  receiver_spatial_DE_processed = get_non_spatial_de(niches = niches, spatial_info = spatial_info, type = "receiver", lr_network = lr_network)
  receiver_spatial_DE_processed = receiver_spatial_DE_processed %>% mutate(scaled_receptor_score_spatial = scale_quantile_adapted(receptor_score_spatial))
}
```

# Analysis

```{r include=FALSE}
exprs_sender_receiver = lr_network %>% 
  inner_join(exprs_tbl_ligand, by = c("ligand")) %>% 
  inner_join(exprs_tbl_receptor, by = c("receptor")) %>% inner_join(DE_sender_receiver %>% distinct(niche, sender, receiver))
  
ligand_scaled_receptor_expression_fraction_df = exprs_sender_receiver %>% group_by(ligand, receiver) %>% mutate(rank_receptor_expression = dense_rank(receptor_expression), rank_receptor_fraction  = dense_rank(receptor_fraction)) %>% mutate(ligand_scaled_receptor_expression_fraction = 0.5*( (rank_receptor_fraction / max(rank_receptor_fraction)) + ((rank_receptor_expression / max(rank_receptor_expression))) ) )  %>% distinct(ligand, receptor, receiver, ligand_scaled_receptor_expression_fraction) %>% distinct() %>% ungroup() 
```

```{r include=FALSE}
prioritizing_weights = c("scaled_ligand_score" = 5,
                         "scaled_ligand_expression_scaled" = 1,
                         "ligand_fraction" = 1,
                         "scaled_ligand_score_spatial" = 0, 
                         "scaled_receptor_score" = 0.5,
                         "scaled_receptor_expression_scaled" = 0.5,
                          "receptor_fraction" = 1, 
                         "ligand_scaled_receptor_expression_fraction" = 1,
                         "scaled_receptor_score_spatial" = 0,
                         "scaled_activity" = 0,
                         "scaled_activity_normalized" = 1)
```

```{r include=FALSE}
output = list(DE_sender_receiver = DE_sender_receiver, ligand_scaled_receptor_expression_fraction_df = ligand_scaled_receptor_expression_fraction_df, sender_spatial_DE_processed = sender_spatial_DE_processed, receiver_spatial_DE_processed = receiver_spatial_DE_processed,
         ligand_activities_targets = ligand_activities_targets, DE_receiver_processed_targets = DE_receiver_processed_targets, exprs_tbl_ligand = exprs_tbl_ligand,  exprs_tbl_receptor = exprs_tbl_receptor, exprs_tbl_target = exprs_tbl_target)
prioritization_tables = get_prioritization_tables(output, prioritizing_weights)
```

```{r include=FALSE}


prioritization_tables$prioritization_tbl_ligand_receptor %>% filter(receiver == niches[[1]]$receiver) %>% head(10)

prioritization_tables$prioritization_tbl_ligand_target %>% filter(receiver == niches[[1]]$receiver) %>% head(10)


prioritization_tables$prioritization_tbl_ligand_receptor %>% filter(receiver == niches[[2]]$receiver) %>% head(10)

prioritization_tables$prioritization_tbl_ligand_target %>% filter(receiver == niches[[2]]$receiver) %>% head(10)


prioritization_tables$prioritization_tbl_ligand_receptor = prioritization_tables$prioritization_tbl_ligand_receptor %>% mutate(receiver = factor(receiver, levels = c("Neuroendocrine_young","Neuroendocrine_adult")), niche = factor(niche, levels = c("young_niche","old_niche"))) 
prioritization_tables$prioritization_tbl_ligand_target = prioritization_tables$prioritization_tbl_ligand_target %>% mutate(receiver = factor(receiver, levels = c("Neuroendocrine_young","Neuroendocrine_adult")), niche = factor(niche, levels = c("young_niche","old_niche"))) 
```

# Visualization

## Young

```{r include=FALSE}
top_ligand_niche_df = prioritization_tables$prioritization_tbl_ligand_receptor %>% select(niche, sender, receiver, ligand, receptor, prioritization_score) %>% group_by(ligand) %>% top_n(1, prioritization_score) %>% ungroup() %>% select(ligand, receptor, niche) %>% rename(top_niche = niche)
top_ligand_receptor_niche_df = prioritization_tables$prioritization_tbl_ligand_receptor %>% select(niche, sender, receiver, ligand, receptor, prioritization_score) %>% group_by(ligand, receptor) %>% top_n(1, prioritization_score) %>% ungroup() %>% select(ligand, receptor, niche) %>% rename(top_niche = niche)

ligand_prioritized_tbl_oi = prioritization_tables$prioritization_tbl_ligand_receptor %>% select(niche, sender, receiver, ligand, prioritization_score) %>% group_by(ligand, niche) %>% top_n(1, prioritization_score) %>% ungroup() %>% distinct() %>% inner_join(top_ligand_niche_df) %>% filter(niche == top_niche) %>% group_by(niche) %>% top_n(50, prioritization_score) %>% ungroup() # get the top50 ligands per niche
```

```{r include=FALSE}
receiver_oi = "Neuroendocrine_young" 

filtered_ligands = ligand_prioritized_tbl_oi %>% filter(receiver == receiver_oi) %>% pull(ligand) %>% unique()

prioritized_tbl_oi = prioritization_tables$prioritization_tbl_ligand_receptor %>% filter(ligand %in% filtered_ligands) %>% select(niche, sender, receiver, ligand,  receptor, ligand_receptor, prioritization_score) %>% distinct() %>% inner_join(top_ligand_receptor_niche_df) %>% group_by(ligand) %>% filter(receiver == receiver_oi) %>% top_n(2, prioritization_score) %>% ungroup() 
```

```{r echo=FALSE, fig.height=12, fig.width=8, message=FALSE, warning=FALSE}
lfc_plot = make_ligand_receptor_lfc_plot(receiver_oi, prioritized_tbl_oi, prioritization_tables$prioritization_tbl_ligand_receptor, plot_legend = FALSE, heights = NULL, widths = NULL)
lfc_plot 
```

low expression of IFITM1 was associated with a poor prognosis in PRAD (Prostate adenocarcinoma). 

In a recent study published in Cell, Gungabeesoon and colleagues1 demonstrated that a specific subpopulation of neutrophils, characterized by a distinct Sell(hi) (CD62L(hi)) phenotype with an interferon gene signature, acutely accumulates in tumors during successful immunotherapy and is associated with better treatment outcome.

we conclude that OSM produced by recruited neutrophils tunes early innate immune signaling pathways, improving pneumonia outcomes.




```{r fig.height=10, fig.width=34, include=FALSE}
exprs_activity_target_plot = make_ligand_activity_target_exprs_plot(receiver_oi, prioritized_tbl_oi,  prioritization_tables$prioritization_tbl_ligand_receptor,  prioritization_tables$prioritization_tbl_ligand_target, output$exprs_tbl_ligand,  output$exprs_tbl_target, lfc_cutoff, ligand_target_matrix, plot_legend = FALSE, heights = NULL, widths = NULL)
exprs_activity_target_plot$combined_plot
```

```{r echo=FALSE, fig.height=10, fig.width=34, message=FALSE, warning=FALSE}
exprs_activity_target_plot = make_ligand_activity_target_exprs_plot(receiver_oi, prioritized_tbl_oi,  prioritization_tables$prioritization_tbl_ligand_receptor, 
                                                                    prioritization_tables$prioritization_tbl_ligand_target, output$exprs_tbl_ligand,  output$exprs_tbl_target, lfc_cutoff, ligand_target_matrix,
                                                                    scaled_ligand_activity_limits = "IQR", plot_legend = FALSE, heights = NULL, widths = NULL)
exprs_activity_target_plot$combined_plot
```

```{r message=FALSE, warning=FALSE, include=FALSE}

filtered_ligands = ligand_prioritized_tbl_oi %>% filter(receiver == receiver_oi) %>% top_n(15, prioritization_score) %>% pull(ligand) %>% unique() %>% c('S100A8', "S100A9")

prioritized_tbl_oi = prioritization_tables$prioritization_tbl_ligand_receptor %>% filter(ligand %in% filtered_ligands) %>% select(niche, sender, receiver, ligand,  receptor, ligand_receptor, prioritization_score) %>% distinct() %>% inner_join(top_ligand_receptor_niche_df) %>% group_by(ligand) %>% filter(receiver == receiver_oi) %>% top_n(2, prioritization_score) %>% ungroup() 

colors_sender = brewer.pal(n = prioritized_tbl_oi$sender %>% unique() %>% sort() %>% length(), name = 'Spectral') %>% magrittr::set_names(prioritized_tbl_oi$sender %>% unique() %>% sort())
colors_receiver = c("lavender")  %>% magrittr::set_names(prioritized_tbl_oi$receiver %>% unique() %>% sort())

circos_output = make_circos_lr(prioritized_tbl_oi, colors_sender, colors_receiver, separate_legend = F)
```

```{r fig.height=5, fig.width=10}
circos_output
```

## Adult

```{r include=FALSE}
receiver_oi = "Neuroendocrine_adult" 

filtered_ligands = ligand_prioritized_tbl_oi %>% filter(receiver == receiver_oi) %>% pull(ligand) %>% unique()

prioritized_tbl_oi = prioritization_tables$prioritization_tbl_ligand_receptor %>% filter(ligand %in% filtered_ligands) %>% select(niche, sender, receiver, ligand,  receptor, ligand_receptor, prioritization_score) %>% distinct() %>% inner_join(top_ligand_receptor_niche_df) %>% group_by(ligand) %>% filter(receiver == receiver_oi) %>% top_n(2, prioritization_score) %>% ungroup() 
```

```{r echo=FALSE, fig.height=12, fig.width=8, message=FALSE, warning=FALSE}
lfc_plot = make_ligand_receptor_lfc_plot(receiver_oi, prioritized_tbl_oi, prioritization_tables$prioritization_tbl_ligand_receptor, plot_legend = FALSE, heights = NULL, widths = NULL)
lfc_plot
```

Recent studies have demonstrated that epiregulin (EREG) is upregulated in various cancer types, which contributes to cancer progression by triggering the EGFR signaling pathway.

ADAM17 is contributory to the occurrence and development of cancers

ytokine CXCL12 (also known as stromal-derived factor 1α) and its receptor CXCR4 represent the most promising actionable targets for this strategy. Both are overexpressed in various cancer types, and this aberrant expression strongly promotes proliferation, migration and invasion through multiple signal pathways.


```{r fig.height=10, fig.width=34, include=FALSE}
exprs_activity_target_plot = make_ligand_activity_target_exprs_plot(receiver_oi, prioritized_tbl_oi,  prioritization_tables$prioritization_tbl_ligand_receptor,  prioritization_tables$prioritization_tbl_ligand_target, output$exprs_tbl_ligand,  output$exprs_tbl_target, lfc_cutoff, ligand_target_matrix, plot_legend = FALSE, heights = NULL, widths = NULL)
exprs_activity_target_plot$combined_plot
```

```{r echo=FALSE, fig.height=10, fig.width=34, message=FALSE, warning=FALSE}
exprs_activity_target_plot = make_ligand_activity_target_exprs_plot(receiver_oi, prioritized_tbl_oi,  prioritization_tables$prioritization_tbl_ligand_receptor, 
                                                                    prioritization_tables$prioritization_tbl_ligand_target, output$exprs_tbl_ligand,  output$exprs_tbl_target, lfc_cutoff, ligand_target_matrix,
                                                                    scaled_ligand_activity_limits = "IQR", plot_legend = FALSE, heights = NULL, widths = NULL)
exprs_activity_target_plot$combined_plot
```

```{r message=FALSE, warning=FALSE, include=FALSE}

filtered_ligands = ligand_prioritized_tbl_oi %>% filter(receiver == receiver_oi) %>% top_n(15, prioritization_score) %>% pull(ligand) %>% unique()

prioritized_tbl_oi = prioritization_tables$prioritization_tbl_ligand_receptor %>% filter(ligand %in% filtered_ligands) %>% select(niche, sender, receiver, ligand,  receptor, ligand_receptor, prioritization_score) %>% distinct() %>% inner_join(top_ligand_receptor_niche_df) %>% group_by(ligand) %>% filter(receiver == receiver_oi) %>% top_n(2, prioritization_score) %>% ungroup() 

colors_sender = brewer.pal(n = prioritized_tbl_oi$sender %>% unique() %>% sort() %>% length(), name = 'Spectral') %>% magrittr::set_names(prioritized_tbl_oi$sender %>% unique() %>% sort())
colors_receiver = c("lavender")  %>% magrittr::set_names(prioritized_tbl_oi$receiver %>% unique() %>% sort())

circos_output = make_circos_lr(prioritized_tbl_oi, colors_sender, colors_receiver)
```

```{r echo=FALSE, fig.height=5, fig.width=10}
circos_output
```


# Heatmaps

```{r include=FALSE}
# Human gene names
human_genes <- c(
  "BCL2", "MAPK1", "BAX", "NFKB1", "JUN", "TNF", "IL6", "AP1", "CXCL1", "CCL5", "CCL7", 
  "SLC39A10", "LCN2", "ZC3H12A", "ENPP2", "MMP10", "MMP1", "MMP2", "MMP9", 
  "PIK3CA", "RAC1",
  "CASP1", "CASP3", "CASP4", "CASP5", "CASP7", "CASP8", "CASP9", "CASP10", "CASP14",
  "XAF1", "ULK1", "XIAP", "H2AFX", "TP53BP1", "ATG13", "RB1CC1", "AMBRA1", "BECN1", 
  "PIK3C3", "ATG14", "UVRAG", "ATG5", "ATG12", "ATG16L1", "MAP1LC3A", "ATG7", "ATG3", 
  "ATG10", "RAB7A", "LAMP1", "LAMP2", "STX17", "MTOR", "PRKAA1", "TFEB", "BAK1", 
  "BNIP3", "BBC3", "PMAIP1", "FAS", "MCL1", "BIRC5", "DNM1L", "VDAC1", "MPTP", 
  "TFAM", "ATP5F1A", "TNFRSF1A", "DDIT3", "ATF4", "TLR4", "AGER", "P2RX7", "NLRP3", 
  "CAPN1", "NFE2L2", "CYCS", "GSK3B", "IL18", "AIFM1", "NOX1", "NOX2", "NOX3", "NOX4", "NOX5", 
  "SOD1", "SOD2", "SOD3", "CAT", "GPX1", "GPX4", "TXN", "TXNRD1", "PRDX1", "PRDX2", "PRDX3", 
  "XDH", "AHR", 
  # Fibroblast related
  "VIM", "STAT3", "SMAD2", "SMAD3", "IL33", "IL1B", "ACTA2", "CXCL8", "COL1A1", "PDGFB",
  # Macrophage related
  "ARG1", "P2RX7", "SOCS3", "NLRP3", "MYD88", "NOS2"
)

# Mouse gene names
mouse_genes <- c(
  "Bcl2", "Mapk1", "Bax", "Nfkb1", "Jun", "Tnf", "Il6", "Ap1", "Cxcl1", "Ccl5", "Ccl7", 
  "Slc39a10", "Lcn2", "Zc3h12a", "Enpp2", "Mmp10", "Mmp1", "Mmp2", "Mmp9", 
  "Pik3ca", "Rac1",
  "Casp1", "Casp3", "Casp4", "Casp5", "Casp7", "Casp8", "Casp9", "Casp10", "Casp14",
  "Xaf1", "Ulk1", "Xiap", "H2afx", "Trp53bp1", "Atg13", "Rb1cc1", "Ambra1", "Becn1", 
  "Pik3c3", "Atg14", "Uvrag", "Atg5", "Atg12", "Atg16l1", "Map1lc3a", "Atg7", "Atg3", 
  "Atg10", "Rab7a", "Lamp1", "Lamp2", "Stx17", "Mtor", "Prkaa1", "Tfeb", "Bak1", 
  "Bnip3", "Bbc3", "Pmaip1", "Fas", "Mcl1", "Birc5", "Dnm1l", "Vdac1", "Mptp", 
  "Tfam", "Atp5f1a", "Tnfrsf1a", "Ddit3", "Atf4", "Tlr4", "Ager", "P2rx7", "Nlrp3", 
  "Capn1", "Nfe2l2", "Cycs", "Gsk3b", "Il18", "Aifm1", "Nox1", "Nox2", "Nox3", "Nox4", "Nox5", 
  "Sod1", "Sod2", "Sod3", "Cat", "Gpx1", "Gpx4", "Txn", "Txnrd1", "Prdx1", "Prdx2", "Prdx3", 
  "Xdh", "Ahr", 
  # Fibroblast related
  "Vim", "Stat3", "Smad2", "Smad3", "Il33", "Il1b", "Acta2", "Cxcl8", "Col1a1", "Pdgfb",
  # Macrophage related
  "Arg1", "P2rx7", "Socs3", "Nlrp3", "Myd88", "Nos2"
)
```

```{r include=FALSE}
DEG_sc_mouse <- qread(paste0(PREPRDATADIR, 'DEG_sc_mouse.qs'))
DEG_bulk_human <- qread(paste0(PREPRDATADIR, 'DEG_ATLAS_pseudobulk_norep_results.qs'))
mouse <- qread(paste0(PREPRDATADIR, 'seu_mouse_neutrophils.qs'))
```

```{r echo=FALSE, fig.height=8, fig.width=5}
for (i in unique(DEG_sc_mouse$celltype)) {
highlight_genes <- unique(mouse_genes[mouse_genes %in% (DEG_sc_mouse %>% dplyr::filter(celltype == i & p_val_adj < 0.05 & abs(avg_log2FC) > 0.15) %>% pull(gene))])

# 3. Plot with EnhancedVolcano
print(EnhancedVolcano(
  DEG_sc_mouse %>% dplyr::filter(celltype == i),
  lab = DEG_sc_mouse %>% dplyr::filter(celltype == i) %>% pull(gene),
  selectLab = highlight_genes,
  x = 'avg_log2FC',
  y = 'p_val_adj',
  col = c("grey30", "#8FBC8BFF", "#6495EDFF", "#BA55D3FF"),
  drawConnectors = TRUE,
  pCutoff = 0.05,
  FCcutoff = 0.5,
  max.overlaps = Inf,  # allow all 17 labels to show
  titleLabSize = 14,
  subtitleLabSize = 12,
  captionLabSize = 12,
  legendLabSize = 12,
  legendIconSize = 3,
  title = paste(i, 'mouse')
) + labs(subtitle = "young vs adult"))
}
```

```{r fig.height=8, fig.width=5}
for (i in names(DEG_bulk_human$meta_results)) {
  highlight_genes <- unique(human_genes[human_genes %in% (DEG_bulk_human$meta_results[[i]] %>% dplyr::filter(FDR < 0.05 & abs(mean_logFC) > 0.15) %>% pull(gene))])

print(EnhancedVolcano(
  DEG_bulk_human$meta_results[[i]],
  lab = DEG_bulk_human$meta_results[[i]]$gene,
  selectLab = highlight_genes,
  x = 'mean_logFC',
  y = 'FDR',
  col = c("grey30", "#8FBC8BFF", "#6495EDFF", "#BA55D3FF"),
  drawConnectors = TRUE,
  pCutoff = 0.05,
  FCcutoff = 0.5,
  max.overlaps = Inf,  # allow all 17 labels to show
  titleLabSize = 14,
  subtitleLabSize = 12,
  captionLabSize = 12,
  legendLabSize = 12,
  legendIconSize = 3,
  title = paste(i, 'human')
) + labs(subtitle = "< 18 months vs > 18 months"))
}
```

```{r eval=FALSE, include=FALSE}
# =================================================================
# SAVE ONLY CORE NICHENET DATA - Calculate tops/filters in Python
# =================================================================

# 1. TARGET GENE LISTS (Core results)
write.csv(data.frame(gene = geneset_NEyoung), paste0(OUTNAME, "nichenet_targets_young.csv"), row.names = FALSE)
write.csv(data.frame(gene = geneset_NEadult), paste0(OUTNAME, "nichenet_targets_adult.csv"), row.names = FALSE)

# 2. MAIN PRIORITIZATION TABLES (All ligand-receptor data)
write.csv(prioritization_tables$prioritization_tbl_ligand_receptor, 
          paste0(OUTNAME, "nichenet_ligand_receptor_prioritization.csv"), row.names = FALSE)

# 3. LIGAND-TARGET RELATIONSHIPS (All ligand-target data)  
write.csv(prioritization_tables$prioritization_tbl_ligand_target, 
          paste0(OUTNAME, "nichenet_ligand_target_prioritization.csv"), row.names = FALSE)

# Print what was saved
cat("Saved 4 core files:\n")
cat("1. nichenet_targets_young.csv -", length(geneset_NEyoung), "target genes\n")
cat("2. nichenet_targets_adult.csv -", length(geneset_NEadult), "target genes\n") 
cat("3. nichenet_ligand_receptor_prioritization.csv - All L-R pairs\n")
cat("4. nichenet_ligand_target_prioritization.csv - All L-T relationships\n")
```



