---
title: "Data overview"
output: html_document
date: "2024-06-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
library(data.table)
library(Seurat)
library(scDblFinder)
library(DoubletFinder)
library(dplyr)
library(remotes)
# remotes::install_github("carmonalab/STACAS")
library(STACAS)
library(SingleR)
# install.packages("scGate")
library(scGate)
library(ggplot2)
library(BiocParallel)
library(harmony)
library(RColorBrewer)
library(stringr)
library(cowplot)
library(scuttle)
# install.packages("devtools")
# BiocManager::install("Homo.sapiens")
# BiocManager::install("jpmam1/scalop") #forked copy of a really good guy 
# devtools::install_github("jlaffy/infercna")
library(infercna)
library(infercnv)
library(parallel)
library(future)
library(clustree)
library(dittoSeq)
# install.packages("ggh4x")
library(ggh4x)
library(EnhancedVolcano)
library(clusterProfiler)
library("org.Hs.eg.db")
library(ReactomePA)
library(msigdbr)
library(qs)
```

```{r read data, include=FALSE}
set.seed(100)
DATADIR = '~/neuroblastoma/data/'
WORKDIR = '~/neuroblastoma/'
PREPRDATADIR = '~/neuroblastoma/preprocessing_data/'

my_colors <- c("#9ACD32FF", "#4682B4FF",  "#DDA0DDFF", "#FFA07AFF", "#8FBC8BFF",
               "#40E0D0FF",  "#F0E68CFF", "#5F9EA0FF", "#D2B48CFF",  
               "#6495EDFF", "#FF69B4FF", "#BA55D3FF",  "#F08080FF", "#32CD32FF",  
               "#FFDAB9FF",  "#87CEEBFF")

seu <- readRDS(paste0(PREPRDATADIR, 'seu_MYCN_MLLT11.rds'))
```

```{r metadata load, include=FALSE}
sample_metadata <- read.csv(paste0(DATADIR, 'samples_metadata_v2.csv'))
sample_metadata[sample_metadata==""] <- NA

rownames(sample_metadata) <- sample_metadata$Sample_dataset


seu$sex <- sample_metadata[seu$SampleID,]$Sex
seu$less18M <- sample_metadata[seu$SampleID,]$less18M
seu$Survival.Death <- sample_metadata[seu$SampleID,]$Survival.Death
seu$Site <- sample_metadata[seu$SampleID,]$Site
seu$INSS_stage <- sample_metadata[seu$SampleID,]$INSS_stage

seu$Study <- sapply(str_split(seu$SampleID, "_"), function(x) {
  if (length(x) == 2) {
    return(x[2])
  } else if (length(x) == 3) {
    return(x[3])
  } else {
    return(NA)
  }
})
```

# Integration

```{r integration plot, fig.height=4, fig.width=12, message=FALSE, warning=FALSE}
plot_grid(DimPlot(seu, group.by = 'Study', cols = my_colors) +
    ggtitle('By studies'),
  DimPlot(seu, group.by = 'SampleID') +
    ggtitle('By samples') +
    NoLegend(), nrow=1, rel_widths = c(0.5, 0.4))
```

```{r}
FeaturePlot(seu, 'percent_mito', cols=c("gray90", '#BA55D3FF')) 
```

# Annotation 

## Define malignant cells

Firstly, I was trying to define malignant cells by known markers

```{r echo=FALSE, fig.height=8, fig.width=12}
FeaturePlot(seu, c('MYCN', 'SLC6A2', 'MLLT11'), cols=c("gray90", '#BA55D3FF'), ncol = 2)
```

Some cells are really different they highly express MYCN and don't express MLLT11. I decided to look at this cluster separately and I realized that almost all cells of this cluster are from one patient.

```{r cluster 15 plot, fig.height=4, fig.width=10}
seu$annotation_simple <- Idents(seu)
Idents(seu) <- seu$RNA_snn_res.0.1

p1 <- DimPlot(seu, cells.highlight = WhichCells(seu, idents = "15"), cols.highlight = '#6495EDFF') +
  NoLegend() +
  ggtitle('Cluster 15')

Idents(seu) <- seu$SampleID
p2 <- DimPlot(seu, cells.highlight = WhichCells(seu, idents = "NB37_verhoeven"), cols.highlight='#FF69B4FF') +
  NoLegend() +
  ggtitle('NB37_verhoeven')

Idents(seu) <- seu$annotation_simple

plot_grid(p1, p2)
rm(p1, p2)
```

Secondly, I used the tool InferCNV than can find CNV (copy number variations) in the data. 

```{r infercnv plot, echo=FALSE}
DimPlot(seu, group.by='infercnv_result') +
  scale_color_manual(values = c('#FF69B4FF', 'grey')) +
  ggtitle('inferCNV')
```

### T-cells

```{r t cells}
FeaturePlot(seu, 'CD7', cols=c("gray90", '#BA55D3FF'))
```

### B-cells

```{r b cells}
FeaturePlot(seu, 'CD79A', cols=c("gray90", '#BA55D3FF'))
```


### Myeloid cells

```{r fig.height=4, fig.width=12}
FeaturePlot(seu, c('CD14', 'CD141', 'CD1C'), cols=c("gray90", '#BA55D3FF'))
```


## Final annotation

```{r}
DimPlot(seu, cols = my_colors)
```

# Analysis by sample

## MYCN amplification 

```{r echo=FALSE, fig.width=12}
plot_data <- as.data.frame(table(seu@meta.data$SampleID, seu$MYCN_MLLT11))
colnames(plot_data) <- c("SampleID", "Category", "Count")
plot_data <- merge(plot_data, sample_metadata, by.x='SampleID', by.y='Sample_dataset')
plot_data$less18M <- ifelse(plot_data$less18M, '<18M', '>18M')

# Calculate the percentages
plot_data <- plot_data %>%
  group_by(SampleID) %>%
  mutate(Percent = Count / sum(Count))

healthy_percent <- plot_data %>%
  filter(Category == "normal cells") %>%
  arrange(Percent)

# Order the SampleID factor levels by the percentage of healthy cells
plot_data$SampleID <- factor(plot_data$SampleID, levels = healthy_percent$SampleID)

ggplot(plot_data, aes(x = SampleID, y = Percent, fill = Category)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("normal cells" = "#87CEEBFF", "MLLT11" = "#FFA07AFF", "MYCN" = "#9ACD32FF", "MYCN+MLLT11" = "#FF69B4FF")) +
  labs(y = "Percent of cells", x = "SampleID", title = "MYCN_MLLT11") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid2(~MYCN, scales = "free_x", space = "free_x", switch = "x",
              strip = strip_themed(
                background_x = elem_list_rect(
                  fill = c("#6495EDFF", "#32CD32FF", '#DDA0DDFF'),
                  color = c("#6495EDFF", "#32CD32FF", '#DDA0DDFF')),
                text_x = elem_list_text(size=14))) +
  ggtitle('MYCN amplification')
```

```{r echo=FALSE, fig.width=12}
plot_data2 <- as.data.frame(table(seu@meta.data$SampleID, seu$annotation_simple))
colnames(plot_data2) <- c("SampleID", "Cell_type", "Count")
plot_data2 <- merge(plot_data2, sample_metadata, by.x='SampleID', by.y='Sample_dataset')
plot_data2$less18M <- ifelse(plot_data2$less18M, '<18M', '>18M')

# Calculate the percentages
plot_data2 <- plot_data2 %>%
  group_by(SampleID) %>%
  mutate(Percent = Count / sum(Count))

plot_data2$Cell_type <- factor(plot_data2$Cell_type, levels = c("T-cells", "B-cells", "Myeloid cells", "Stromal cells", "Tumor cells"))

tumor_percent <- plot_data2 %>%
  filter(Cell_type == "Tumor cells") %>%
  arrange(Percent)


plot_data2$SampleID <- factor(plot_data2$SampleID, levels = tumor_percent$SampleID)

ggplot(plot_data2, aes(x = SampleID, y = Percent, fill = Cell_type)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("Stromal cells" = "#87CEEBFF", "Myeloid cells" = "#FFA07AFF", 
                               "B-cells" = "#9ACD32FF", "T-cells" = "#FF69B4FF", 
                               "Tumor cells" = "#FFDAB9FF")) +
  labs(y = "Percent of cells", x = "SampleID") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid2(~MYCN, scales = "free_x", space = "free_x", switch = "x",
              strip = strip_themed(
                background_x = elem_list_rect(
                  fill = c("#6495EDFF", "#32CD32FF", '#DDA0DDFF'),
                  color = c("#6495EDFF", "#32CD32FF", '#DDA0DDFF')),
                text_x = elem_list_text(size=14))) +
  ggtitle('MYCN amplification')
```

## Age 

```{r fig.width=12}
ggplot(plot_data, aes(x = SampleID, y = Percent, fill = Category)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("normal cells" = "#87CEEBFF", "MLLT11" = "#FFA07AFF", "MYCN" = "#9ACD32FF", "MYCN+MLLT11" = "#FF69B4FF")) +
  labs(y = "Percent of cells", x = "SampleID", title = "MYCN_MLLT11") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid2(~less18M, scales = "free_x", space = "free_x", switch = "x",
              strip = strip_themed(
                background_x = elem_list_rect(
                  fill = c("#6495EDFF", "#32CD32FF"),
                  color = c("#6495EDFF", "#32CD32FF")),
                text_x = elem_list_text(size=14))) +
  ggtitle('Age')
```

```{r fig.width=12}
ggplot(plot_data2, aes(x = SampleID, y = Percent, fill = Cell_type)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("Stromal cells" = "#87CEEBFF", "Myeloid cells" = "#FFA07AFF", 
                               "B-cells" = "#9ACD32FF", "T-cells" = "#FF69B4FF", 
                               "Tumor cells" = "#FFDAB9FF")) +
  labs(y = "Percent of cells", x = "SampleID") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid2(~less18M, scales = "free_x", space = "free_x", switch = "x",
              strip = strip_themed(
                background_x = elem_list_rect(
                  fill = c("#6495EDFF", "#32CD32FF"),
                  color = c("#6495EDFF", "#32CD32FF")),
                text_x = elem_list_text(size=14))) +
  ggtitle('Age')
```

## Sirvival

```{r fig.width=12}
ggplot(plot_data, aes(x = SampleID, y = Percent, fill = Category)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("normal cells" = "#87CEEBFF", "MLLT11" = "#FFA07AFF", "MYCN" = "#9ACD32FF", "MYCN+MLLT11" = "#FF69B4FF")) +
  labs(y = "Percent of cells", x = "SampleID", title = "MYCN_MLLT11") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid2(~Survival.Death, scales = "free_x", space = "free_x", switch = "x",
              strip = strip_themed(
                background_x = elem_list_rect(
                  fill = c("#6495EDFF", "#32CD32FF", '#DDA0DDFF'),
                  color = c("#6495EDFF", "#32CD32FF", '#DDA0DDFF')),
                text_x = elem_list_text(size=14))) +
  ggtitle('Survival')
```

```{r fig.width=12}
ggplot(plot_data2, aes(x = SampleID, y = Percent, fill = Cell_type)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("Stromal cells" = "#87CEEBFF", "Myeloid cells" = "#FFA07AFF", 
                               "B-cells" = "#9ACD32FF", "T-cells" = "#FF69B4FF", 
                               "Tumor cells" = "#FFDAB9FF")) +
  labs(y = "Percent of cells", x = "SampleID") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid2(~Survival.Death, scales = "free_x", space = "free_x", switch = "x",
              strip = strip_themed(
                background_x = elem_list_rect(
                  fill = c("#6495EDFF", "#32CD32FF", '#DDA0DDFF'),
                  color = c("#6495EDFF", "#32CD32FF", '#DDA0DDFF')),
                text_x = elem_list_text(size=14))) +
  ggtitle('Survival')
```

## INSS stage

```{r fig.width=12}
ggplot(plot_data, aes(x = SampleID, y = Percent, fill = Category)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("normal cells" = "#87CEEBFF", "MLLT11" = "#FFA07AFF", "MYCN" = "#9ACD32FF", "MYCN+MLLT11" = "#FF69B4FF")) +
  labs(y = "Percent of cells", x = "SampleID", title = "MYCN_MLLT11") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid2(~INSS_stage, scales = "free_x", space = "free_x", switch = "x",
              strip = strip_themed(
                background_x = elem_list_rect(
                  fill = c("#6495EDFF", "#32CD32FF", '#DDA0DDFF', 
                           "#8FBC8BFF", "#40E0D0FF",  "#F0E68CFF"),
                  color = c("#6495EDFF", "#32CD32FF", '#DDA0DDFF', 
                            "#8FBC8BFF", "#40E0D0FF",  "#F0E68CFF")),
                text_x = elem_list_text(size=11))) +
  ggtitle('INSS stage')
```

```{r fig.width=12}
ggplot(plot_data2, aes(x = SampleID, y = Percent, fill = Cell_type)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("Stromal cells" = "#87CEEBFF", "Myeloid cells" = "#FFA07AFF", 
                               "B-cells" = "#9ACD32FF", "T-cells" = "#FF69B4FF", 
                               "Tumor cells" = "#FFDAB9FF")) +
  labs(y = "Percent of cells", x = "SampleID") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid2(~INSS_stage, scales = "free_x", space = "free_x", switch = "x",
              strip = strip_themed(
                background_x = elem_list_rect(
                  fill = c("#6495EDFF", "#32CD32FF", '#DDA0DDFF', 
                           "#8FBC8BFF", "#40E0D0FF",  "#F0E68CFF"),
                  color = c("#6495EDFF", "#32CD32FF", '#DDA0DDFF', 
                            "#8FBC8BFF", "#40E0D0FF",  "#F0E68CFF")),
                text_x = elem_list_text(size=11))) +
  ggtitle('INSS stage')
```

# Analysis of immune cells by age

I perfomed differential expression analysis in immune cells by age using pseudobulk.

```{r fig.height=8, fig.width=15}
seu$less18M <- as.character(seu$less18M)
bulk <- AggregateExpression(seu, return.seurat = T, slot = "counts", assays = "RNA", 
                            group.by = c("annotation_simple", "SampleID", "less18M"))
Idents(bulk) <- "less18M"

t_cells.de.markers <- FindMarkers(subset(bulk, annotation_simple == 'T-cells'), 
                                  ident.1 = "TRUE", ident.2 = "FALSE", slot = "counts",
                                  test.use = "DESeq2", verbose = F)
write.csv(t_cells.de.markers, paste0(PREPRDATADIR, 't_cells_de_markers_bulk.csv'))
b_cells.de.markers <- FindMarkers(subset(bulk, annotation_simple == 'B-cells'), 
                                  ident.1 = "TRUE", ident.2 = "FALSE", slot = "counts",
                                  test.use = "DESeq2", verbose = F)
write.csv(b_cells.de.markers, paste0(PREPRDATADIR, 'b_cells_de_markers_bulk.csv'))
myeloid.de.markers <- FindMarkers(subset(bulk, annotation_simple == 'Myeloid cells'), 
                                  ident.1 = "TRUE", ident.2 = "FALSE", slot = "counts",
                                  test.use = "DESeq2", verbose = F)
write.csv(myeloid.de.markers, paste0(PREPRDATADIR, 'myeloid_de_markers_bulk.csv'))

volcano_plot <- function(de.markers) {
  EnhancedVolcano(de.markers,
    lab = rownames(de.markers),
    x = 'avg_log2FC',
    y = 'p_val',
    col = c("grey30", "#9ACD32FF", "#6495EDFF", "#FF69B4FF"),
    drawConnectors = TRUE,
    pCutoff = 0.05) +
  labs(subtitle = "< 18M vs > 18M")
}

plot_grid(volcano_plot(t_cells.de.markers) +
            ggtitle('T-cells'),
          volcano_plot(b_cells.de.markers) +
            ggtitle('B-cells'),
          volcano_plot(myeloid.de.markers) +
            ggtitle('Myeloid cells'), 
          nrow = 1)
```

## Enrichment analysis

That's an enrichment analysis of genes **up**-regulated in patients that are younger 18 months

```{r fig.height=4, fig.width=20}
H_t2g <- msigdbr(species = "Homo sapiens", category = "H") %>% 
  dplyr::select(gs_name, entrez_gene)


enrichplot_pos <- function(de.markers) {
  em <- enricher(de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC > 1) %>% 
                            rownames() %>% 
                            mapIds(org.Hs.eg.db, 
                                   keys = .,
                                   column = "ENTREZID",
                                   keytype = "SYMBOL"),
                       TERM2GENE=H_t2g)
  barplot(em) 
}

plot_grid(enrichplot_pos(t_cells.de.markers) +
            ggtitle('T-cells'),
          enrichplot_pos(b_cells.de.markers) +
            ggtitle('B-cells'),
          enrichplot_pos(myeloid.de.markers) +
            ggtitle('Myeloid cells'), 
          nrow = 1)
```

That's an enrichment analysis of genes **down**-regulated in patients that are younger 18 months. There are not significant pathways for T-cells.  

```{r fig.height=4, fig.width=20}
enrichplot_neg <- function(de.markers) {
  em <- enricher(de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC < -1) %>% 
                            rownames() %>% 
                            mapIds(org.Hs.eg.db, 
                                   keys = .,
                                   column = "ENTREZID",
                                   keytype = "SYMBOL"),
                       TERM2GENE=H_t2g)
  barplot(em) 
}

plot_grid('',
          '',
          enrichplot_neg(myeloid.de.markers) +
            ggtitle('Myeloid cells'), 
          nrow = 1)
```

```{r}
t_cells <- qread(paste0(PREPRDATADIR, 't_cells_v2_minusS2fetahu.qs'))
b_cells <- qread(paste0(PREPRDATADIR, 'b_cells_v2_minusS2fetahu.qs'))
myeloid <- qread(paste0(PREPRDATADIR, 'myeloid_cells_v2_minusS2fetahu.qs'))
```

Genes are **up**-regulated in patients that are younger 18 months.

```{r fig.height=15, fig.width=20}
heatmap_tcells <- dittoHeatmap(t_cells, t_cells.de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC > 1) %>% 
                            dplyr::arrange(desc(avg_log2FC)) %>% 
                            rownames(),
             order.by = 'less18M', annot.by = c('less18M', 'Study', 'SampleID'), silent=T,
             # annot.colors = c("#87CEEBFF", "#DDA0DDFF"), 
             main = 'T cells', 
             breaks=seq(-2, 2, 0.2), heatmap.colors = colorRampPalette(c("#6495EDFF", "white", "#FF69B4FF"))(length(seq(-2, 2, 0.2)))) 

heatmap_bcells <- dittoHeatmap(b_cells, b_cells.de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC > 1) %>% 
                            dplyr::arrange(desc(avg_log2FC)) %>% 
                            rownames(),
             order.by = 'less18M', annot.by = c('less18M', 'Study', 'SampleID'), silent=T,
             # annot.colors = c("#87CEEBFF", "#DDA0DDFF"), 
             main = 'B cells', 
             breaks=seq(-2, 2, 0.2), heatmap.colors = colorRampPalette(c("#6495EDFF", "white", "#FF69B4FF"))(length(seq(-2, 2, 0.2)))) 

heatmap_myeloid <- dittoHeatmap(myeloid, myeloid.de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC > 1) %>% 
                            dplyr::arrange(desc(avg_log2FC)) %>% 
                            rownames(),
             order.by = 'less18M', annot.by = c('less18M', 'Study', 'SampleID'), silent=T,
             # annot.colors = c("#87CEEBFF", "#DDA0DDFF"), 
             main = 'Myeloid cells',
             breaks=seq(-2, 2, 0.2), heatmap.colors = colorRampPalette(c("#6495EDFF", "white", "#FF69B4FF"))(length(seq(-2, 2, 0.2)))) 

plot_grid(heatmap_tcells[[4]], heatmap_bcells[[4]], heatmap_myeloid[[4]], nrow = 1)
```

Genes are **down**-regulated in patients that are younger 18 months.

```{r fig.height=15, fig.width=20}
heatmap_tcells_neg <- dittoHeatmap(t_cells, t_cells.de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC < -1) %>% 
                            dplyr::arrange(desc(avg_log2FC)) %>% 
                            rownames(),
             order.by = 'less18M', annot.by = c('less18M', 'Study', 'SampleID'), silent=T,
             # annot.colors = c("#87CEEBFF", "#DDA0DDFF"), 
             main = 'T cells', 
             breaks=seq(-2, 2, 0.2), heatmap.colors = colorRampPalette(c("#6495EDFF", "white", "#FF69B4FF"))(length(seq(-2, 2, 0.2)))) 

heatmap_bcells_neg <- dittoHeatmap(b_cells, b_cells.de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC < -1) %>% 
                            dplyr::arrange(desc(avg_log2FC)) %>% 
                            rownames(),
             order.by = 'less18M', annot.by = c('less18M', 'Study', 'SampleID'), silent=T,
             # annot.colors = c("#87CEEBFF", "#DDA0DDFF"), 
             main = 'B cells', 
             breaks=seq(-2, 2, 0.2), heatmap.colors = colorRampPalette(c("#6495EDFF", "white", "#FF69B4FF"))(length(seq(-2, 2, 0.2)))) 

heatmap_myeloid_neg <- dittoHeatmap(myeloid, myeloid.de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC < -1) %>% 
                            dplyr::arrange(desc(avg_log2FC)) %>% 
                            rownames(),
             order.by = 'less18M', annot.by = c('less18M', 'Study', 'SampleID'),silent=T,
             # annot.colors = c("#87CEEBFF", "#DDA0DDFF"),
             main = 'Myeloid cells',
             breaks=seq(-2, 2, 0.2), heatmap.colors = colorRampPalette(c("#6495EDFF", "white", "#FF69B4FF"))(length(seq(-2, 2, 0.2)))) 

plot_grid(heatmap_tcells_neg[[4]], heatmap_bcells_neg[[4]], heatmap_myeloid_neg[[4]], nrow = 1)
```
