---
title: "T_cells_v1"
author: "Kate Petrenko"
date: "2024-07-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
library(data.table)
library(Seurat)
library(scDblFinder)
library(DoubletFinder)
library(dplyr)
library(remotes)
# remotes::install_github("carmonalab/STACAS")
library(STACAS)
library(SingleR)
# install.packages("scGate")
library(scGate)
library(ggplot2)
library(BiocParallel)
library(harmony)
library(RColorBrewer)
library(stringr)
library(cowplot)
library(scuttle)
# install.packages("devtools")
# BiocManager::install("Homo.sapiens")
# BiocManager::install("jpmam1/scalop") #forked copy of a really good guy 
# devtools::install_github("jlaffy/infercna")
library(infercna)
library(infercnv)
library(parallel)
library(future)
library(clustree)
library(dittoSeq)
# install.packages("ggh4x")
library(ggh4x)
library(EnhancedVolcano)
library(clusterProfiler)
library("org.Hs.eg.db")
library(ReactomePA)
library(msigdbr)
library(qs)
# BiocManager::install('multtest')
# install.packages('metap')
```

```{r read data, include=FALSE}
set.seed(100)
DATADIR = '~/neuroblastoma/data/'
WORKDIR = '~/neuroblastoma/'
PREPRDATADIR = '~/neuroblastoma/preprocessing_data/'

my_colors <- c("#9ACD32FF", "#4682B4FF",  "#DDA0DDFF", "#FFA07AFF", "#8FBC8BFF",
               "#40E0D0FF",  "#F0E68CFF", "#5F9EA0FF", "#D2B48CFF",  
               "#6495EDFF", "#FF69B4FF", "#BA55D3FF",  "#F08080FF", "#32CD32FF",  
               "#FFDAB9FF",  "#87CEEBFF")

t_cells <- qread(paste0(PREPRDATADIR, 't_cells_v2.qs'))
```

# Overview of all immune cells

I decided to redo the analysis after reading a lot about integrated analysis. I discovered that it isn’t statistically correct to count differential expression (DE) by cells. They advised using pseudobulk for this purpose, so I did it. Here, I present DE by pseudobulk for each patient. 

```{r fig.height=8, fig.width=15}
t_cells.de.markers <- read.csv(paste0(PREPRDATADIR, 't_cells_de_markers_bulk.csv'), row.names = 1)
b_cells.de.markers <- read.csv(paste0(PREPRDATADIR, 'b_cells_de_markers_bulk.csv'), row.names = 1)
myeloid.de.markers <- read.csv(paste0(PREPRDATADIR, 'myeloid_de_markers_bulk.csv'), row.names = 1)

volcano_plot <- function(de.markers) {
  EnhancedVolcano(de.markers,
    lab = rownames(de.markers),
    x = 'avg_log2FC',
    y = 'p_val_adj',
    col = c("grey30", "#9ACD32FF", "#6495EDFF", "#FF69B4FF"),
    drawConnectors = TRUE,
    pCutoff = 0.05) +
  labs(subtitle = "< 18M vs > 18M")
}

plot_grid(volcano_plot(t_cells.de.markers) +
            ggtitle('T-cells'),
          volcano_plot(b_cells.de.markers) +
            ggtitle('B-cells'),
          volcano_plot(myeloid.de.markers) +
            ggtitle('Myeloid cells'), 
          nrow = 1)
```

## Enrichment

```{r include=FALSE}
H_t2g <- msigdbr(species = "Homo sapiens", category = "H") %>% 
  dplyr::select(gs_name, entrez_gene)

ImmuneSigDB <- msigdbr(species = "Homo sapiens", category = "C5") %>% 
  dplyr::select(gs_name, entrez_gene)

enrichplot_pos <- function(de.markers, gene.set, logFC_treshlod=1) {
  em <- enricher(de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC > logFC_treshlod) %>% 
                            rownames() %>% 
                            mapIds(org.Hs.eg.db, 
                                   keys = .,
                                   column = "ENTREZID",
                                   keytype = "SYMBOL"),
                       TERM2GENE=gene.set)
  barplot(em) 
}

enrichplot_neg <- function(de.markers, gene.set, logFC_treshlod=-1) {
  em <- enricher(de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC < logFC_treshlod) %>% 
                            rownames() %>% 
                            mapIds(org.Hs.eg.db, 
                                   keys = .,
                                   column = "ENTREZID",
                                   keytype = "SYMBOL"),
                       TERM2GENE=gene.set)
  barplot(em) 
}
```

```{r fig.height=7, fig.width=20}
plot_grid(enrichplot_pos(t_cells.de.markers %>% 
                            filter(p_val_adj < 0.05 & avg_log2FC > 1) %>% rownames(), H_t2g) +
            ggtitle('Positive markers'), 
          enrichplot_pos(t_cells.de.markers, ImmuneSigDB) +
            ggtitle('Positive markers'), 
          nrow = 1)
```

```{r fig.height=7, fig.width=20}
plot_grid(enrichplot_neg(b_cells.de.markers, H_t2g) +
            ggtitle('Negative markers'), 
          enrichplot_neg(b_cells.de.markers1, ImmuneSigDB) +
            ggtitle('Negative markers'), 
          nrow = 1)
```

## DE
```{r}
plots <- VlnPlot(t_cells, features = c("LYZ", "ISG15", "CXCL10"), split.by = "stim", group.by = "seurat_annotations",
    pt.size = 0, combine = FALSE)
wrap_plots(plots = plots, ncol = 1)
```

# T-cells overview

```{r include=FALSE}
t_cells <- t_cells %>% 
  FindNeighbors(reduction = "integrated.cca.tcells", dims = 1:20) %>%
  FindClusters(resolution = c(0.2)) %>%
  RunUMAP(reduction = "integrated.cca.tcells", dims = 1:15)

clustree(t_cells)

t_cells <- subset(t_cells, subset = RNA_snn_res.0.2 != '3')
t_cells <- subset(t_cells, subset = RNA_snn_res.0.2 != '8')

t_cells <- t_cells %>% 
  FindNeighbors(reduction = "integrated.cca.tcells", dims = 1:20) %>%
  FindClusters(resolution = c(0.1)) %>%
  RunUMAP(reduction = "integrated.cca.tcells", dims = 1:15)

Idents(t_cells) <- 'RNA_snn_res.0.1'

DimPlot(t_cells, group.by = 'RNA_snn_res.0.1', cols = my_colors)
```

```{r fig.height=4, fig.width=10}
plot_grid(#DimPlot(t_cells, group.by = 'celltype_humatlas_fine') + NoLegend(),
          DimPlot(t_cells, group.by = 'Study', cols = my_colors),
          #DimPlot(t_cells, group.by = 'scGate_multi'),
          DimPlot(t_cells, group.by = 'less18M', cols = my_colors[10:12]), 
          nrow=1)
```

```{r fig.height=4, fig.width=10}
plot_grid(dittoBarPlot(t_cells, 'less18M', group.by = 'Study', color.panel = my_colors[2:3]),
          dittoBarPlot(t_cells, 'less18M', group.by = 'sex', color.panel = my_colors[2:3]),
          nrow = 1)
```

# Analysis by clusters

Here I analyze cells by clusters. 

On the plot B we can see phases of cell cycle that I calculated based on cell cycle genes.

```{r fig.height=4, fig.width=10, message=FALSE, warning=FALSE}
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
t_cells <- CellCycleScoring(t_cells, s.features = s.genes, g2m.features = g2m.genes)

plot_grid(DimPlot(t_cells, cols = my_colors) + ggtitle('Clusters'),
          DimPlot(t_cells, group.by = 'Phase', cols = my_colors),
          labels = c('A', "B"))
```

Distribution of age by clusters.

```{r fig.height=4, fig.width=10.5}
p1 <- dittoBarPlot(t_cells, "less18M", group.by = "RNA_snn_res.0.1", color.panel = my_colors) +
  ggtitle('') +
  xlab('Clusters') +
  NoLegend()
p2 <- dittoBarPlot(t_cells, "less18M", group.by = "RNA_snn_res.0.1", 
                   scale='count', color.panel = my_colors) +
  ggtitle('') +
  xlab('Clusters') +
  guides(fill=guide_legend(title="< 18 months"))
p3 <- dittoBarPlot(t_cells, "RNA_snn_res.0.1", group.by = "less18M", 
                   color.panel = my_colors) +
  ggtitle('') +
  guides(fill=guide_legend(title="Clusters"))

cowplot::plot_grid(p1, p2, p3, 
          nrow=1, rel_widths = c(1, 1.5, 1),
          labels = c('A', "B", "C"))
```

Distribution of studies by clusters. 

```{r fig.height=4, fig.width=10.5}
p1 <- dittoBarPlot(t_cells, "Study", group.by = "RNA_snn_res.0.1", color.panel = my_colors) +
  ggtitle('') +
  xlab('Clusters') +
  NoLegend()
p2 <- dittoBarPlot(t_cells, "Study", group.by = "RNA_snn_res.0.1", 
                   scale='count', color.panel = my_colors) +
  ggtitle('') +
  xlab('Clusters') +
  guides(fill=guide_legend(title="Study"))
p3 <- dittoBarPlot(t_cells, "RNA_snn_res.0.1", group.by = "Study", 
                   color.panel = my_colors) +
  ggtitle('') +
  guides(fill=guide_legend(title="Clusters"))

cowplot::plot_grid(p1, p2, p3, 
          nrow=1, rel_widths = c(1, 1.5, 1),
          labels = c('A', "B", "C"))
```

Distribution of patients by clusters. 

```{r fig.height=6, fig.width=10}
dittoBarPlot(t_cells, "SampleID", group.by = "RNA_snn_res.0.1") +
  ggtitle('By samples') +
  xlab('Clusters') +
  guides(fill=guide_legend(title="SampleID"))
```

## Cluster markers

There are markers of our clusters. 

```{r fig.height=4, fig.width=10, message=FALSE, warning=FALSE}
t_cells.clusters.markers <- FindAllMarkers(t_cells, only.pos = TRUE, min.pct = 0.25,
                                     logfc.threshold = 0.25)

top5 <- t_cells.clusters.markers %>% 
  dplyr::filter(p_val_adj < 0.05) %>% 
  group_by(cluster) %>% 
  top_n(n = 5, wt = avg_log2FC)


dittoDotPlot(t_cells, vars = unique(top5$gene), group.by = "RNA_snn_res.0.1", 
             max.color='#6495EDFF', min.color='white') +
  ylab('Clusters') +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

## Enrichment analysis by clusters

That's an enrichment analysis of cluster markers. I took all significnatn markers (not only 5 as in previous plot). We can describe subtypes of B-cells based on these plots. 

Here I used hallmarks gene sets from Human MSigDB Collections as reference.

```{r fig.height=12, fig.width=16}
plot_grid(enrichplot_pos(t_cells.clusters.markers %>% 
                           filter(cluster == '0') %>% 
                           dplyr::filter(p_val_adj < 0.05), H_t2g, 0.5) +
            ggtitle('Cluster 0'),
          enrichplot_pos(t_cells.clusters.markers %>% 
                           filter(cluster == '1') %>% 
                           dplyr::filter(p_val_adj < 0.05), H_t2g, 0.1) +
            ggtitle('Cluster 1'),
          enrichplot_pos(t_cells.clusters.markers %>% 
                           filter(cluster == '2') %>% 
                           dplyr::filter(p_val_adj < 0.05), H_t2g, 0.1) +
            ggtitle('Cluster 2'),
          enrichplot_pos(t_cells.clusters.markers %>% 
                           filter(cluster == '3') %>% 
                           dplyr::filter(p_val_adj < 0.05), H_t2g, 0.5) +
            ggtitle('Cluster 3'),
          enrichplot_pos(t_cells.clusters.markers %>% 
                           filter(cluster == '4') %>% 
                           dplyr::filter(p_val_adj < 0.05), H_t2g, 0.5) +
            ggtitle('Cluster 4'),
          enrichplot_pos(t_cells.clusters.markers %>% 
                           filter(cluster == '5') %>% 
                           dplyr::filter(p_val_adj < 0.05), H_t2g, 0.5) +
            ggtitle('Cluster 5'),
          nrow = 3)
```

# Annotation 

## T-cells common markers

```{r}
FeaturePlot(t_cells, c('CD3E', 'CD3D', 'CD3G'))
```

```{r fig.height=8, fig.width=10}
FeaturePlot(t_cells, c('CD4', 'CD8A', 'CD8B'))
```
# Need to finish ^

# Enrichment 




